### 2014-02-22 paper.tex
Add introduction.

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{ICFP '14}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Type-Safe Client-Server Communication in Web Applications}
\subtitle{Subtitle Text, if any}

\authorinfo{Koen Claessen ans Anton Ekblad}
           {Chalmers University of Technology}
           {\{koen,antonek\}@chalmers.se}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2
%
\keywords
web

\section{Introduction}

Development of web applications using conventional means is often no task for
the faint of heart. Code written in Javascript, the only widely supported
language for client-side web applications, is often confusing and error-prone,
much due to the language's lack of modularity, encapsulation facilities and
type safety.

Worse, most web applications, being intended to facilitate communication, data
storage and other tasks involving some centralized resource, also require a
significant server component. This component is usually implemented as a
completely separate program, and communicates with the client code over some
network protocol.

Having your application rely on some network protocol for
its internal functionality involves writing a lot of boilerplate network code
and some serious advance planning of your application's features to be able to
design a suitable protocol, hampering incremental development. Were we to
discount this initial design burden, ensuring that the protocol is used
correctly throughout the code base as the application evolves and gains new
features would still put an ever increasing burden on the developer.

This state of things is not a concious design choice - most web applications
are conceptually a single entity, not two programs which just happen to talk
to each other over a network - but a consequence of there being a large,
distributed network between the client and server parts.
However, such implementation details should not be allowed to dictate the way
we structure and reason about our applications - clearly, an abstraction is
called for.

In this paper, we propose a programming model in which web applications are
written as a single entity from which client and server binaries are generated
during compilation. Type annotations in the source program control which parts
are executed on the server and which are executed on the client, and the two
communicate using type safe RPC calls.

Other approaches exist which attempt to tackle the same issue[CITE: STRATIFIEDJS, DUETTO, SUNROOF]
which solve some, but not all, of the associated problems. In particular, our
solution has the following properties:

\begin{itemize}
  \item Type safety: StratifiedJS, being a close relative of the original
        Javascript language, provides type safety for neither RPC calls nor the
        core language and Duetto, being a C++ dialect, is neither memory safe
        nor type safe.
  \item Modularity: our programming model is built as a library on top of
        Haskell and is thus easily portable to any Haskell-to-Javascrit
        compiler. In contrast, StratifiedJS and Duetto are compilers in their
        own right, not usable together with other similar tools.
  \item Purely functional semantics: in contrast to StratifiedJS, Duetto and
        Sunroof, our programming model enables the use of the full Haskell
        language both on the server and the client.
  \item Code reuse: with Sunroof, client code is written in an embedded
        language whereas server code is written in full Haskell, necessitating
        duplicate implementations of functions which may be of use to both the
        client and the server parts. In our solution, once a function is
        written it is usable on both the server and the client.
\end{itemize}

\section{The Haste compiler}
\section{A seamless programming model}
\section{Implementation}
\section{Discussion and related work}

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

### 2014-02-26 paper.tex
Add related work and discussion, restructure introduction.

@@
 
-Development of web applications using conventional means is often no task for
-the faint of heart. Code written in Javascript, the only widely supported
-language for client-side web applications, is often confusing and error-prone,
-much due to the language's lack of modularity, encapsulation facilities and
-type safety.
+Development of web applications is no task for the faint of heart.
+The conventional method involves splitting your program into two logical parts,
+writing the one in Javascript, which is notorious even among its proponents for
+being wonky and error-prone, and the other in the current language du jour.
+Then, two are glued together using whichever home-grown network protocol seems
+to fit the application. However, most web applications are conceptually
+monolithic entities, making this forced split an undesirable hindrance which
+introduces new possibilities for defects, adds development overhead and
+prevents code reuse.
+
+Several attempts have been made at tackling this problem, but none is, in our
+view, satisfactory. In this paper, we propose a programming model in which a
+web application is written as a single program from which client and server
+binaries are generated during compilation. Type annotations in the source
+program control which parts are executed on the server and which are executed
+on the client, and the two communicate using type safe RPC calls. Functions
+which are not explicitly declared as server-side or client-side are usable by
+either side. Our solution is implemented as a Haskell library requiring minimal
+to no compiler support, ensuring portability and modularity.
+
+\paragraph{Motivation}
+
+Code written in Javascript, the only widely supported language for client-side
+web applications, is often confusing and error-prone, much due to the
+language's lack of modularity, encapsulation facilities and type safety.
 
@@
 
-In this paper, we propose a programming model in which web applications are
-written as a single entity from which client and server binaries are generated
-during compilation. Type annotations in the source program control which parts
-are executed on the server and which are executed on the client, and the two
-communicate using type safe RPC calls.
-
-Other approaches exist which attempt to tackle the same issue[CITE: STRATIFIEDJS, DUETTO, SUNROOF]
-which solve some, but not all, of the associated problems. In particular, our
-solution has the following properties:
-
-\begin{itemize}
-  \item Type safety: StratifiedJS, being a close relative of the original
-        Javascript language, provides type safety for neither RPC calls nor the
-        core language and Duetto, being a C++ dialect, is neither memory safe
-        nor type safe.
-  \item Modularity: our programming model is built as a library on top of
-        Haskell and is thus easily portable to any Haskell-to-Javascrit
-        compiler. In contrast, StratifiedJS and Duetto are compilers in their
-        own right, not usable together with other similar tools.
-  \item Purely functional semantics: in contrast to StratifiedJS, Duetto and
-        Sunroof, our programming model enables the use of the full Haskell
-        language both on the server and the client.
-  \item Code reuse: with Sunroof, client code is written in an embedded
-        language whereas server code is written in full Haskell, necessitating
-        duplicate implementations of functions which may be of use to both the
-        client and the server parts. In our solution, once a function is
-        written it is usable on both the server and the client.
-\end{itemize}
-
-\section{The Haste compiler}
 \section{A seamless programming model}
 \section{Implementation}
+\section{The Haste compiler}
 \section{Discussion and related work}
 
-\appendix
-\section{Appendix Title}
+Other approaches to this problem exist, notably in the form of the
+Conductance\ \cite{conductance}, Duetto\ \cite{duetto} and
+Sunroof\ \cite{sunroof} projects, all of which implement some synchronous
+abstraction over the network between the client and the server, much like our
+solution.
 
-This is the text of the appendix, if you need one.
+\paragraph{Conductance} Conductance is an application server built on
+StratifiedJS, a Javascript language extension which adds a few niceties such as
+cooperative multitasking and more concise syntax for many common tasks.
+Conductance uses an RPC-based model for client-server communication, much like
+our own, but also adds the possibility for the server to independently transmit
+data back to the client through the use of shared variables or call back into
+the client by way of function objects received via RPC call, as well as the
+possibility for both client and server to seamlessly modify variables located
+on the opposite end of the network.
+
+While Conductance gets rid of the callback-based programming model endemic to
+regular Javascript, it still suffers from many of its usual drawbacks. In
+particular, the the weak typing of Javascript poses a problem in that the
+programmer is in no way reprimanded by her tools for using server APIs
+incorrectly or trying to transmit values which can not be sensibly serialized
+and de-serialized, such as DOM nodes. Wrongly typed programs will thus crash, or
+even worse, gleefully keep running with erroneous state due to implicit type
+conversions, rather than give the programmer some advance warning that something
+is amiss.
+
+We are also not completely convinced that the ability to implicitly back and
+forth over the network is a unilaterally good thing; while this indeed provides
+the programmer some extra convenience, it also requires the programmer to
+exercise extra caution to avoid inadvertently sending large amounts of data
+back and forth over the network.
+
+\paragraph{Duetto} Duetto is a C++ compiler targeting the web, written from
+the ground up to produce code for both client and server simultaneously. It
+utilizes the new attributes mechanism introduced in C++11 to designate
+functions and data to live on either client or server side. Any calls to a
+function on the other side of the network and attempts to access remote data
+are implicit, requiring no extra annotations or scaffolding at the call site.
+
+Much like Conductance, Duetto suffers somewhat from its heritage: while the
+client-side code is not memory-unsafe, as it is not possible to generate
+memory-unsafe Javascript code, its server-side counterpart unfortunately is.
+Our reservations expressed about how network communication in Duetto can be
+initiated implicitly apply to Duetto as well.
+
+\paragraph{Sunroof} In contrast to the Conductance and Duetto, Sunroof is an
+embedded language. Implemented as a Haskell library, it allows the programmer
+to use Haskell to write code which is compiled to Javascript and executed on
+the client. The language can best be described as having Javascript semantics
+with Haskell's type system. Communication between client and server is
+accomplished through the use of ``downlinks'' and ``uplinks'', allowing for
+data to be sent to and from the client respectively.
+
+Sunroof is completely type-safe, in the DSL itself as well as in the
+communication with the Haskell host. However, the fact that client and server
+must be written in two separate languages - any code used to generate
+Javascript must of course be built solely from the primitives of the Sunroof
+language, precluding use of general Haskell code - makes code reuse hard.
+
+\paragraph{Advantages to our approach} We believe that our approach has a
+number of distinct advantages to the aforementioned attacks on the problem.
+
+Our approach gives the programmer access to the same strongly typed,
+general-purpose functional language on both client and server; any code which
+may be of use to both client and server is effortlessly shared, leading to less
+duplication of code and increased possibilities for reusing third party
+libraries.
+
+Any and all communication between client and server is both strongly typed
+and made explicit by the use of the `onServer` function, with the programmer
+having complete control over the serialization and de-serialization of data
+using the appropriate type classes. Aside from the obvious advantages of type
+safety, making the crossing of the network boundary explicit helps the
+programmer make an informed decision as to when and where server communication
+is appropriate.
+
+Our programming model is implemented as a library, assuming only two Haskell
+compilers, one targeting Javascript and one targeting the programmer's server
+platform of choice. While we use Haste as our Javascript-targeting compiler,
+modifying our implementation to use GHCJS or even the Javascript backend of UHC
+would be entirely straightforward. This implementation not only allows for
+greater flexibility, but also eliminates any potentially hairy interaction with
+compiler internals.
+
+\paragraph{Limitations} Unlike any of the related works, our approach takes a
+firm stand, regarding the client as the driver in the client-server
+relationship with the server taking on the role of a passive computational
+or storage resource. The server may thus not call back into the client at
+arbitrary points but is instead limited to returning answers to client-side
+queries. This is clearly less flexible than the back-and-forth model of Sunroof
+and Duetto or the shared variables of Conductance. However, we believe that
+this restriction makes program flow easier to follow and comprehend. Like the
+immutability of Haskell, this model gives programmers a not-so-subtle hint
+as to how they may want to structure their programs. Should this model prove
+insufficiently expressive, extending our existing model with an `onClient`
+counterpart to `onServer` would be relatively straightforward.
+
+\section{Future work}
+
+\paragraph{Information flow control} Web applications often make use of a wide
+range of third party code for user tracking, advertising, collecition of
+statistics and a wide range of other tasks. Any piece of code executing in the
+context of a particular web session may not only interact with any other piece
+of code executing in the same context, but may also perform basically limitless
+communication with third parties and may thus, inadvertently or not, leak
+information about the application state. This is obviously highly undesirable
+for many applications, which is why there is ongoing work in controlling the
+information flow within web applications\ \cite{jsflow}.
+
+While this would indeed provide an effective defence towards attackers and
+programming mistakes alike, there is value in being able to tell the two apart,
+as well as in catching policy violations resulting from programming mistakes
+as early as possible. An interesting venue of research would be to investigate
+whether we can take advantage of our strong typing to generate security policies
+for such an information flow control scheme, as well as ensure that this policy
+is not violated at compile time. This could shorten development cycles as well
+as give a reasonable level of confidence that any run time policy violation is
+indeed an attempted attack.
+
+\paragraph{Reactive client-server toolkit} While we believe that our work
+represents a step forward in making web applications more robust and easier to
+develop, the basic model for updating the application's user-visible state is
+still essentially imperative. Implementing an FRP-style\ \cite{ppfrp} GUI
+programming model on top of our client-server abstraction may prove a fruitful
+way to further improve the programmability of web-based distributed
+applications.
 
@@
 
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
+\bibitem[Bracker, Gill (2014)]{sunroof}
+J. Bracker and A. Gill, Sunroof: A Monadic DSL for Generating JavaScript, Practical Aspects of Declarative Languages, 65-80
+
+\bibitem[Conductance application server (2014)]{conductance}
+THE INTERNETS, http://conductance.io
+
+\bibitem[Duetto compiler (2014)]{duetto}
+THE INTERNETS, http://leaningtech.com/duetto/
+
+\bibitem[Elliott (2009)]{ppfrp}
+C. M. Elliott, Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell
+
+\bibitem[Hedin et al (2014)]{jsflow}
+D. Hedin, A. Birgisson, L. Bello and A. Sabelfeld, JSFlow: Tracking Information Flow in JavaScript and its APIs
 

### 2014-02-26 paper.tex
Explain choice of Haste and benefits of compiling from STG.

@@
 \section{The Haste compiler}
+
+In order to allow the same language to be used on both client and server, we
+obviously need some way to compile that language into Javascript. To this end,
+we make use of the Haste compiler\ \cite{haste}, developed as a MSc thesis and
+extended as part of this work. Haste builds on the GHC compiler to provide
+the full Haskell language, including most GHC-specific extensions, in the
+browser.
+
+\paragraph{Choosing a compiler} Haste is by no means the only
+Javascript-targeting compiler for a purely functional language. In particular,
+the GHC-based GHCJS\ \cite{ghcjs} and UHC\ \cite{uhc} compilers are both
+capable of compiling standard Haskell into Javascript; the Fay\ \cite{fay}
+language was designed from the ground up to target the web space using a subset
+of Haskell; and there exist solutions for compiling Erlang\ \cite{jserlang} and
+Clean\ \cite{jsclean} to Javascript as well.
+
+In the face of this multitude of options, why did we settle on Haste? We make
+the following demands of a candidate source language:
+
+\begin{itemize}
+  \item The language must provide a static type system, since one of our
+        primary concerns is to reduce defect rates through static typing of
+        the client-server communication channel.
+  \item The language must be compilable to both Javascript and a format
+        suitable for server-side execution as we want our web applications
+        to be written as a single, monolithic program.
+  \item We also want our language to be practical, in the sense that its output
+        should not be disproportionately large or slow. This is more important
+        for the client part, where we have to contend with slow connections and
+        the overhead of running within an interpreter, than for the server part.
+\end{itemize}
+
+Our first criterion rules out the dynamically typed Erlang. The second rules
+out Fay which unfortunately lacks critical Haskell features, most notably type
+classes.
+
+This still leaves us with four options: Clean, GHCJS, Haste and UHC. However,
+the output from GHCJS is prohibitively large - often in the range of several
+megabytes for very simple programs - which rules it out for our use case, in
+spite of providing slightly better compatibility with low-level GHC features
+than Haste. Similarly, the code produced by the UHC Javascript backend tends to
+be relatively slow due to it lacking the more mature optimization capabilities
+of GHC, as well as being in the same ballpark as GHCJS for size.\ \cite{haste}
+
+This leaves us choosing between Clean and Haste for our client compiler. Since
+most of our abstractions, both the cooperative multitasking and the type-safe
+client-server communication itself, are easily expressed as monads, well
+supported in Haskell thanks to do-notation, the choice fell to Haste.
+
+\paragraph{Implementation} As previously mentioned, Haste offloads much of the
+heavy lifting of compilation - parsing, type checking, intermediate code
+generation and many optimizations - onto GHC, and takes over code generation
+after the STG generation step, at the very end of the compilation process.
+STG is the last intermediate representation used by the GHC compiler before
+the final code generation takes place and has several benefits:
+
+\begin{itemize}
+  \item STG is still a functional intermediate representation, based on the
+        lambda calculus. When generating code for a high level target language
+        such as Javascript, where functions are first class objects, this
+        allows for a higher level translation than when doing traditional
+        compilation to lower level targets like stack machines or register
+        machines. This in turn allows us to make more efficient use of the
+        target language's runtime, leading to smaller, faster code.
+  \item In contrast to Haskell itself and GHC's intermediate Core language, STG
+        represents `thunks`, the construct used by GHC to implement non-strict
+        evaluation, as closures which are explicitly created and evaluated.
+        Closures are decorated with a wealth of information, such as their set
+        of captured varibles, any type information needed for code generation,
+        and so on. While extracting this information manually is not very hard,
+        having this done for us means we can get away with a simpler
+        compilation pipeline.
+  \item The language is very small, essentially only comprising lambda
+        abstraction and application, plus primitive operations and facilities
+        for calling out to other languages. Again, this allows the Haste
+        compiler to be a very simple thing indeed.
+  \item Any extensions to the Haskell language implemented by GHC will already
+        have been translated into this very simple intermediate format,
+        allowing us to support basically any extension GHC supports without
+        effort.
+\end{itemize}
+
 \section{Discussion and related work}
@@
 
-\paragraph{Reactive client-server toolkit} While we believe that our work
+\paragraph{Reactive client-server GUI toolkit} While we believe that our work
 represents a step forward in making web applications more robust and easier to
@@
 \bibitem[Bracker, Gill (2014)]{sunroof}
-J. Bracker and A. Gill, Sunroof: A Monadic DSL for Generating JavaScript, Practical Aspects of Declarative Languages, 65-80
+J. Bracker and A. Gill: Sunroof: A monadic DSL for generating JavaScript, Practical Aspects of Declarative Languages, 65-80, 2014
 
 \bibitem[Conductance application server (2014)]{conductance}
-THE INTERNETS, http://conductance.io
+THE INTERNETS: http://conductance.io
+
+\bibitem[Dijkstra et al (2013)]{uhc}
+A. Dijkstra, J. Stutterheim, A. Vermeulen, S. D. Swierstra: Building JavaScript applications with Haskell, Implementation and Application of Functional Languages, pp 37-52, 2013
+
+\bibitem[Domoszlai et al (2011)]{jsclean}
+L. Domoszlai, E. Bruël, J. M. Jansen: Implementing a non-strict purely functional language in JavaScript, in Acta Universitatis Sapientiae, 2011
 
 \bibitem[Duetto compiler (2014)]{duetto}
-THE INTERNETS, http://leaningtech.com/duetto/
+THE INTERNETS: http://leaningtech.com/duetto/
+
+\bibitem[Ekblad (2012)]{haste}
+A. Ekblad: Towards a declarative web, MSc thesis, University of Gothenburg, 2012
 
 \bibitem[Elliott (2009)]{ppfrp}
-C. M. Elliott, Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell
+C. M. Elliott: Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell, 2009
+
+\bibitem[The Fay language (2014)]{fay}
+THE INTERNETS: http://fay-lang.org
+
+\bibitem[Guthrie (2014)]{jserlang}
+G. Guthrie: http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/, 2014
 
 \bibitem[Hedin et al (2014)]{jsflow}
-D. Hedin, A. Birgisson, L. Bello and A. Sabelfeld, JSFlow: Tracking Information Flow in JavaScript and its APIs
+D. Hedin, A. Birgisson, L. Bello and A. Sabelfeld: JSFlow: Tracking Information Flow in JavaScript and its APIs, 2014
+
+\bibitem[Nazarov (2012)]{ghcjs}
+V. Nazarov: GHCJS Haskell to Javascript Compiler, https://github.com/ghcjs/ghcjs
 

### 2014-02-26 paper.tex
Add reference for STG.

@@
 after the STG generation step, at the very end of the compilation process.
-STG is the last intermediate representation used by the GHC compiler before
-the final code generation takes place and has several benefits:
+STG\ \cite{stg} is the last intermediate representation used by GHC before
+the final code generation takes place and has several benefits for use as
+Haste's target language:
 
@@
 
+\bibitem[Peyton Jones (1992)]{stg}
+S. Peyton Jones: Implementing lazy functional languages on stock hardware: the Spineless Tagless G-machine, Journal of Functional Programming, 1992
+
 \end{thebibliography}

### 2014-02-26 paper.tex
Add section on Haste implementation.

@@
 
+\paragraph{Design of the Haste compiler} Haste is first and foremost intended
+to be a practical compiler; the ability to produce code which is lean and fast
+enough to be used in the real world is critical. To this end, we make several
+design choices and tradeoffs, some of which would be viewed as detrimental in a
+more purist context.
+
+One such tradeoff is the decision to make use of the facilities provided by the
+web browser - most prominently the Javascript engine's native garbage
+collector - as far as possible rather than implementing parts of the GHC
+runtime in Javascript. In the same vein, a conscious effort is made to generate
+as ``regular-looking'' code as possible. In some rare cases, this means having
+to sacrifice compatibility with low-level GHC features. Most notably, Haste
+does not support weak references or pre-emptive multitasking, as these would
+require modifications to the runtime system and generated code which confound
+the structure of the resulting Javascript and have a detrimental impact on code
+size and execution speed.
+
+GHCJS takes the opposite view, that faithful emulation of the GHC runtime is
+paramount. Of course, this brings an impressive degree of compatibility, at the
+cost of the blowup in code size mentioned previously.
+
 \paragraph{Implementation} As previously mentioned, Haste offloads much of the
@@
         effort.
+  \item Application of external functions is always saturated, as is
+        application of most other functions. This allows for compiling most
+        function applications into simple Javascript function calls, limiting
+        the use of the slower dynamic techniques required to handle curried
+        functions in the general case\ \cite{fastcurry} to cases where it is
+        simply not possible to statically determine the arity of a function.
 \end{itemize}
+
+In light of its heavy reliance on STG, it may be more correct to categorize
+Haste as an STG compiler rather than a Haskell compiler.
+
+The runtime data representation of Haste programs is kept as close to regular
+Javascript programs as possible. The numeric types are represented using the
+Javascript \texttt{Number} type, which is defined as the IEEE754 double
+precision floating point type. This obviously adds some overhead to operations
+on integers as overflow and non-integer divisions must be handled. However, this
+is common practice in hand-written Javascript as well, and is generally handled
+efficiently by Javascript engines.
+
+Values of non-primitive data types in Haskell consist of a data constructor and
+zero or more arguments. In Haste, these values are represented using arrays,
+with the first element representing the data constructor and the following
+values representing its arguments. For instance, the value \texttt{42 :: Int}
+is represented as \texttt{[0, 42]}, the leading \texttt{0} representing the
+zeroth constructor of the \texttt{Int} type and the \texttt{42}, obviously,
+representing the ``machine'' integer. It may seem strange that a limited
+precision integer is represented using one level of indirection rather than as
+a simple number, but recall that the \texttt{Int} type is defined by GHC as
+\texttt{data Int = I\# Int\#}.
+
+Functions are represented as plain Javascript functions, one of the blessings
+of targeting a high level language, and application is can therefore be
+implemented as its Javascript counterpart in most cases. In the general case,
+however, functions may be curried. For such cases where the arity of an applied
+function can not be determined statically, application is implemented using the
+eval/apply method described in \cite{fastcurry} instead.
 
@@
 
-We are also not completely convinced that the ability to implicitly back and
-forth over the network is a unilaterally good thing; while this indeed provides
-the programmer some extra convenience, it also requires the programmer to
-exercise extra caution to avoid inadvertently sending large amounts of data
-back and forth over the network.
+We are also not completely convinced that the ability to implicitly pass data
+back and forth over the network is a unilaterally good thing; while this indeed
+provides the programmer some extra convenience, it also requires the programmer
+to exercise extra caution to avoid inadvertently sending large amounts of data
+over the network.
 
@@
 Any and all communication between client and server is both strongly typed
-and made explicit by the use of the `onServer` function, with the programmer
-having complete control over the serialization and de-serialization of data
-using the appropriate type classes. Aside from the obvious advantages of type
-safety, making the crossing of the network boundary explicit helps the
+and made explicit by the use of the \texttt{onServer} function, with the
+programmer having complete control over the serialization and de-serialization
+of data using the appropriate type classes. Aside from the obvious advantages
+of type safety, making the crossing of the network boundary explicit helps the
 programmer make an informed decision as to when and where server communication
@@
 as to how they may want to structure their programs. Should this model prove
-insufficiently expressive, extending our existing model with an `onClient`
-counterpart to `onServer` would be relatively straightforward.
+insufficiently expressive, extending our existing model with an
+\texttt{onClient} counterpart to \texttt{onServer} would be relatively
+straightforward.
 
@@
 
+\bibitem[Marlow, Peyton Jones (2004)]{fastcurry}
+S. Marlow, S. Peyton Jones: Making a fast curry: push/enter vs. eval/apply for higher order languages, 2004
+
 \bibitem[Nazarov (2012)]{ghcjs}
@@
 
+\bibitem[Peyton Jones (2001)]{ffi}
+S. Peyton Jones: Tackling the awkward squad: monadic input/output, concurrency, exceptions, and foreign-language calls in Haskell
+
 \end{thebibliography}

### 2014-02-26 bad.js
Add some example code, to be included in paper later.

@@ -1,0 +1,20 @@
+window.onload = function() {
+    var logbox = document.getElementById('log');
+    var msgbox = document.getElementById('message');
+    var sock = new WebSocket('ws://localhost:24601');
+
+    sock.onmessage = function(e) {
+        logbox.value = e.data + '\n' + logbox.value;
+    };
+
+    sock.onopen = function(e) {
+        sock.send('helo');
+        msgbox.addEventListener('keydown', function(e) {
+            if(e.keyCode == 13) {
+                var msg = msgbox.value;
+                msgbox.value = '';
+                sock.send('text' + msg);
+            }
+        });
+    };
+};

### 2014-02-26 haste.hs

@@ -1,0 +1,37 @@
+import Haste.App
+import Haste.App.Concurrent as HC
+import qualified Control.Concurrent as CC
+import qualified Data.List as L
+
+main = runApp (defaultConfig "ws://localhost:24601" 24601) $ do
+  clients <- liftServerIO $ CC.newMVar []
+
+  hello <- export $ mkUseful clients >>= \clients -> do
+    sid <- getSessionID
+    liftIO $ CC.modifyMVar clients $ \cs -> do
+      mv <- CC.newEmptyMVar
+      return ((sid, mv):cs ,())
+
+  awaitMsg <- export $ mkUseful clients >>= \clients -> do
+    sid <- getSessionID
+    liftIO $ do
+      cs <- CC.withMVar clients return
+      case L.find ((== sid) . fst) cs of
+        Just (_, mv) -> CC.takeMVar mv
+        _            -> fail "Client did not say hello - abort session!"
+
+  sendMsg <- export $ \msg -> mkUseful clients >>= \clients -> liftIO $ do
+    recipients <- CC.withMVar clients return
+    mapM_ (CC.forkIO . flip CC.putMVar msg . snd) recipients
+
+  runClient $ withElems ["log", "message"] $ \[log, msgbox] -> do
+    onServer hello
+
+    mbox <- HC.statefully [] $ \oldlines newline -> do
+      setProp log "value" $ unlines $ newline:oldlines
+      return . Just $ newline:oldlines
+    HC.fork . forever $ mbox <! onServer awaitMsg
+    
+    msgbox `onEvent` OnKeyPress $ \13 -> do
+      getProp msgbox "value" >>= onServer . (sendMsg <.>)
+      setProp msgbox "value" ""

### 2014-02-26 server.hs

@@ -1,0 +1,30 @@
+import qualified Network.WebSockets as WS
+import qualified Control.Concurrent as Conc
+import qualified Control.Exception as Ex
+import qualified Data.ByteString as BS
+import qualified Data.ByteString.UTF8 as BS
+
+main = do
+  clients <- Conc.newMVar []
+  nextClientID <- Conc.newMVar 0
+  WS.runServer "0.0.0.0" 24601 $ \pending -> do
+    conn <- WS.acceptRequest pending
+    cid <- Conc.modifyMVar nextClientID $ \cid -> do
+      return (cid+1, cid)
+
+    let endSession :: Ex.SomeException -> IO ()
+        endSession e = Conc.modifyMVar clients $ \cs -> do
+          return ([c | c <- cs, fst c /= cid],())
+
+    flip Ex.catch endSession $ do
+      let loop = do
+            msg <- WS.receiveData conn :: IO BS.ByteString
+            case splitAt 4 $ BS.toString msg of
+              ("helo", "") -> do
+                Conc.modifyMVar clients $ \cs -> return ((cid,conn):cs,())
+              ("text", msg) -> do
+                recipients <- Conc.withMVar clients return
+                flip mapM_ recipients $ do
+                  \(_,r) -> Conc.forkIO $ WS.sendTextData r (BS.fromString msg)
+            loop
+        in loop

### 2014-02-26 paper.tex
Add stuff on unsafety in the face of unsafePerformIO/IO-dependent free vars.

@@
 
+While our solution completely safe from the invocation point of the
+\texttt{runApp} function onward, capturing IO-dependent variables introduced
+before this invocation may lead to the client and server execution paths
+diverging. Consider the following example:
+
+\begin{verbatim}
+main = do
+  captured <- randomRIO (False, True)
+
+  runApp config $ do
+    diverged <- export $ \\transmitted -> do
+      if captured == transmitted
+        then return ``All OK''
+        else return ``Diverged''
+
+    runClient $ do
+      divergence <- onServer (diverged <.> captured)
+      alert divergence
+\end{verbatim}
+
+The dialog box displayed by this program will read ``All OK'' roughly half
+of the times it is run, while in half of the runs it will read ``Diverged''.
+The reason for this is that the \texttt{randomRIO} computation will be executed
+twice - once on the client, and once on the server. Whenever the computation
+returns differening results, the client and the server will have conflicting
+opinions on the actual value of \texttt{captured}, causing any code
+which - quite reasonably - assumes that variables are consistent between
+client and server to fail.
+
+This is easily remedied however, by a slight compiler modification. Instead of
+using \texttt{main :: IO ()} as the program entry point, introduce a function
+\texttt{appMain :: App Done} for this purpose, and let the runtime system start
+the application by invoking \texttt{runApp config appMain}, where
+\texttt{config} contains the URL and port configuration for the server, given
+by the programmer using a compiler flag. As this modification only affects the
+program's entry point, which is already given ``special'' treatment in Haskell
+compilers, we believe this to be a practical, non-intrusive way to remove this
+limitation, and in fact made this modification to the Haste compiler for our
+implementation.
+
+The astute reader may notice that the same applies to code using
+\texttt{unsafePerformIO} to declare IO-dependent top-level values, and that
+this problem is unaffected by changing the type of the program's entry point.
+While this is indeed true, we feel that since it is quite possible to break
+any Haskell program using \texttt{unsafePerformIO}, even subverting the type
+system, this is not a problem with our solution.
+
 \section{Future work}

### 2014-02-26 paper.tex
Add some macros and a bit of text about Haste.Foreign.

--- 2014-02-26 paper.tex
@@
 \usepackage{amsmath}
+\input{macros}
 
@@
 
-\authorinfo{Koen Claessen ans Anton Ekblad}
+\authorinfo{Koen Claessen and Anton Ekblad}
            {Chalmers University of Technology}
@@
 Javascript programs as possible. The numeric types are represented using the
-Javascript \texttt{Number} type, which is defined as the IEEE754 double
+Javascript \lstinline!Number! type, which is defined as the IEEE754 double
 precision floating point type. This obviously adds some overhead to operations
@@
 with the first element representing the data constructor and the following
-values representing its arguments. For instance, the value \texttt{42 :: Int}
-is represented as \texttt{[0, 42]}, the leading \texttt{0} representing the
-zeroth constructor of the \texttt{Int} type and the \texttt{42}, obviously,
-representing the ``machine'' integer. It may seem strange that a limited
-precision integer is represented using one level of indirection rather than as
-a simple number, but recall that the \texttt{Int} type is defined by GHC as
-\texttt{data Int = I\# Int\#}.
+values representing its arguments. For instance, the value \lstinline!42 :: Int!
+is represented as \lstinline![0, 42]!, the leading \lstinline!0! representing
+the zeroth constructor of the \lstinline!Int! type and the \lstinline!42!,
+obviously, representing the ``machine'' integer. It may seem strange that a
+limited precision integer is represented using one level of indirection rather
+than as a simple number, but recall that the \lstinline!Int! type is defined by
+GHC as \lstinline!data Int = I# Int#!.
 
@@
 eval/apply method described in \cite{fastcurry} instead.
+
+\paragraph{Interfacing with Javascript} While Haste supports the Foreign
+Function Interface inherited from GHC, with its usual features and limitations
+\ \cite{ffi}, it is often impractical to work within the confines of an
+interface designed for communication on a very low level. For this reason
+Haste sports its own method for interacting with Javascript as well, which
+allows the programmer to pass any value back and forth between Haskell and
+Javascript, as long as she can come up with a way to translate this value
+between its Haskell and Javascript representations. Of course, not performing
+any translation at all is also a valid ``translation'', which allows Haskell
+code to store any Javascript value for later retrieval without inspecting it
+and vice versa. The following example implements mutable variables using this
+custom Javascript interface.
+
+\begin{code}
+import Haste.Foreign
+
+-- A MutableVar is completely opaque to Haskell code
+-- and is only ever manipulated in Javascript. Thus,
+-- we use the Unpacked type to represent it,
+-- indicating a completely opaque value.
+newtype MutableVar a = MV Unpacked
+
+instance Marshal (MutableVar a) where
+  pack          = MV
+  unpack (MV x) = x
+
+newMutable :: Marshal a => a -> IO (MutableVar a)
+newMutable = ffi "(function(x) {return {val: x};})"
+
+setMutable :: Marshal a => MutableVar a -> a -> IO ()
+setMutable = ffi "(function(m, x) {m.val = x;})"
+
+getMutable :: Marshal a => MutableVar a -> IO a
+getMutable = ffi "(function(m) {return m.val;})"
+\end{code}
+
+The core of this interface consists of the \lstinline!ffi! function, which
+allows the programmer to create a Haskell function from arbitrary Javascript
+code. This function exploits Javascript's ability to parse and execute
+arbitrary strings at run-time using the \lstinline!eval! function, coupled with
+the fact that functions in Haste and in Javascript share the same
+representation, to dynamically create a function object at runtime.
+The \lstinline!ffi! function is typed using a well-known type class trick for
+creating statically typed variadic functions\ \cite{printf}, and works very
+much like the \lstinline!printf! function of Haskell's standard library.
+When applied to one or more arguments instantiating the \lstinline!Marshal!
+type class, the \lstinline!pack! function is applied to each argument,
+marshalling them into their respective Javascript representations, before they
+are passed to the dynamically created function. When that function returns,
+the inverse \lstinline!unpack! function is applied to its return value before
+it is passed back into Haskell land.
+
+As the marshalling functions chosen for each argument and the foreign
+function's return value depends on its type, the programmer must explicitly
+specify the type of each function imported using \lstinline!ffi!; in this,
+Haste's custom method is no different from the conventional FFI.
+
+There are several benefits to this method, the most obvious being that new
+marshallable types can be added by simply instantiating a type class. Thanks
+to the lazy evaluation employed by Haste, each foreign function object is only
+created once and then cached; any further calls to the same (Haskell) function
+will reuse the cached function object. Implementation-wise, this method is also
+very non-intrusive, requiring only the use of the normal FFI to import
+Javascript's \lstinline!eval! function; no modification of the compiler is
+needed.
 
@@
 regular Javascript, it still suffers from many of its usual drawbacks. In
-particular, the the weak typing of Javascript poses a problem in that the
+particular, the weak typing of Javascript poses a problem in that the
 programmer is in no way reprimanded by her tools for using server APIs
@@
 Any and all communication between client and server is both strongly typed
-and made explicit by the use of the \texttt{onServer} function, with the
+and made explicit by the use of the \lstinline!onServer! function, with the
 programmer having complete control over the serialization and de-serialization
@@
 insufficiently expressive, extending our existing model with an
-\texttt{onClient} counterpart to \texttt{onServer} would be relatively
+\lstinline!onClient! counterpart to \lstinline!onServer! would be relatively
 straightforward.
+
+While our solution completely safe from the invocation point of the
+\lstinline!runApp! function onward, capturing IO-dependent variables introduced
+before this invocation may lead to the client and server execution paths
+diverging. Consider the following example:
+
+\begin{code}
+main = do
+  captured <- randomRIO (False, True)
+
+  runApp config $ do
+    diverged <- export $ \transmitted -> do
+      if captured == transmitted
+        then return ``All OK''
+        else return ``Diverged''
+
+    runClient $ do
+      divergence <- onServer (diverged <.> captured)
+      alert divergence
+\end{code}
+
+The dialog box displayed by this program will read ``All OK'' roughly half
+of the times it is run, while in half of the runs it will read ``Diverged''.
+The reason for this is that the \lstinline!randomRIO! computation will be
+executed twice - once on the client, and once on the server. Whenever the
+computation returns differening results, the client and the server will have
+conflicting opinions on the actual value of \lstinline!captured!, causing any
+code which - quite reasonably - assumes that variables are consistent between
+client and server to fail.
+
+This is easily remedied however, by a slight compiler modification. Instead of
+using \lstinline!main :: IO ()! as the program entry point, introduce a function
+\lstinline!appMain :: App Done! for this purpose, and let the runtime system
+start the application by invoking \lstinline!runApp config appMain!, where
+\lstinline!config! contains the URL and port configuration for the server, given
+by the programmer using a compiler flag. As this modification only affects the
+program's entry point, which is already given ``special'' treatment in Haskell
+compilers, we believe this to be a practical, non-intrusive way to remove this
+limitation, and in fact made this modification to the Haste compiler for our
+implementation.
+
+The astute reader may notice that the same applies to code using
+\lstinline!unsafePerformIO! to declare IO-dependent top-level values, and that
+this problem is unaffected by changing the type of the program's entry point.
+While this is indeed true, we feel that since it is quite possible to break
+any Haskell program using \lstinline!unsafePerformIO!, even subverting the type
+system, this is not a problem with our solution.
 
@@
 
+\bibitem[Taylor (2013)]{printf}
+C. Taylor: Polyvariadic Functions and Printf, http://chris-taylor.github.io/blog/2013/03/01/how-haskell-printf-works/, 2013
+
 \end{thebibliography}

### 2014-02-26 paper.tex
Add the beginning of Jen's story to the introduction.

@@
 called for.
+
+To put this all into perspective, let's take a look at the completely fictious
+case of Jen, a software developer tasked with implementing a simple ``chatbox''
+for a client's web site. Jen is a senior developer, but has no previous
+experience with web applications. She is, however, a competend engineer and a
+fast learner, and so manages to produce the following piece of client-side code.
+
+\begin{code}
+function handshake(sock) {sock.send('helo');}
+function chat(sock, msg) {sock.send('text' + msg);}
+
+window.onload = function() {
+    var logbox = document.getElementById('log');
+    var msgbox = document.getElementById('message');
+    var sock = new WebSocket('ws://example.com:24601');
+
+    sock.onmessage = function(e) {
+        logbox.value = e.data + NEWLINE + logbox.value;
+    };
+
+    sock.onopen = function(e) {
+        handshake(sock);
+        msgbox.addEventListener('keydown', function(e) {
+            if(e.keyCode == 13) {
+                var msg = msgbox.value;
+                msgbox.value = '';
+            }
+        });
+    };
+};
+\end{code}
+
+Since the ``chatbox'' application is very simple - users should only be able to
+send and receive text messages in real time - Jen opts for a very simple
+design. Two UI elements, \lstinline!logbox! and \lstinline!msgbox!, represent
+the chat log and the text area where the user inputs their messages
+respectively. When a message arrives, it is prepended to the chat log, making
+the most recent message appear at the of the log window, and when the user hits
+the return key in the input text box the message contained therein is sent and
+the input text box is cleared.
+
+Messages are transmitted as strings, with the initial four characters
+indicating the type of the message and the rest being the optional payload.
+There are only two messages; a handshake indicating that a user wants to join
+the conversation, and a broadcast message which sends a line of text to all
+connected users via the server.
+
+The code looks solid enough, but Jen is not happy. This simple piece of code
+contains no less than three asynchronous callbacks, making the program flow
+non-obvious. Moreover, she is concerned about the network communication. The
+protocol will clearly need to be redesigned if the application is to be
+upgraded with new features later, but the client is not willing to pay Jen to
+spend development time on features he has not requested and perhaps never will.
+If the protocol will need to be updated in the future, how much time will Jen
+need to spend on ensuring that all code using the network protocol stays
+updated? Then there is the server component to think; of how much extra work
+will it take to ensure that the two code bases stay in sync? Coming from a
+desktop background, Jen wonders why something so seemingly simple should
+involve client/server architectures and network protocols at all.
 

### 2014-02-27 paper.tex
Add Conclusion section, so I dont forget that part later.
Put Jen's Javascript in a figure instead of inline. So pretty!

@@
 experience with web applications. She is, however, a competend engineer and a
-fast learner, and so manages to produce the following piece of client-side code.
+fast learner, and so manages to produce the following client-side code in
+figure \ref{lst:javascript-client}.
 
+\begin{listingfloat}
 \begin{code}
@@
         msgbox.value = '';
+        chat(msg);
       }
@@
 \end{code}
+\caption{Javascript chatbox implementation}
+\label{lst:javascript-client}
+\end{listingfloat}
 
@@
 protocol will clearly need to be redesigned if the application is to be
-upgraded with new features later, but the client is not willing to pay Jen to
+enhanced with new features later, but the client is not willing to pay Jen to
 spend development time on features he has not requested and perhaps never will.
+
 If the protocol will need to be updated in the future, how much time will Jen
 need to spend on ensuring that all code using the network protocol stays
-updated? Then there is the server component to think; of how much extra work
-will it take to ensure that the two code bases stay in sync? Coming from a
-desktop background, Jen wonders why something so seemingly simple should
-involve client/server architectures and network protocols at all.
+updated? Then there is the server component to think of; how much extra work
+will it take to ensure that the two code bases stay in sync? How much code
+will need to be written twice, one version for the client and one for the
+server, due to the unfortunate fact that the two can't share any code?
+
+Coming from a desktop background, Jen wonders why something so seemingly simple
+should involve client/server architectures and network protocols at all.
 
@@
 
+\section{Conclusion}
+
 \acks

### 2014-02-27 paper.tex
Jen wasn't serious enough, so she had to go.

--- 2014-02-26 paper.tex
@@
 called for.
+
+To put make all a bit more concrete, let's take a look at an example. Let's say
+what we want to implement a simple ``chatbox'' component for a website, to allow
+visitors to discuss the site's content in real time. Using mainstream
+development practives and recent technologies such as WebSockets, we may come
+up with something like the program in figure \ref{lst:javascript-client}.
+
+\begin{listingfloat}
+\begin{code}
+function handshake(sock) {sock.send('helo');}
+function chat(sock, msg) {sock.send('text' + msg);}
+
+window.onload = function() {
+  var logbox = document.getElementById('log');
+  var msgbox = document.getElementById('message');
+  var sock = new WebSocket('ws://example.com:24601');
+
+  sock.onmessage = function(e) {
+    logbox.value = e.data + NEWLINE + logbox.value;
+  };
+
+  sock.onopen = function(e) {
+    handshake(sock);
+    msgbox.addEventListener('keydown', function(e) {
+      if(e.keyCode == 13) {
+        var msg = msgbox.value;
+        msgbox.value = '';
+        chat(msg);
+      }
+    });
+  };
+};
+\end{code}
+\caption{Javascript chatbox implementation}
+\label{lst:javascript-client}
+\end{listingfloat}
+
+Since the ``chatbox'' application is very simple - users should only be able to
+send and receive text messages in real time - we opt for a very simple
+design. Two UI elements, \lstinline!logbox! and \lstinline!msgbox!, represent
+the chat log and the text area where the user inputs their messages
+respectively. When a message arrives, it is prepended to the chat log, making
+the most recent message appear at the of the log window, and when the user hits
+the return key in the input text box the message contained therein is sent and
+the input text box is cleared.
+
+Messages are transmitted as strings, with the initial four characters
+indicating the type of the message and the rest being the optional payload.
+There are only two messages; a handshake indicating that a user wants to join
+the conversation, and a broadcast message which sends a line of text to all
+connected users via the server.
+
+This code looks solid enough by web standards, but even this simple piece of
+code contains no less than three asynchronous callbacks, two of which both read
+and and modify the application's global state. This makes the program flow
+non-obvious, and introduces unnecessary risk and complexity through the
+haphazard state modifications.
+
+Moreover, this code is not very extensible. If this simple application is to be
+enhanced with new features down the road, the network protocol will clearly
+need to be redesigned. However, if we were developing this application for a
+client, said client would likely not want to pay the added cost for the design
+and implementation of features she did not ask for, and perhaps never will.
+
+Should the protocol need updating in the future, how much time will we need to
+spend on ensuring that the protocol is used properly across our entire program,
+and how much extra work will it take to keep the client and server in sync?
+How much code will need to be written twice, once for the client and once for
+the server, due to the unfortunate fact that the two parts are implemented as
+separate programs, possibly in separate languages?
+
+Above all, is it really necessary for such a simple program to involve
+client/server architectures and network protocol design at all?
 

### 2014-02-27 paper.tex
Add general description of our solution plus Hello Server example.

@@
   var msgbox = document.getElementById('message');
-  var sock = new WebSocket('ws://example.com:24601');
+  var sock = new WebSocket('ws://example.com);
 
@@
 \section{A seamless programming model}
+
+There are many conveivable improvements to the mainstream web development model
+described in the previous section. We propose an alternative programming model
+based on Haskell, in which web applications are written as monolithic entities
+rather than as two independent parts that just so happen to talk to each other.
+
+Our proposed model, dubbed ``Haste.App'' in an ingenious attempt at
+capitalizing on the recent popularity of the ``app'' buzzword, has the
+following properties:
+
+\begin{itemize}
+  \item The programming model is synchronous, giving the programmer a simple,
+        linear view of the program flow, eliminating the need to program with
+        callbacks and continuations.
+  \item Side-effecting code is explicitly designated to run on either the
+        client or the server using the type system while pure code can be
+        shared by both. Additionally, both client and server code may lift
+        general IO computations, allowing for safe IO code reuse within the
+        confines of the client or server designated code sections.
+  \item Client-server network communication is handled through statically typed
+        RPC function calls, extending the reach of Haskell's type checker over
+        the network and giving the programmer advance warning when she uses
+        network services incorrectly or forgets to update communication code
+        as the application's internal protocol changes.
+  \item The implementation is built as a library on top of the GHC and Haste
+        Haskell compilers, requiring little to no compiler support.
+\end{itemize}
+
+\paragraph{A first example} While explaining of the properties of our solution
+is all well and good, nothing compares to a good old Hello World example to
+convey the idea. We begin by implementing a function which prints a greeting
+to the server's console.
+
+\begin{code}
+import Haste.App
+
+helloServer :: String -> Server ()
+helloServer name =
+  liftIO $ putStrLn (name ++ " says hello!")
+\end{code}
+
+Computations exclusive to the server side live in the \lstinline!Server! monad.
+This is basically an IO monad with a few extra operations for session handling;
+its main purpose is to prevent the programmer from accidentally attempting to
+perform client-exclusive operations, such as popping up a browser dialog box,
+on the server.
+
+Next, we need to make the \lstinline!helloServer! function available as an RPC
+function and call it from the client.
+
+\begin{code}
+appMain :: App Done
+appMain = do
+  greetings <- export helloServer
+
+  runClient $ do
+    name <- prompt "Hi there, what is your name?"
+    onServer (greetings <.> name)
+\end{code}
+
+The \lstinline!export! function takes an arbitrary function, provided that all
+its arguments as well as its return value are serializable through the aptly
+named \lstinline!Serializable! type class, and returns a typed ``handle'' of
+sorts. In this example, the type of \lstinline!greetings! is
+\lstinline!Export (String -> Server ())!, indicating that the handle identifies
+an exported function with a single \lstinline!String! argument and no return
+value; exported functions all live in the \lstinline!Server! monad.
+
+In order to make an RPC call using this handle, we must supply it with an
+argument. This is done using the \lstinline!<.>! operator. Upon closer
+inspection, the attentive reader may notice that its type,
+\linebreak
+\lstinline!Serialize a => Export (a -> b) -> a -> Export b!,
+is very similar to the type of the \lstinline!<*>! operator over applicative
+functors. This is not a coincidence; the \lstinline!<.>! performs the same role
+for the \lstinline!Export! type as \lstinline!<*>! performs for applicative
+functors. The reason for using a separate operator for this instead of making
+\lstinline!Export! an instance of \lstinline!Applicative! is that since
+functions embedded in the \lstinline!Export! type exist only to be called over
+a network, such functions must only be applied to arguments which can be
+serialized and sent over a network connection.
+
+After applying the value obtained by \lstinline!prompt!ing the user for her
+name to the exported function, we apply the \lstinline!onServer! function to
+the result, which dispatches the RPC call to the server. \lstinline!onServer!
+will then block until the RPC call returns.
+
+Finally, we need to launch our application, and inform the client part of the
+whereabouts of its server counterpart; in this example the server will listen
+on port 1111, and since we will only run this application locally the client
+should attempt to contact the server at our local machine.
+
+\begin{code}
+main :: IO ()
+main =
+  runApp (mkConfig "ws://localhost:1111" 1111) appMain
+\end{code}
+
+This part is entirely straightforward; the URL of the server is provided for
+the benefit of the client and the port to listen on is provided for the
+benefit of the server. The final argument is the main function of our
+application. Figure \ref{lst:hello-server} gives the code of this example in
+its entirety.
+
+\begin{listingfloat}
+\begin{code}
+import Haste.App
+
+helloServer :: String -> Server ()
+helloServer name =
+  liftIO $ putStrLn (name ++ " says hello!")
+
+appMain :: App Done
+appMain = do
+  greetings <- export helloServer
+
+  runClient $ do
+    name <- prompt "Hi there, what is your name?"
+    onServer (greetings <.> name)
+
+main :: IO ()
+main =
+  runApp (mkConfig "ws://localhost:1111" 1111) appMain
+\end{code}
+\caption{A seamless programming model: Hello Server}
+\label{lst:hello-server}
+\end{listingfloat}
+
 \section{Implementation}
@@
 program's entry point, which is already given ``special'' treatment in Haskell
-compilers, we believe this to be a practical, non-intrusive way to remove this
-limitation, and in fact made this modification to the Haste compiler for our
-implementation.
+compilers, and may even be implemented using the plugin mechanism of GHC,
+requiring no modification of the compiler code at all, we believe this to be a
+practical, non-intrusive way to remove this limitation, and are planning on
+including this modification in our implementation.
 

### 2014-02-27 paper.tex
Describe runApp, runClient, Client.

@@
 Computations exclusive to the server side live in the \lstinline!Server! monad.
-This is basically an IO monad with a few extra operations for session handling;
-its main purpose is to prevent the programmer from accidentally attempting to
-perform client-exclusive operations, such as popping up a browser dialog box,
-on the server.
+This is basically an IO monad, as can be seen from the regular
+\lstinline!putStrLn! \lstinline!IO! computation being lifted into it, with a
+few extra operations for session handling; its main purpose is to prevent the
+programmer from accidentally attempting to perform client-exclusive operations,
+such as popping up a browser dialog box, on the server.
 
@@
 
-In order to make an RPC call using this handle, we must supply it with an
-argument. This is done using the \lstinline!<.>! operator. Upon closer
-inspection, the attentive reader may notice that its type,
+After the \lstinline!export! call, we enter the domain of client-exclusive code
+with the application of \lstinline!runClient!. This function, unsurprisingly,
+execute computations in the \lstinline!Client! monad which is essentially an
+IO monad with cooperative multitasking added on top, to mitigate the fact that
+Javascript has no native concurrency support. \lstinline!runClient! does not
+return, and is the only function with a return type of \lstinline!App Done!,
+which ensures that each \lstinline!App! computation contains exactly one client
+computation.
+
+In order to make an RPC call using a handle obtained from \lstinline!export!,
+we must supply it with an argument. This is done using the \lstinline!<.>!
+operator. Upon closer inspection, the attentive reader may notice that
 \linebreak
-\lstinline!Serialize a => Export (a -> b) -> a -> Export b!,
+its type, \lstinline!Serialize a => Export (a -> b) -> a -> Export b!,
 is very similar to the type of the \lstinline!<*>! operator over applicative
@@
 its entirety.
+
+Like \lstinline!runClient!, \lstinline!runApp! does not return, and is intended
+to be the only computation executed by \lstinline!main!. In fact, performing
+non-deterministic IO and allowing \lstinline!runApp! or its arguments to
+capture the result may lead to unpredictable behavior. A remedy for this is
+discussed in section \ref{sec:discussion}.
 
@@
 
-\section{Discussion and related work}
+\section{Discussion and related work}\label{sec:discussion}
 

### 2014-02-27 paper.tex
Add blurb about our solution being client centric.
Introduce some subsections.

@@
 Several attempts have been made at tackling this problem, but none is, in our
-view, satisfactory. In this paper, we propose a programming model in which a
-web application is written as a single program from which client and server
-binaries are generated during compilation. Type annotations in the source
+view, satisfactory. In this paper, we propose a functional programming model in
+which a web application is written as a single program from which client and
+server binaries are generated during compilation. Type annotations in the source
 program control which parts are executed on the server and which are executed
@@
         as the application's internal protocol changes.
+  \item Our model takes the view that the client side is the main driver when
+        developing web applications and accordingly delegates the server to
+        the role of a computational and/or storage resource, tasked with
+        servicing client requests rather than driving the program.
   \item The implementation is built as a library on top of the GHC and Haste
@@
 
-\paragraph{A first example} While explaining of the properties of our solution
-is all well and good, nothing compares to a good old Hello World example to
-convey the idea. We begin by implementing a function which prints a greeting
-to the server's console.
+\subsection{A first example}
+
+While explaining of the properties of our solution is all well and good,
+nothing compares to a good old Hello World example to convey the idea.
+We begin by implementing a function which prints a greeting to the server's
+console.
 
@@
 
-\paragraph{Choosing a compiler} Haste is by no means the only
-Javascript-targeting compiler for a purely functional language. In particular,
-the GHC-based GHCJS\ \cite{ghcjs} and UHC\ \cite{uhc} compilers are both
-capable of compiling standard Haskell into Javascript; the Fay\ \cite{fay}
-language was designed from the ground up to target the web space using a subset
-of Haskell; and there exist solutions for compiling Erlang\ \cite{jserlang} and
-Clean\ \cite{jsclean} to Javascript as well.
+\subsection{Choosing a compiler}
+
+Haste is by no means the only Javascript-targeting compiler for a purely
+functional language. In particular, the GHC-based GHCJS\ \cite{ghcjs} and
+UHC\ \cite{uhc} compilers are both capable of compiling standard Haskell into
+Javascript; the Fay\ \cite{fay} language was designed from the ground up to
+target the web space using a subset of Haskell; and there exist solutions for
+compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to Javascript as
+well.
 
@@
 
-\paragraph{Design of the Haste compiler} Haste is first and foremost intended
-to be a practical compiler; the ability to produce code which is lean and fast
-enough to be used in the real world is critical. To this end, we make several
-design choices and tradeoffs, some of which would be viewed as detrimental in a
-more purist context.
+\subsection{Design of the Haste compiler}
+
+Haste is first and foremost intended to be a practical compiler; the ability
+to produce code which is lean and fast enough to be used in the real world is
+critical. To this end, we make several design choices and tradeoffs, some of
+which would be viewed as detrimental in a more purist context.
 
@@
 
-\paragraph{Implementation} As previously mentioned, Haste offloads much of the
-heavy lifting of compilation - parsing, type checking, intermediate code
-generation and many optimizations - onto GHC, and takes over code generation
-after the STG generation step, at the very end of the compilation process.
-STG\ \cite{stg} is the last intermediate representation used by GHC before
-the final code generation takes place and has several benefits for use as
-Haste's target language:
+\subsection{Implementation overview}
+
+As previously mentioned, Haste offloads much of the heavy lifting of
+compilation - parsing, type checking, intermediate code generation and many
+optimizations - onto GHC, and takes over code generation after the STG
+generation step, at the very end of the compilation process. STG\ \cite{stg} is
+the last intermediate representation used by GHC before the final code
+generation takes place and has several benefits for use as Haste's target
+language:
 
@@
 
+\subsection{Data representation}
+
 The runtime data representation of Haste programs is kept as close to regular
@@
 
-\paragraph{Interfacing with Javascript} While Haste supports the Foreign
-Function Interface inherited from GHC, with its usual features and limitations
-\ \cite{ffi}, it is often impractical to work within the confines of an
-interface designed for communication on a very low level. For this reason
-Haste sports its own method for interacting with Javascript as well, which
-allows the programmer to pass any value back and forth between Haskell and
-Javascript, as long as she can come up with a way to translate this value
-between its Haskell and Javascript representations. Of course, not performing
-any translation at all is also a valid ``translation'', which allows Haskell
-code to store any Javascript value for later retrieval without inspecting it
-and vice versa. The following example implements mutable variables using this
-custom Javascript interface.
+\subsection{Interfacing with Javascript}
 
+While Haste supports the Foreign Function Interface inherited from GHC, with
+its usual features and limitations \ \cite{ffi}, it is often impractical to
+work within the confines of an interface designed for communication on a very
+low level. For this reason Haste sports its own method for interacting with
+Javascript as well, which allows the programmer to pass any value back and
+forth between Haskell and Javascript, as long as she can come up with a way to
+translate this value between its Haskell and Javascript representations.
+Of course, not performing any translation at all is also a valid
+``translation'', which allows Haskell code to store any Javascript value for
+later retrieval without inspecting it and vice versa. The example given in
+figure \ref{lst:ffi} implements mutable variables using this custom Javascript
+interface.
+
+\begin{listingfloat}
 \begin{code}
@@
 \end{code}
+\caption{Mutable variables with \lstinline!Haste.Foreign!}
+\label{lst:ffi}
+\end{listingfloat}
 
@@
 \section{Discussion and related work}\label{sec:discussion}
+
+\subsection{Related work}
 
@@
 
-\paragraph{Limitations} Unlike any of the related works, our approach takes a
-firm stand, regarding the client as the driver in the client-server
+\subsection{Limitations}
+
+\paragraph{Client-centricity} Unlike any of the related works, our approach
+takes a firm stand, regarding the client as the driver in the client-server
 relationship with the server taking on the role of a passive computational
@@
 
-While our solution completely safe from the invocation point of the
-\lstinline!runApp! function onward, capturing IO-dependent variables introduced
-before this invocation may lead to the client and server execution paths
-diverging. Consider the following example:
+\paragraph{Preventing divergence} While our solution completely safe from the
+invocation point of the \lstinline!runApp! function onward, capturing
+IO-dependent variables introduced before this invocation may lead to the client
+and server execution paths diverging. Consider the following example:
 
@@
 system, this is not a problem with our solution.
+
+\paragraph{Environment consistency} As our programming model uses two different
+compilers to generate client and server code, it is crucial to keep the package
+environments of the two in sync. A situation where, for instance, a module is
+visible to one compiler but not to the other will render many programs
+uncompilable until this inconsistency is fixed.
+
+This kind of divergence can be worked around using conditional compilation, but
+is highly problematic even so; using a unified package database between the two
+compilers, while problematic due to the differing natures of native and
+Javascript compilation respectively, would be a significant improvement in this
+area.
 

### 2014-02-27 paper.tex
Add subsection about server-side state.

@@
   \item The implementation is built as a library on top of the GHC and Haste
-        Haskell compilers, requiring little to no compiler support.
+        Haskell compilers, requiring little to no compiler support. Programs
+        are compiled twice; once with Haste and once with GHC, to produce the
+        final client and server side code respectively.
 \end{itemize}
 
-\subsection{A first example}
+\subsection{A first example}\label{sec:helloserver}
 
@@
 non-deterministic IO and allowing \lstinline!runApp! or its arguments to
-capture the result may lead to unpredictable behavior. A remedy for this is
-discussed in section \ref{sec:discussion}.
+capture the result may lead to unpredictable behavior. The \lstinline!appMain!
+function should for all intents and purposes be regarded as the program's entry
+point; a non-intrusive method of enforcing this using the type system is
+described in section \ref{sec:limitations}, but is not used in the examples
+given here as it relies on the use of the GHC plugin system.
 
@@
 \label{lst:hello-server}
+\end{listingfloat}
+
+\subsection{Using server-side state}
+
+While the Hello Server example illustrates how client-server communication is
+handled, most web applications need to keep some server-side state as well.
+At first glance, this would seem to be a trivial problem; can we not just
+create our state holding elements in \lstinline!main! as usual?
+Figure \ref{lst:bad-state} describes an attempt at a solution using this idea.
+
+\begin{listingfloat}
+\begin{code}
+main = do
+  visitorref <- newIORef 0
+  runApp config $ do
+    count <- export . liftIO $ do
+      atomicModifyIORef visitorref (\v -> (v+1, v+1))
+
+    runClient $ do
+      visitors <- onServer count
+      alert ("Your are visitor #" ++ show visitors)
+\end{code}
+\caption{Server-side state: a misguided first attempt}
+\label{lst:bad-state}
+\end{listingfloat}
+
+While this may at first seem to solve the problem of server-side state, this
+approach raises a number of questions. Does \lstinline!visitorref! live on the
+client or the server, or perhaps on both? What happens if our state holding
+element is initialized to a non-deterministic value, and what if we assign it
+a value on the server and then attempt to read that value on the client?
+Recall that, as explained in section \ref{sec:helloserver}, the function
+passed to \lstinline!runApp! should for all intents and purposes be regarded as
+the program's entry point.
+
+To allow the use of server state while treating the final argument to
+\lstinline!runApp! (from here on always referred to as \lstinline!appMain!)
+as the program's entry point, we may want to allow arbitrary IO computations to
+be lifted into the \lstinline!App! monad:
+
+\begin{code}
+main = runApp config $ do
+  visitorref <- liftIO $ newIORef 0
+  ...
+\end{code}
+
+However, this does not answer the question of where \lstinline!visitorref!
+lives or where any side effects caused by its creation will take place.
+To solve this problem once and for all, we need to introduce a way to lift
+arbitrary IO computations, but ensure that said computations are executed on
+the server and nowhere else. This is accomplished using a more restricted
+version of \lstinline!liftIO!:
+
+\begin{code}
+liftServerIO :: IO a -> App (Useless a)
+\end{code}
+
+In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
+is wrapped in the \lstinline!Useless! type. As the name implies, this type is
+completely opaque and thus useless unless you can somehow extract its contained
+value, something which is only possible using a single function:
+
+\begin{code}
+mkUseful :: Useless a -> Server a
+\end{code}
+
+Therein lies the secret to restricting \lstinline!liftServerIO! to the
+server: the only function capable of turning a \lstinline!Useless! value into
+a useful one lives in the server monad. Any client side code is thus free to
+completely ignore executing computations lifted using \lstinline!liftServerIO!;
+since the result of such a computation is never observable on the client,
+the client has no obligation to even produce such a value.
+Figure \ref{lst:good-state} shows how to make proper use of server-side state.
+
+\begin{listingfloat}
+\begin{code}
+main =
+  runApp config $ do
+    uselessref <- liftServerIO $ newIORef 0
+
+    count <- export $ do
+      r <- mkUseful uselessref
+      liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))
+
+    runClient $ do
+      visitors <- onServer count
+      alert ("Your are visitor #" ++ show visitors)
+\end{code}
+\caption{Server-side state: doing it properly}
+\label{lst:good-state}
 \end{listingfloat}
@@
 
-\section{Discussion and related work}\label{sec:discussion}
+\section{Discussion and related work}
 
@@
 \subsection{Limitations}
+\label{sec:limitations}
 
@@
 IO-dependent variables introduced before this invocation may lead to the client
-and server execution paths diverging. Consider the following example:
+and server execution paths diverging. Recall that our solution does not modify
+either of the compilers used. As such, both will produce code to execute all
+computations called by \lstinline!main!. Consider the following example:
 

### 2014-02-27 haste.hs
De-uglify haste.hs slightly.

@@ -1,7 +1,7 @@
+import Data.List
 import Haste.App
-import Haste.App.Concurrent as HC
+import Haste.App.Concurrent
 import qualified Control.Concurrent as CC
-import qualified Data.List as L
 
 main = runApp (defaultConfig "ws://localhost:24601" 24601) $ do
   clients <- liftServerIO $ CC.newMVar []
@@ -16,7 +16,7 @@ main = runApp (defaultConfig "ws://localhost:24601" 24601) $ do
     sid <- getSessionID
     liftIO $ do
       cs <- CC.withMVar clients return
-      case L.find ((== sid) . fst) cs of
+      case find ((== sid) . fst) cs of
         Just (_, mv) -> CC.takeMVar mv
         _            -> fail "Client did not say hello - abort session!"
 
@@ -27,10 +27,10 @@ main = runApp (defaultConfig "ws://localhost:24601" 24601) $ do
   runClient $ withElems ["log", "message"] $ \[log, msgbox] -> do
     onServer hello
 
-    mbox <- HC.statefully [] $ \oldlines newline -> do
+    mbox <- statefully [] $ \oldlines newline -> do
       setProp log "value" $ unlines $ newline:oldlines
       return . Just $ newline:oldlines
-    HC.fork . forever $ mbox <! onServer awaitMsg
+    fork . forever $ mbox <! onServer awaitMsg
     
     msgbox `onEvent` OnKeyPress $ \13 -> do
       getProp msgbox "value" >>= onServer . (sendMsg <.>)

### 2014-02-27 paper.tex
Add checklist for implementation section.

@@
 a network, such functions must only be applied to arguments which can be
-serialized and sent over a network connection.
+serialized and sent over a network connection. When an \lstinline!Export!
+function is applied to an argument using \lstinline!<.>!, the argument is
+serialized and stored inside the resulting \lstinline!Export! object, awaiting
+dispatch.
 
@@
 
+\subsection{The chatbox, revisited}
+
 \section{Implementation}
+
+\begin{itemize}
+  \item Poor man's concurrency monad
+  \item Conditional compilation
+  \item Printf trick to turn export into [JSON] -> f
+  \item JSON over WebSockets used for transport
+  \item Server event loop awaits commands, dispatches functions, sends data back
+  \item Client sends command with nonce and blocks on MVar, event loop fills
+        MVar when return value with proper nonce arrives
+  \item Throw away unused stuff on client
+\end{itemize}
+
 \section{The Haste compiler}

### 2014-02-27 paper.tex
Add conclusion section.

@@
 of data using the appropriate type classes. Aside from the obvious advantages
-of type safety, making the crossing of the network boundary explicit helps the
-programmer make an informed decision as to when and where server communication
-is appropriate.
+of type safety, making the crossing of the network boundary explicit aids the
+programmer in making an informed decision as to when and where server
+communication is appropriate, as well as helps prevents accidental transmission
+of sensitive information intended to stay on the client.
 
@@
 
+We have presented a programming model which improves on the current state of
+the art in client-server web application development. In particular, our
+solution combines type safe communication between the client and the server
+with purely functional semantics, clear demarcations as to when data is
+transmitted and where a particular piece of code is executed, and the ability to
+effortlessly share code between the client and the server.
+
+Our model is based on a simple blocking concurrency model rather than explicit
+continuations, is well suited for use with a GUI programming style based on
+communicating processes with local state, and requires essentially no
+modification of existing tools or compilers.
+
 \acks

### 2014-02-27 paper.tex
Remove redundant paragraph in intro.

@@
 
-Having your application rely on some network protocol for
-its internal functionality involves writing a lot of boilerplate network code
-and some serious advance planning of your application's features to be able to
-design a suitable protocol, hampering incremental development. Were we to
-discount this initial design burden, ensuring that the protocol is used
-correctly throughout the code base as the application evolves and gains new
-features would still put an ever increasing burden on the developer.
-
-This state of things is not a concious design choice - most web applications
+This state of things is not a conscious design choice - most web applications
 are conceptually a single entity, not two programs which just happen to talk
@@
 
-To put make all a bit more concrete, let's take a look at an example. Let's say
-what we want to implement a simple ``chatbox'' component for a website, to allow
-visitors to discuss the site's content in real time. Using mainstream
-development practives and recent technologies such as WebSockets, we may come
-up with something like the program in figure \ref{lst:javascript-client}.
+To put make the problem a bit more concrete, let's take a look at an example.
+Let's say what we want to implement a simple ``chatbox'' component for a
+website, to allow visitors to discuss the site's content in real time. Using
+mainstream development practives and recent technologies such as WebSockets,
+we may come up with something like the program in figure
+\ref{lst:javascript-client}.
 
@@
 the conversation, and a broadcast message which sends a line of text to all
-connected users via the server.
+connected users via the server. The only messages received from the server are
+new chat messages, delivered as simple strings.
 
@@
 client, said client would likely not want to pay the added cost for the design
-and implementation of features she did not ask for, and perhaps never will.
+and implementation of features she did not - and perhaps never will - ask for.
 
@@
 \subsection{The chatbox, revisited}
+
+Now that we have seen how to implement both network communication, we are ready
+to revisit the chatbox program, this time using our improved programming model.
 

### 2014-02-27 paper.tex
Fix some INTERNETS references.

@@
 \bibitem[Conductance application server (2014)]{conductance}
-THE INTERNETS: http://conductance.io
+The Conductance application server, http://conductance.io
 
@@
 \bibitem[Domoszlai et al (2011)]{jsclean}
-L. Domoszlai, E. Bruël, J. M. Jansen: Implementing a non-strict purely functional language in JavaScript, in Acta Universitatis Sapientiae, 2011
+L. Domoszlai, E. Bruël, J. M. Jansen: Implementing a non-strict purely functional language in JavaScript, Acta Universitatis Sapientiae, 2011
 
-\bibitem[Duetto compiler (2014)]{duetto}
-THE INTERNETS: http://leaningtech.com/duetto/
+\bibitem[Done (2012)]{fay}
+C. Done: Fay, JavaScript, etc., http://chrisdone.com/posts/fay, 2012
 
@@
 C. M. Elliott: Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell, 2009
-
-\bibitem[The Fay language (2014)]{fay}
-THE INTERNETS: http://fay-lang.org
 
@@
 
+\bibitem[Pignotti (2013)]{duetto}
+A. Pignotti: Duetto: a C++ compiler for the Web going beyond emscripten and node.js, http://leaningtech.com/duetto/blog/2013/10/31/Duetto-Released/, 2013
+
 \bibitem[Taylor (2013)]{printf}

### 2014-02-27 paper.tex
Add half of the Haste.App chatbox example to paper.tex.

@@
 to revisit the chatbox program, this time using our improved programming model.
+Since we are now writing the entire application, both client and server, as
+opposed to the client part from our motivating example, our program has three
+new responsibilities.
+
+\begin{itemize}
+  \item We need to add connecting users to a list of message recipients;
+  \item users leaving the site need to be removed from the recipient list; and
+  \item chat messages need to be distributed to all users in the list.
+\end{itemize}
+
+With this in mind, we begin by importing a few modules we are going to need and
+define the type for our recipient list.
+
+\begin{code}
+import Data.List (find)
+import Haste.App
+import Haste.App.Concurrent
+import qualified Control.Concurrent as CC
+
+type Recipient = (SessionID, CC.MVar String)
+type RcptList = CC.MVar [Recipient]
+\end{code}
+
+We use an \lstinline!MVar! from \lstinline!Control.Concurrent! to store the
+list of recipients. A recipient will be represented by a \lstinline!SessionID!,
+an ID used by Haste.App to identify user sessions, and an \lstinline!MVar! into
+which new chat messages sent to the recipient will be written. Next, we define
+our handshake RPC function.
+
+\begin{code}
+srvHello :: Useless RcptList -> Server ()
+srvHello uRcpts = do
+  recipients <- mkUseful uRcpts
+  sid <- getSessionID
+  rcptMVar <- liftIO CC.newEmptyMVar
+  liftIO $ CC.modifyMVar recipients $ \cs -> do
+    return ((sid, rcptMVar):cs ,())
+\end{code}
+
+The handshake is a simple operation; the connecting client just gets an
+\lstinline!MVar! associated with its session identifier, and the two are then
+prepended to the recipient list. Notice how the application's server state is
+passed in as the function's first \lstinline!Useless RcptList! argument.
+
+\begin{code}
+srvSend :: Useless RcptList -> String -> Server ()
+srvSend uRcpts message = do
+    rcpts <- mkUseful uRcpts
+    liftIO $ do
+      recipients <- CC.withMVar rcpts return
+      mapM_ (CC.forkIO . deliver message) recipients
+  where
+    deliver :: String -> Recipient -> IO ()
+    deliver message (_, rcptMVar) =
+      CC.putMVar rcptMVar message
+\end{code}
+
+The send function is slightly more complex. The incoming message is written to
+the \lstinline!MVar! corresponding to each active session. Note the use of
+\lstinline!forkIO!. Some recipients may be slow to empty their \lstinline!MVar!
+to make room for a new message. We don't want to force all recipients to wait
+for the slowest one, so we spawn a new lightweight thread for each recipient.
+
+\begin{code}
+srvAwait :: Useless RcptList -> Server String
+srvAwait uRcpts = do
+  rcpts <- mkUseful uRcpts
+  sid <- getSessionID
+  liftIO $ do
+    recipients <- CC.withMVar rcpts return
+    case find ((== sid) . fst) recipients of
+      Just (_, mv) -> CC.takeMVar mv
+      _            -> fail "Unregistered session; aborting"
+\end{code}
+
+The final server operation, notifying users of pending messages, finds the
+appropriate \lstinline!MVar! to wait on by searching the recipient list for the
+session identifier of the calling user, and then blocks until a message arrives
+in said \lstinline!MVar!. This is a little different from the other two
+operations, which perform their work as quickly as possible and then return
+immediately.
 

### 2014-02-27 haste.hs
Make haste.hs prettier.

@@ -3,26 +3,45 @@ import Haste.App
 import Haste.App.Concurrent
 import qualified Control.Concurrent as CC
 
-main = runApp (defaultConfig "ws://localhost:24601" 24601) $ do
-  clients <- liftServerIO $ CC.newMVar []
+type Recipient = (SessionID, CC.MVar String)
+type RcptList = CC.MVar [Recipient]
 
-  hello <- export $ mkUseful clients >>= \clients -> do
-    sid <- getSessionID
-    liftIO $ CC.modifyMVar clients $ \cs -> do
-      mv <- CC.newEmptyMVar
-      return ((sid, mv):cs ,())
+srvHello :: Useless RcptList -> Server ()
+srvHello uRcpts = do
+  recipients <- mkUseful uRcpts
+  sid <- getSessionID
+  rcptMVar <- liftIO CC.newEmptyMVar
+  liftIO $ CC.modifyMVar recipients $ \cs -> do
+    return ((sid, rcptMVar):cs ,())
 
-  awaitMsg <- export $ mkUseful clients >>= \clients -> do
-    sid <- getSessionID
+srvSend :: Useless RcptList -> String -> Server ()
+srvSend uRcpts message = do
+    rcpts <- mkUseful uRcpts
     liftIO $ do
-      cs <- CC.withMVar clients return
-      case find ((== sid) . fst) cs of
-        Just (_, mv) -> CC.takeMVar mv
-        _            -> fail "Client did not say hello - abort session!"
+      recipients <- CC.withMVar rcpts return
+      mapM_ (CC.forkIO . deliver message) recipients
+  where
+    deliver :: String -> Recipient -> IO ()
+    deliver message (_, rcptMVar) =
+      CC.putMVar rcptMVar message
 
-  sendMsg <- export $ \msg -> mkUseful clients >>= \clients -> liftIO $ do
-    recipients <- CC.withMVar clients return
-    mapM_ (CC.forkIO . flip CC.putMVar msg . snd) recipients
+srvAwait :: Useless RcptList -> Server String
+srvAwait uRcpts = do
+  rcpts <- mkUseful uRcpts
+  sid <- getSessionID
+  liftIO $ do
+    recipients <- CC.withMVar rcpts return
+    case find ((== sid) . fst) recipients of
+      Just (_, mv) -> CC.takeMVar mv
+      _            -> fail "Unregistered session; aborting"
+
+appMain :: App Done
+appMain = do
[…]

### 2014-02-27 paper.tex
Finish up chatbox revisited section.

@@
         the role of a computational and/or storage resource, tasked with
-        servicing client requests rather than driving the program.
+        servicing client requests rather than driving the program. While it is
+        entirely possible to implement a server-to-client communication channel
+        on top of our model, we believe that choosing one side of the
+        heterogenous client-server relation as the master helps keeping the
+        program flow linear and predictable.
   \item The implementation is built as a library on top of the GHC and Haste
@@
 completely ignore executing computations lifted using \lstinline!liftServerIO!;
-since the result of such a computation is never observable on the client,
-the client has no obligation to even produce such a value.
+since the result of a server lifted computation is never observable on the
+client, the client has no obligation to even produce such a value.
 Figure \ref{lst:good-state} shows how to make proper use of server-side state.
@@
   sid <- getSessionID
-  rcptMVar <- liftIO CC.newEmptyMVar
-  liftIO $ CC.modifyMVar recipients $ \cs -> do
+  liftIO $ do
+    rcptMVar <- CC.newEmptyMVar
+    CC.modifyMVar recipients $ \cs ->
       return ((sid, rcptMVar):cs ,())
@@
   where
-    deliver :: String -> Recipient -> IO ()
-    deliver message (_, rcptMVar) =
-      CC.putMVar rcptMVar message
+    deliver msg (_, rcptMVar) = CC.putMVar rcptMVar msg
 \end{code}
@@
       Just (_, mv) -> CC.takeMVar mv
-      _            -> fail "Unregistered session; aborting"
+      _            -> fail "Unregistered session!"
 \end{code}
@@
 immediately.
+
+If the caller's session identifier could not be found in the
+recipient list, it has for some reason not completed its handshake with the
+server. If this is the case, we simply drop the session by throwing an error;
+the session will be automatically cleaned up on the server side and an exception
+is thrown to the client.
+
+Having implemented our three server operations, all that's left is to tie them
+to the client. We start by creating an \lstinline!MVar! to hold the server
+state and exporting our three server operations.
+
+\begin{code}
+appMain :: App Done
+appMain = do
+  recipients <- liftServerIO $ CC.newMVar []
+
+  hello <- export $ srvHello recipients
+  awaitMsg <- export $ srvAwait recipients
+  sendMsg <- export $ srvSend recipients
+
+  runClient $ do
+    withElems ["log","message"] $ \[log,msgbox] -> do
+      onServer hello
+\end{code}
+
+Notice that the \lstinline!recipients! list is passed to our three server
+operations \emph{before} they are exported; since \lstinline!recipients! is
+created on the server and inaccessible to the client side, it is not possible
+to pass it over the network as an RPC argument. Even if it were possible,
+passing server-private state back and forth over the network would be highly
+undesirable.
+
+The \lstinline!withElems! function is part of Haste's DOM manipulation library;
+it locates references to the DOM nodes with the given identifiers, in this case
+the variable \lstinline!log! is bound to the node with the identifier ``log'',
+and \lstinline!inp! is bound to the node identifier by ``msg''. These are the
+same DOM nodes that were referenced in our original example, and refer
+to the chat log window and the text input field respectively. After locating
+all the needed UI elements, the client proceeds to register itself with the
+server's recipient list using the \lstinline!hello! export.
+
+\begin{code}
+      let recvLoop chatlines = do
+            setProp log "value" $ unlines chatlines
+            message <- onServer awaitMsg
+            recvLoop (message : chatlines)
+      fork $ recvLoop []
+\end{code}
+
+The \lstinline!recvLoop! function perpetually asks the server for new messages
+and updates the chat log whenever one arrives. Note that unlike the
+\lstinline!onmessage! callback of the Javascript version of this example,
+\lstinline!recvLoop! is acting as a completely self-contained process with
+linear program flow, keeping track of its own state and only reaching out to
+the outside world to write its state to the chat log whenever necessary. As
+the \lstinline!awaitMsg! function blocks until a message arrives,
+\lstinline!recvLoop! will make exactly one iteration per received message.
+
+\begin{code}
+      msgbox `onEvent` OnKeyPress $ \13 -> do
+        msg <- getProp msgbox "value"
+        setProp msgbox "value" ""
+        onServer (sendMsg <.> msg)
+\end{code}
+
+This is the final part of our program; we set up an event handler to clear the
+input box and send its contents off to the server whenever the user hits return
+(character code 13) while the input box has focus. Finally, all that is left
+is to actually launch our application.
+
+\begin{code}
+main :: IO ()
+main =
+  runApp (mkConfig "ws://localhost:1111" 1111) appMain
+\end{code}
+
+Using this approach, creating web applications with strongly typed
+communication becomes easy. The discerning reader may be slightly annoyed at
+the need to extract the contents from \lstinline!Useless! values at each point
+of use. Indeed, in a simple example such as this, the source clutter caused by
+this becomes a disproportionate irritant. Fortunately, most web applications
+tend to have more complex client-server interactions, reducing this overhead
+significantly.
 
@@
 Our model is based on a simple blocking concurrency model rather than explicit
-continuations, is well suited for use with a GUI programming style based on
+continuations which is well suited for use with a GUI programming style based on
 communicating processes with local state, and requires essentially no

### 2014-02-27 haste.hs

@@ -21,9 +21,7 @@ srvSend uRcpts message = do
       recipients <- CC.withMVar rcpts return
       mapM_ (CC.forkIO . deliver message) recipients
   where
-    deliver :: String -> Recipient -> IO ()
-    deliver message (_, rcptMVar) =
-      CC.putMVar rcptMVar message
+    deliver msg (_, rcptMVar) = CC.putMVar rcptMVar msg
 
 srvAwait :: Useless RcptList -> Server String
 srvAwait uRcpts = do
@@ -43,18 +41,21 @@ appMain = do
   awaitMsg <- export $ srvAwait recipients
   sendMsg <- export $ srvSend recipients
 
-  runClient $ withElems ["log", "message"] $ \[log, msgbox] -> do
-    onServer hello
+  runClient $ do
+    withElems ["log","message"] $ \[log,msgbox] -> do
+      onServer hello
 
-    mbox <- statefully [] $ \oldlines newline -> do
-      setProp log "value" $ unlines $ newline:oldlines
-      return . Just $ newline:oldlines
-    fork . forever $ mbox <! onServer awaitMsg
+      let rcvLoop chatlines = do
+            setProp log "value" $ unlines chatlines
+            message <- onServer awaitMsg
+            rcvLoop (message : chatlines)
+      fork $ rcvLoop []
     
-    msgbox `onEvent` OnKeyPress $ \13 -> do
-      msg <- getProp msgbox "value"
-      onServer (sendMsg <.> msg)
-      setProp msgbox "value" ""
+      msgbox `onEvent` OnKeyPress $ \13 -> do
+        msg <- getProp msgbox "value"
+        setProp msgbox "value" ""
+        onServer (sendMsg <.> msg)
 
 main :: IO ()
-main = runApp (mkConfig "ws://localhost:1111" 1111) appMain
+main =
+  runApp (mkConfig "ws://localhost:1111" 1111) appMain

### 2014-02-28 paper.tex
Don't Overuse Capitals In The Title
Update keywords.
Add explanation as to why there's a bunch of Haste stuff here.

@@
 
-\title{Type-Safe Client-Server Communication in Web Applications}
+\title{Type-safe client-server communication for web applications}
 \subtitle{Subtitle Text, if any}
@@
 \keywords
-web
+web applications; distributed systems; network communication
 
@@
 to no compiler support, ensuring portability and modularity.
+
+This paper also describes the Haste Haskell compiler in some detail, as it is a
+crucial part of our implementation.
 

### 2014-02-28 paper.tex
Add related work for defunctionalization and cloud haskell.

@@
 
+\paragraph{Inspiration and alternatives to \lstinline!export!} One crucial
+aspect of implementing cross-network function calls is, of course, the issue of
+data representation: the client side of things must be able to obtain some
+representation of any function it may want to call. In our solution, this
+representation is obtained through the use of the \lstinline!export! function,
+which when executed on the server pairs a function with a unique identifier and
+on the client, and when executed on the client returns said identifier so that
+the client may now refer to the function. While this has the advantage of being
+simple to implement, one major drawback of this method is that all functions
+must be explicitly exported in the \lstinline!App! monad prior to being called
+over the network.
+
+This approach was inspired by Cloud Haskell \cite{cloudhaskell}, which
+introduces the notion of ``static values''; values which are known at compile
+time. Codifying this in the type system, to enable it to be used as a basis for
+remote procedure calls, unfortunately requires some major changes to the
+compiler. Cloud Haskell has a stopgap measure for unmodified compilers wherein
+a remote table, pairing values with unique identifiers, are kept. This explicit
+bookkeeping relies on the programmer to assign appropriate types to both values
+themselves and their identifiers, breaking type safety.
+
+The astute reader may notice that this is exactly what the \lstinline!export!
+does as well, the difference being that \export!export! links the identifier to
+the value it represents on the type level, making it impossible to call
+non-existent remote functions and break the program's type safety in other ways.
+
+Another approach to this problem is defunctionalization
+\cite{defunctionalization}, a program transformation wherein functions are
+translated into algebraic data types. This approach would allow the client and
+server to use the same actual code; rather than passing an identifier around,
+the client would instead pass the actual defunctionalized function. This would
+have the added benefit of allowing functions to be arbitrarily composed before
+being remotely invoked.
+
+This approach also requires significant changes to the compiler, making it
+unsuitable for our use case. Moreover, we are not entirely convinced about the
+wisdom of allowing server-side execution of what is essentially arbitrary code
+sent from the client which, for web applications, is completely untrustworthy.
+
 \subsection{Limitations}
@@
 
+\bibitem[Danvy, Nielsen (2001)]{defunctionalization}
+O. Danvy, L. R. Nielsen: Defunctionalization at work, 2001
+
 \bibitem[Dijkstra et al (2013)]{uhc}
@@
 
+\bibitem[Epstein et al (2011)]{cloudhaskell}
+J. Epstein, A. P. Black, S. Peyton Jones: Towards Haskell in the Cloud, 2011
+
 \bibitem[Guthrie (2014)]{jserlang}
@@
 
+\bibitem[Svenningsson, Svensson (2013)]{bjornbenny}
+J. Svenningsson, J. Svensson: Simple and Compositional Reification of Monadic Embedded Languages, 2013
+
 \bibitem[Taylor (2013)]{printf}

### 2014-02-28 paper.tex
Add that "Haste is small and awesome"

@@
 be relatively slow due to it lacking the more mature optimization capabilities
-of GHC, as well as being in the same ballpark as GHCJS for size.\ \cite{haste}
+of GHC, as well as being in the same ballpark as GHCJS for size. Haste, in
+contrast, produces code which is generally one to two orders of magnitude
+smaller than that produced by GHCJS or UHC.\ \cite{haste}
 

### 2014-02-28 paper.tex
Fix silly typo, label for Haste section.

@@
 
-\begin{itemize}
-  \item Poor man's concurrency monad
-  \item Conditional compilation
-  \item Printf trick to turn export into [JSON] -> f
-  \item JSON over WebSockets used for transport
-  \item Server event loop awaits commands, dispatches functions, sends data back
-  \item Client sends command with nonce and blocks on MVar, event loop fills
-        MVar when return value with proper nonce arrives
-  \item Throw away unused stuff on client
-\end{itemize}
-
-\section{The Haste compiler}
+\section{The Haste compiler}\label{sec:haste}
 
@@
 obviously need some way to compile that language into Javascript. To this end,
-we make use of the Haste compiler\ \cite{haste}, developed as a MSc thesis and
-extended as part of this work. Haste builds on the GHC compiler to provide
+we make use of the Haste compiler\ \cite{haste}, started as an MSc thesis and
+continued as part of this work. Haste builds on the GHC compiler to provide
 the full Haskell language, including most GHC-specific extensions, in the
@@
 The astute reader may notice that this is exactly what the \lstinline!export!
-does as well, the difference being that \export!export! links the identifier to
-the value it represents on the type level, making it impossible to call
+does as well, the difference being that \lstinline!export! links the identifier
+to the value it represents on the type level, making it impossible to call
 non-existent remote functions and break the program's type safety in other ways.
@@
 
+\bibitem[Claessen (1999)]{concurrencymonad}
+K. Claessen: Functional Pearls: A poor man's concurrency monad, 1999
+
 \bibitem[Conductance application server (2014)]{conductance}

### 2014-02-28 paper.tex
Explain App monad.

@@
 \section{Implementation}
+
+Our implementation is built in three layers: the compiler layer, the
+concurrency layer and the communication layer. The concurrency and
+communication layers are simple Haskell libraries, portable to any other pair
+of standard Haskell compilers with minimal effort.
+
+\paragraph{Two compilers}
+The principal trick to our solution is compiling the same program twice; once
+with a compiler that generates the server binary, and once with one that
+generates Javascript for execution on the client. Conditional compilation is
+used for a select few functions, to enable slightly different behavior on the
+client and on the server as necessary. As we believe that a monadic
+programming style is a good fit for this application, we have chosen to use
+Haskell as the base of our solution, which by default leads us to choose GHC
+as our server-side compiler. For reasons outlined in section \ref{sec:haste},
+the Haste compiler was chosen to provide the client-side code.
+
+\paragraph{The \lstinline!App! monad}
+The \lstinline!App! monad is where remote functions are declared, server state
+is initialized and program flow is finally handed over to the \lstinline!Client!
+monad. Slightly simplified, its definition is as follows.
+
+\begin{code}
+type Method = [JSON] -> IO JSON
+type AppState = (CallID, [(CallID, Method)])
+newtype App a = App (StateT AppState IO a)
+\end{code}
+
+As we can see, \lstinline!App! is a simple state monad, with underlying IO
+capabilities to allow server-side computations to be forked from within it.
+Aside from the \lstinline!runClient! function, which is only used to launch the
+client program, This monad has two primary operations.
+
+\begin{code}
+liftServerIO :: IO a -> App (Useless a)
+export :: Exportable a => a -> App (Export a)
+\end{code}
+
+Aside from the \lstinline!App! monad itself, these functions operate on two
+types we have not seen before.
+
+\begin{code}
+data Useless a = Useful a | Useless
+data Export a = Export CallID [JSON]
+\end{code}
+
+The definition of \lstinline!liftServerIO! is very simple. Through the use of
+conditional compilation, on the client it simply ignores its argument and
+returns \lstinline!Useless!; recall that server-lifted IO computations are not
+observable on the client. On the server side, \lstinline!liftServerIO! executes
+its argument and wraps the computation's return value using the
+\lstinline!Useful! constructor. The counterpart function, \lstinline!mkUseful!
+simply unwraps the \lstinline!Useful! constructor and returns the contained
+value in the \lstinline!Server! monad.
+
+\paragraph{The \lstinline!Client! monad and the \lstinline!onServer! function}
+As synchronous network communication is one of our stated goals, it is clear
+that we will need some kind of blocking primitive. Since Javascript does not
+support any kind of blocking, we will have to implement this ourselves.
+
+A solution is given in the \emph{poor man's concurrency
+monad}\ \cite{concurrencymonad}. Making use of a continuation monad with
+primitive operations for forking a computation and atomically lifting an IO
+computation into the monad, it is possible to implement cooperative
+multitasking on top of the non-concurrent Javascript runtime. This monad
+allows us to implement \lstinline!MVar!s as our blocking primitive, with the
+same semantics as their regular Haskell counterpart.\ \cite{ffi}
+This concurrency-enhanced monad is used as the basis of the \lstinline!Client!
+monad.
 
@@
 function can not be determined statically, application is implemented using the
-eval/apply method described in \cite{fastcurry} instead.
+eval/apply method described in\ \cite{fastcurry} instead.
 
@@
 
-This approach was inspired by Cloud Haskell \cite{cloudhaskell}, which
+This approach was inspired by Cloud Haskell\ \cite{cloudhaskell}, which
 introduces the notion of ``static values''; values which are known at compile
@@
 
-Another approach to this problem is defunctionalization
-\cite{defunctionalization}, a program transformation wherein functions are
-translated into algebraic data types. This approach would allow the client and
-server to use the same actual code; rather than passing an identifier around,
-the client would instead pass the actual defunctionalized function. This would
-have the added benefit of allowing functions to be arbitrarily composed before
-being remotely invoked.
+Another approach to this problem is
+defunctionalization\ \cite{defunctionalization}, a program transformation
+wherein functions are translated into algebraic data types. This approach would
+allow the client and server to use the same actual code; rather than passing
+an identifier around, the client would instead pass the actual defunctionalized
+function. This would have the added benefit of allowing functions to be
+arbitrarily composed before being remotely invoked.
 
@@
 sent from the client which, for web applications, is completely untrustworthy.
+While analyzing code for improper behavior is certainly possible, implementing
+and enforcing a sufficiently strict-yet-flexible security policy for remote
+execution would be an unwelcome burden on the programmer.
 
@@
 
-\acks
-
-Acknowledgments, if needed.
+%\acks
+%
+%Acknowledgments, if needed.
 

### 2014-02-28 paper.tex
Finish explanation of App monad and its operations.

@@
 
-\title{Type-safe client-server communication for web applications}
-\subtitle{Subtitle Text, if any}
+\title{Type-safe client-server communication for functional web applications}
 
@@
 its arguments as well as its return value are serializable through the aptly
-named \lstinline!Serializable! type class, and returns a typed ``handle'' of
+named \lstinline!Serialize! type class, and returns a typed ``handle'' of
 sorts. In this example, the type of \lstinline!greetings! is
@@
 
+To pass data back and forth over the network, messages are serialized using
+JSON, a fairly lightweight format used by many web applications, and sent using
+the HTML5 WebSockets API. This choice is completely arbitrary, guided purely
+by implementation convenience. It is certainly not the most performant choice,
+but can be easily replaced with something more suitable as the need arises.
+
 \paragraph{Two compilers}
@@
 \begin{code}
+type CallID = Int
 type Method = [JSON] -> IO JSON
 type AppState = (CallID, [(CallID, Method)])
-newtype App a = App (StateT AppState IO a)
+type App = StateT AppState IO
 \end{code}
@@
 capabilities to allow server-side computations to be forked from within it.
+Its \lstinline!CallID! state element contains the identifier to be given to the
+next exported function, and its other state element contains a mapping from
+identifiers to exported functions. Of course, in reality, \lstinline!App! is a
+\lstinline!newtype!, to prevent arbitrary lifting of IO computations.
+
 Aside from the \lstinline!runClient! function, which is only used to launch the
@@
 \begin{code}
+data Useless a = Useful a | Useless
+data Export a = Export CallID [JSON]
+
 liftServerIO :: IO a -> App (Useless a)
 export :: Exportable a => a -> App (Export a)
-\end{code}
-
-Aside from the \lstinline!App! monad itself, these functions operate on two
-types we have not seen before.
-
-\begin{code}
-data Useless a = Useful a | Useless
-data Export a = Export CallID [JSON]
 \end{code}
@@
 value in the \lstinline!Server! monad.
+
+\lstinline!export! is a more interesting function. Looking at its type, we see
+that it accepts any \lstinline!Exportable! value for its argument.
+\lstinline!Exportable! is defined as follows.
+
+\begin{code}
+class Exportable a where
+  mkExported :: a -> ([JSON] -> IO JSON)
+
+instance Serialize a =>
+         Exportable (Server a)
+instance (Serialize a, Exportable b) =>
+         Exportable (a -> b)
+\end{code}
+
+In essence, any function, over any number of arguments, which returns a
+serializable value in the \lstinline!Server! monad can be exported. The
+\lstinline!mkExported! function makes use a well-known type class trick for
+creating statically typed variadic functions, and works very much like the
+\lstinline!printf! function of Haskell's standard library.\ \cite{printf}
+From an \emph{n}-ary function, it creates a unary function which expects
+a list of \emph{n} JSON-encoded elements as its argument. When applied, its
+argument list is deserialized and passed to the exported function, and its
+return value is serialized. This allows us to easily bundle up all of an
+exported function's arguments and pass them across the network.
+
+The client implementation of \lstinline!export! is straightforward.
+
+\begin{code}
+export _ = do
+  (next_id, exports) <- get
+  put (next_id+1, exports)
+  return (Export next_id [])
+\end{code}
+
+As the exported function lives on the server, the client only needs an
+identifier to be able to call on it. The exported function is thus ignored,
+so that it can be optimized out of existence in the client version of the code.
+The server version is only marginally more complex in that it updates the
+identifier-function mapping table as well.
+
+\begin{code}
+export f = do
+  (next_id, exports) <- get
+  put (next_id+1, (next_id, mkExported f) : exports)
+  return (Export next_id [])
+\end{code}
+
+The second argument of the \lstinline!Export! constructor is, as the observant
+reader might have guessed, the list of arguments to be sent to the export.
+Essentially, an explicit closure is built up using the \lstinline!<.>! operator
+and sent off over the network for execution.
+
+\begin{code}
+(<.>) :: Serialize a
+      => Export (a -> b) -> a -> Export b
+(Export identifier args) <.> arg =
+  Export identifier (serialize arg : args)
+\end{code}
 
@@
 representation, to dynamically create a function object at runtime.
-The \lstinline!ffi! function is typed using a well-known type class trick for
-creating statically typed variadic functions\ \cite{printf}, and works very
-much like the \lstinline!printf! function of Haskell's standard library.
+The \lstinline!ffi! function is typed using the same method as the
+\lstinline!serializify! function described in section \ref{sec:impl}.
 When applied to one or more arguments instantiating the \lstinline!Marshal!

### 2014-02-28 paper.tex
Add abstract.

@@
 \begin{abstract}
-This is the text of the abstract.
+Developing web applications, especially applications which need to communicate
+with a server, is often a painful experience. What is conceptually a single
+monolithic application is forcefully split into a client and a server
+explicitly communicating over the network, with all the code duplication,
+complexity and lost type safety this entails.
+
+We propose instead a development model in which a web application is written
+and compiled as a whole, allowing for synchronous, type safe client-server
+communication without the need for explicit networking code, as well as the
+sharing of code between client and server.
 \end{abstract}

### 2014-02-28 paper.tex
Finish explaining Client monad.

@@
 sorts. In this example, the type of \lstinline!greetings! is
+\linebreak
 \lstinline!Export (String -> Server ())!, indicating that the handle identifies
@@
   where
-    deliver msg (_, rcptMVar) = CC.putMVar rcptMVar msg
+    deliver msg (_,rcptMVar) = CC.putMVar rcptMVar msg
 \end{code}
@@
 
-Aside from the \lstinline!runClient! function, which is only used to launch the
-client program, This monad has two primary operations.
+What makes \lstinline!App! interesting is that computations in this monad are
+executed on both the client and the server; once on server startup, and once
+in the startup phase of each client. Its operations behave slightly differently
+depending on whether they are executed on the client or on the server.
+Execution is deterministic, ensuring that the same sequence of
+\lstinline!CallID!s are generated during every run, both on the server and on
+all clients. This is necessary to ensure that any particular call identifier
+always refers to the same server-side function on all clients.
+
+After all common code has been executed, the program flow diverges between the
+client and the server; client-side, \lstinline!runClient! launches the
+application's \lstinline!Client! computation whereas on the server, this
+computation is discarded, and the server instead goes into an event loop,
+waiting for calls from the client.
+
+Aside from \lstinline!runClient!, this monad has two primary operations.
 
@@
 
+After the \lstinline!App! computation finishes, the identifier-function mapping
+accumulated in its state is handed over to the server's event loop, where it is
+used to dispatch the proper functions for incoming calls from the client.
+
 \paragraph{The \lstinline!Client! monad and the \lstinline!onServer! function}
@@
 monad.
+
+\begin{code}
+type Nonce = Int
+type Client = StateT Nonce Conc
+\end{code}
+
+Aside from the added concurrency capabilities, the \lstinline!Client! monad
+only has a single particularly interesting operation: \lstinline!onServer!.
+
+\begin{code}
+onServer :: Serialize a
+         => Export (Server a) -> Client a
+onServer (Export identifier args) = do
+  nonce <- get
+  put (nonce + 1)
+  mv <- createResultMVar nonce
+  webSocketSend $ serialize (nonce, identifier, reverse args)  
+  takeMVar mv
+\end{code}
+
+A nonce is used to pair up return values with the calls that produced them.
+After a call is dispatched, \lstinline!onServer! blocks, waiting for its
+\emph{result variable} to be filled with the result of the call. Filling this
+variable is the responsibility of the \emph{receive callback}, which is
+executed every time a message arrives from the server.
+
+\begin{code}
+onMessage :: JSON -> Client ()
+onMessage msg = do
+  let (nonce, result) = deserialize msg
+  mv <- findResultMVar nonce
+  putMVar mv result
+\end{code}
+
+The corresponding server-side event loop, which handles incoming calls from the
+client, is similarly straightforward.
+
+\begin{code}
+onEvent :: [(CallID, Method)] -> JSON -> IO ()
+onEvent mapping msg = do
+  let (nonce, identifier, args) = deserialize msg
+      f = find identifier mapping
+  result <- f args
+  webSocketSend $ serialize (nonce, result)
+\end{code}
+
+The function corresponding to the RPC call's identifier is looked up in the
+identifier-function mapping generated by the \lstinline!App! computation and
+applied to the list of arguments; the return value is bundled up with the
+call's nonce and sent back to the client.
 
@@
 solution combines type safe communication between the client and the server
-with purely functional semantics, clear demarcations as to when data is
-transmitted and where a particular piece of code is executed, and the ability to
+with functional semantics, clear demarcations as to when data is transmitted
+and where a particular piece of code is executed, and the ability to
 effortlessly share code between the client and the server.
@@
 continuations which is well suited for use with a GUI programming style based on
-communicating processes with local state, and requires essentially no
+self-contained processes with local state, and requires essentially no
 modification of existing tools or compilers.
@@
 
-\bibitem[Svenningsson, Svensson (2013)]{bjornbenny}
-J. Svenningsson, J. Svensson: Simple and Compositional Reification of Monadic Embedded Languages, 2013
-
 \bibitem[Taylor (2013)]{printf}

### 2014-02-28 paper.tex
Remove GUI toolkit from future work. I just can't get it to sound good.

@@
 
-\paragraph{Reactive client-server GUI toolkit} While we believe that our work
-represents a step forward in making web applications more robust and easier to
-develop, the basic model for updating the application's user-visible state is
-still essentially imperative. Implementing an FRP-style\ \cite{ppfrp} GUI
-programming model on top of our client-server abstraction may prove a fruitful
-way to further improve the programmability of web-based distributed
-applications.
-
 \section{Conclusion}
@@
 
-\bibitem[Elliott (2009)]{ppfrp}
-C. M. Elliott: Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell, 2009
-
 \bibitem[Epstein et al (2011)]{cloudhaskell}

### 2014-02-28 paper.tex
Fix Peter's feedback.

@@
 To put make the problem a bit more concrete, let's take a look at an example.
-Let's say what we want to implement a simple ``chatbox'' component for a
+Let's say that we want to implement a simple ``chatbox'' component for a
 website, to allow visitors to discuss the site's content in real time. Using
@@
         client or the server using the type system while pure code can be
-        shared by both. Additionally, both client and server code may lift
-        general IO computations, allowing for safe IO code reuse within the
-        confines of the client or server designated code sections.
+        shared by both. Additionally, general IO computations may be lifted
+        into both client and server code, allowing for safe IO code reuse
+        within the confines of the client or server designated code sections.
   \item Client-server network communication is handled through statically typed
@@
 
-This leaves us choosing between Clean and Haste for our client compiler. Since
-most of our abstractions, both the cooperative multitasking and the type-safe
+This leaves us choosing between Clean and Haste for our client compiler. Most
+of our abstractions, both the cooperative multitasking and the type-safe
 client-server communication itself, are easily expressed as monads, well
-supported in Haskell thanks to do-notation, the choice fell to Haste.
+supported in Haskell thanks to do-notation. Add the fact that one of this
+work's authors is the original author of Haste, giving us a better position to
+troubleshoot any code based thereupon, and Haste becomes the better choice.
 
@@
 Functions are represented as plain Javascript functions, one of the blessings
-of targeting a high level language, and application is can therefore be
+of targeting a high level language, and application can therefore be
 implemented as its Javascript counterpart in most cases. In the general case,

### 2014-03-01 paper.tex
Fix some metadata.
Add note about simplification and error handling to implementation section.

@@
 
-\conferenceinfo{ICFP '14}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
+\conferenceinfo{ICFP~'14}{September 1--3, 2014, Gothenburg, Sweden} 
+\copyrightyear{2014}
+\copyrightdata{978-1-nnnn-nnnn-n/14/09}
 \doi{nnnnnnn.nnnnnnn}
@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
+\titlebanner{-- ICFP '14 draft submission --}        % These are ignored unless
+\preprintfooter{-- -- -- -- draft -- -- -- --}   % 'preprint' option specified.
 
@@
 \begin{code}
-import Data.List (find)
 import Haste.App
@@
     recipients <- CC.withMVar rcpts return
-    case find ((== sid) . fst) recipients of
-      Just (_, mv) -> CC.takeMVar mv
+    case lookup sid recipients of
+      Just mv -> CC.takeMVar mv
       _       -> fail "Unregistered session!"
@@
 
-\section{Implementation}
+\section{Implementation}\label{sec:impl}
 
@@
 but can be easily replaced with something more suitable as the need arises.
+
+The implementation described here is a slight simplification of our
+implementation, removing some performance enhancements and error handling
+clutter in the interest of clarity.
 
@@
 is initialized and program flow is finally handed over to the \lstinline!Client!
-monad. Slightly simplified, its definition is as follows.
+monad. Its definition is as follows.
 
@@
   let (nonce, identifier, args) = deserialize msg
-      f = find identifier mapping
+      Just f = lookup identifier mapping
   result <- f args
@@
 applied to the list of arguments; the return value is bundled up with the
-call's nonce and sent back to the client.
+call's nonce and sent back to the client. Note that during normal operation,
+it is not possible for the client to submit an RPC call with a non-existent
+call identifier, hence the irrefutable pattern match on \lstinline!Just f!.
+Should this pattern match fail, this is a sure sign of malicious tampering;
+the resulting exception is caught and the session is dropped as it is no longer
+meaningful to continue.
 

### 2014-03-01 haste.hs

@@ -1,4 +1,3 @@
-import Data.List
 import Haste.App
 import Haste.App.Concurrent
 import qualified Control.Concurrent as CC
@@ -29,9 +28,9 @@ srvAwait uRcpts = do
   sid <- getSessionID
   liftIO $ do
     recipients <- CC.withMVar rcpts return
-    case find ((== sid) . fst) recipients of
-      Just (_, mv) -> CC.takeMVar mv
-      _            -> fail "Unregistered session; aborting"
+    case lookup sid recipients of
+      Just mv -> CC.takeMVar mv
+      _       -> fail "Unregistered session; aborting"
 
 appMain :: App Done
 appMain = do

### 2014-03-01 paper.tex
Add categories.
Fig overflowing code for onServer.
Fix up references.

@@
 
-\category{CR-number}{subcategory}{third-level}
+\category{D.1.3}{Programming Techniques}{Distributed Programming}
+\category{D.3.2}{Language Classifications}{Applicative (functional) languages}
+\category{H.3.5}{Online Information Services}{Web-based services}
 
@@
   mv <- createResultMVar nonce
-  webSocketSend $ serialize (nonce, identifier, reverse args)  
+  webSocketSend $
+    serialize (nonce, identifier, reverse args)  
   takeMVar mv
@@
 \bibitem[Bracker, Gill (2014)]{sunroof}
-J. Bracker and A. Gill: Sunroof: A monadic DSL for generating JavaScript, Practical Aspects of Declarative Languages, 65-80, 2014
+J. Bracker and A. Gill. "Sunroof: A Monadic DSL for Generating JavaScript." In Practical Aspects of Declarative Languages, pp. 65-80. Springer International Publishing, 2014.
 
 \bibitem[Claessen (1999)]{concurrencymonad}
-K. Claessen: Functional Pearls: A poor man's concurrency monad, 1999
+K. Claessen. "Functional Pearls: A poor man's concurrency monad." Journal of Functional Programming 9 (1999): 313-324.
 
 \bibitem[Conductance application server (2014)]{conductance}
-The Conductance application server, http://conductance.io
+The Conductance application server. Retrieved March 1, 2014, from http://conductance.io.
 
 \bibitem[Danvy, Nielsen (2001)]{defunctionalization}
-O. Danvy, L. R. Nielsen: Defunctionalization at work, 2001
+O. Danvy and L. R. Nielsen. "Defunctionalization at work." In Proceedings of the 3rd ACM SIGPLAN international conference on Principles and practice of declarative programming, pp. 162-174. ACM, 2001.
 
 \bibitem[Dijkstra et al (2013)]{uhc}
-A. Dijkstra, J. Stutterheim, A. Vermeulen, S. D. Swierstra: Building JavaScript applications with Haskell, Implementation and Application of Functional Languages, pp 37-52, 2013
+A.
+Dijkstra, J.
+Stutterheim, A.
+Vermeulen, and S.
+D.
+Swierstra.
+"Building JavaScript applications with Haskell." In Implementation and Application of Functional Languages, pp.
+37-52.
+Springer Berlin Heidelberg, 2013.
 
 \bibitem[Domoszlai et al (2011)]{jsclean}
-L. Domoszlai, E. Bruël, J. M. Jansen: Implementing a non-strict purely functional language in JavaScript, Acta Universitatis Sapientiae, 2011
+L. Domoszlai, E. Bruël, and J. M. Jansen. "Implementing a non-strict purely functional language in JavaScript." Acta Universitatis Sapientiae 3 (2011): 76-98.
 
 \bibitem[Done (2012)]{fay}
-C. Done: Fay, JavaScript, etc., http://chrisdone.com/posts/fay, 2012
+C. Done. (2012, September 15). ``Fay, JavaScript, etc.'', Retrieved March 1, 2014, from http://chrisdone.com/posts/fay.
 
 \bibitem[Ekblad (2012)]{haste}
-A. Ekblad: Towards a declarative web, MSc thesis, University of Gothenburg, 2012
+A. Ekblad. "Towards a declarative web." Master of Science Thesis, University of Gothenburg (2012).
 
 \bibitem[Epstein et al (2011)]{cloudhaskell}
-J. Epstein, A. P. Black, S. Peyton Jones: Towards Haskell in the Cloud, 2011
+J. Epstein, A. P. Black, and S. Peyton-Jones. "Towards Haskell in the cloud." In ACM SIGPLAN Notices, vol. 46, no. 12, pp. 118-129. ACM, 2011.
 
 \bibitem[Guthrie (2014)]{jserlang}
-G. Guthrie: http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/, 2014
+G. Guthrie. (2014, January 1). "Your transpiler to Javascript toolbox". Retrieved March 1, 2014, from http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/.
 
 \bibitem[Hedin et al (2014)]{jsflow}
-D. Hedin, A. Birgisson, L. Bello and A. Sabelfeld: JSFlow: Tracking Information Flow in JavaScript and its APIs, 2014
+D. Hedin, A. Birgisson, L. Bello, and A. Sabelfeld. "JSFlow: Tracking information flow in JavaScript and its APIs." In Proc. 29th ACM Symposium on Applied Computing. 2014.
 
 \bibitem[Marlow, Peyton Jones (2004)]{fastcurry}
-S. Marlow, S. Peyton Jones: Making a fast curry: push/enter vs. eval/apply for higher order languages, 2004
+S. Marlow, and S. Peyton Jones. "Making a fast curry: push/enter vs. eval/apply for higher-order languages." In ACM SIGPLAN Notices, vol. 39, no. 9, pp. 4-15. ACM, 2004.
 
 \bibitem[Nazarov (2012)]{ghcjs}
-V. Nazarov: GHCJS Haskell to Javascript Compiler, https://github.com/ghcjs/ghcjs
+V. Nazarov. "GHCJS Haskell to Javascript Compiler". Retrieved March 1, 2014, from https://github.com/ghcjs/ghcjs.
 
 \bibitem[Peyton Jones (1992)]{stg}
-S. Peyton Jones: Implementing lazy functional languages on stock hardware: the Spineless Tagless G-machine, Journal of Functional Programming, 1992
+S. Peyton Jones. "Implementing lazy functional languages on stock hardware: the Spineless Tagless G-machine." J. Funct. Program. 2, no. 2 (1992): 127-202.
 
 \bibitem[Peyton Jones (2001)]{ffi}
-S. Peyton Jones: Tackling the awkward squad: monadic input/output, concurrency, exceptions, and foreign-language calls in Haskell
+S. Peyton Jones. "Tackling the awkward squad: monadic input/output, concurrency, exceptions, and foreign-language calls in Haskell." Engineering theories of software construction 180 (2001): 47-96.
 
 \bibitem[Pignotti (2013)]{duetto}
-A. Pignotti: Duetto: a C++ compiler for the Web going beyond emscripten and node.js, http://leaningtech.com/duetto/blog/2013/10/31/Duetto-Released/, 2013
+A. Pignotti. (2013, October 31). "Duetto: a C++ compiler for the Web going beyond emscripten and node.js". Retrieved March 1, 2014, from http://leaningtech.com/duetto/blog/2013/10/31/Duetto-Released/.
 
 \bibitem[Taylor (2013)]{printf}
-C. Taylor: Polyvariadic Functions and Printf, http://chris-taylor.github.io/blog/2013/03/01/how-haskell-printf-works/, 2013
+C. Taylor. (2013, March 1). "Polyvariadic Functions and Printf". Retrieved March 1, 2014, from http://chris-taylor.github.io/blog/2013/03/01/how-haskell-printf-works/.
 

### 2014-03-01 paper.tex
IN functional web applications, not for.
Various modifications to things that looked bad. Is decent now.

@@
 
-\title{Type-safe client-server communication for functional web applications}
+\title{Type-safe client-server communication in functional web applications}
 
@@
 writing the one in Javascript, which is notorious even among its proponents for
-being wonky and error-prone, and the other in the current language du jour.
+being wonky and error-prone, and the other in the current server language
+\emph{du jour}.
 Then, two are glued together using whichever home-grown network protocol seems
@@
 
-This paper also describes the Haste Haskell compiler in some detail, as it is a
-crucial part of our implementation.
+As our implementation depends on the Haste Haskell to Javascript compiler, this
+paper also goes into some detail about its design and implementation.
 
@@
 
-To put make the problem a bit more concrete, let's take a look at an example.
-Let's say that we want to implement a simple ``chatbox'' component for a
-website, to allow visitors to discuss the site's content in real time. Using
-mainstream development practives and recent technologies such as WebSockets,
-we may come up with something like the program in figure
-\ref{lst:javascript-client}.
+For a more concrete example, let's say that we want to implement a simple
+``chatbox'' component for a website, to allow visitors to discuss the site's
+content in real time. Using mainstream development practices and recent
+technologies such as WebSockets\ \cite{websockets}, we may come up with
+something like the program in figure \ref{lst:javascript-client}.
 
@@
 indicating the type of the message and the rest being the optional payload.
-There are only two messages; a handshake indicating that a user wants to join
+There are only two messages: a handshake indicating that a user wants to join
 the conversation, and a broadcast message which sends a line of text to all
@@
         into both client and server code, allowing for safe IO code reuse
-        within the confines of the client or server designated code sections.
+        within the confines of the client or server designated functions.
   \item Client-server network communication is handled through statically typed
@@
   \item The implementation is built as a library on top of the GHC and Haste
-        Haskell compilers, requiring little to no compiler support. Programs
-        are compiled twice; once with Haste and once with GHC, to produce the
-        final client and server side code respectively.
+        Haskell compilers, requiring little to no specialized compiler support.
+        Programs are compiled twice; once with Haste and once with GHC, to
+        produce the final client and server side code respectively.
 \end{itemize}
@@
 its arguments as well as its return value are serializable through the aptly
-named \lstinline!Serialize! type class, and returns a typed ``handle'' of
-sorts. In this example, the type of \lstinline!greetings! is
-\linebreak
-\lstinline!Export (String -> Server ())!, indicating that the handle identifies
-an exported function with a single \lstinline!String! argument and no return
-value; exported functions all live in the \lstinline!Server! monad.
+named \lstinline!Serialize! type class, and produces a typed identifier which
+may be used to refer to the exported function. In this example, the type of
+\lstinline!greetings! is \lstinline!Export (String -> Server ())!,
+indicating that the identifier refers
+to an exported function with a single \lstinline!String! argument and no return
+value. Exported functions all live in the \lstinline!Server! monad.
+This part of the program is executed on both the server and the client, albeit
+with slightly different side effects, as described in section \ref{sec:impl}.
 
@@
 
-In order to make an RPC call using a handle obtained from \lstinline!export!,
+In order to make an RPC call using a identifier obtained from \lstinline!export!,
 we must supply it with an argument. This is done using the \lstinline!<.>!
-operator. Upon closer inspection, the attentive reader may notice that
+operator. It might be interesting to note that its type,
 \linebreak
-its type, \lstinline!Serialize a => Export (a -> b) -> a -> Export b!,
+\lstinline!Serialize a => Export (a -> b) -> a -> Export b!,
 is very similar to the type of the \lstinline!<*>! operator over applicative
-functors. This is not a coincidence; the \lstinline!<.>! performs the same role
+functors. This is not a coincidence; \lstinline!<.>! performs the same role
 for the \lstinline!Export! type as \lstinline!<*>! performs for applicative
@@
 serialized and stored inside the resulting \lstinline!Export! object, awaiting
-dispatch.
-
-After applying the value obtained by \lstinline!prompt!ing the user for her
-name to the exported function, we apply the \lstinline!onServer! function to
-the result, which dispatches the RPC call to the server. \lstinline!onServer!
-will then block until the RPC call returns.
-
-Finally, we need to launch our application, and inform the client part of the
-whereabouts of its server counterpart; in this example the server will listen
-on port 1111, and since we will only run this application locally the client
-should attempt to contact the server at our local machine.
+dispatch. \lstinline!Export! values can thus be seen as explicit
+representations of closures.
+
+After applying the value obtained from the user to the exported function,
+we apply the \lstinline!onServer! function to the result, which dispatches
+the RPC call to the server. \lstinline!onServer! will then block until the
+RPC call returns.
+
+Finally, we need to launch our application and inform the client part of the
+whereabouts of its server counterpart. In this example the server will listen
+for connecting clients on port 1111, and since we will only run this
+application locally the client should attempt to contact the server over the
+loopback interface.
 
@@
 function should for all intents and purposes be regarded as the program's entry
-point; a non-intrusive method of enforcing this using the type system is
+point. A non-intrusive method of enforcing this using the type system is
 described in section \ref{sec:limitations}, but is not used in the examples
@@
 client or the server, or perhaps on both? What happens if our state holding
-element is initialized to a non-deterministic value, and what if we assign it
+element is initialized in a non-deterministic way, and what if we assign it
 a value on the server and then attempt to read that value on the client?
@@
 is wrapped in the \lstinline!Useless! type. As the name implies, this type is
-completely opaque and thus useless unless you can somehow extract its contained
-value, something which is only possible using a single function:
+completely opaque and thus useless unless its contained value can somehow be
+extracted, something which is only possible using a single function:
 
@@
 list of recipients. A recipient will be represented by a \lstinline!SessionID!,
-an ID used by Haste.App to identify user sessions, and an \lstinline!MVar! into
-which new chat messages sent to the recipient will be written. Next, we define
-our handshake RPC function.
+an identifier used by Haste.App to identify user sessions, and an
+\lstinline!MVar! into which new chat messages sent to the recipient will be
+written as they arrive. Next, we define our handshake RPC function.
 
@@
 
-The handshake is a simple operation; the connecting client just gets an
-\lstinline!MVar! associated with its session identifier, and the two are then
-prepended to the recipient list. Notice how the application's server state is
-passed in as the function's first \lstinline!Useless RcptList! argument.
+The handshake is a simple operation; an \lstinline!MVar! is associated with the
+connecting client's session identifier, and the pair is prepended to the
+recipient list. Notice how the application's server state is passed in as the
+function's first \lstinline!Useless RcptList! argument.
 
@@
     liftIO $ do
-      recipients <- CC.withMVar rcpts return
+      recipients <- CC.readMVar rcpts
       mapM_ (CC.forkIO . deliver message) recipients
@@
 \lstinline!forkIO!. Some recipients may be slow to empty their \lstinline!MVar!
-to make room for a new message. We don't want to force all recipients to wait
-for the slowest one, so we spawn a new lightweight thread for each recipient.
+to make room for a new message. In this case, we don't want to force all
+recipients to wait for the slowest one, so we spawn a new lightweight thread
+to handle delivery for each recipient.
 
@@
   liftIO $ do
-    recipients <- CC.withMVar rcpts return
+    recipients <- CC.readMVar rcpts
     case lookup sid recipients of
@@
 operations \emph{before} they are exported; since \lstinline!recipients! is
-created on the server and inaccessible to the client side, it is not possible
-to pass it over the network as an RPC argument. Even if it were possible,
-passing server-private state back and forth over the network would be highly
-undesirable.
-
-The \lstinline!withElems! function is part of Haste's DOM manipulation library;
-it locates references to the DOM nodes with the given identifiers, in this case
-the variable \lstinline!log! is bound to the node with the identifier ``log'',
-and \lstinline!inp! is bound to the node identifier by ``msg''. These are the
-same DOM nodes that were referenced in our original example, and refer
-to the chat log window and the text input field respectively. After locating
-all the needed UI elements, the client proceeds to register itself with the
-server's recipient list using the \lstinline!hello! export.
+a mutable reference created on the server and inaccessible to client code,
+it is not possible to pass it over the network as an RPC argument.
+Even if it were possible, passing server-private state back and forth over the
+network would be quite inappropriate due to privacy and security concerns.
+
+The \lstinline!withElems! function is part of the Haste compiler's bundled DOM
+manipulation library; it locates references to the DOM nodes with the given
+identifiers and passes said references to a function.
+In this case the variable \lstinline!log! will be bound to the node
+with the identifier ``log'', and \lstinline!msgbox! will be bound to the node
+identified by ``message''. These are the same DOM nodes that were referenced in
+our original example, and refer to the chat log window and the text input
+field respectively. After locating all the needed UI elements, the client
+proceeds to register itself with the server's recipient list using the
+\lstinline!hello! export.
 
@@
 
-Using this approach, creating web applications with strongly typed
-communication becomes easy. The discerning reader may be slightly annoyed at
-the need to extract the contents from \lstinline!Useless! values at each point
-of use. Indeed, in a simple example such as this, the source clutter caused by
-this becomes a disproportionate irritant. Fortunately, most web applications
-tend to have more complex client-server interactions, reducing this overhead
-significantly.
+The discerning reader may be slightly annoyed at the need to extract the
+contents from \lstinline!Useless! values at each point of use. Indeed, in a
+simple example such as this, the source clutter caused by this becomes a
+disproportionate irritant. Fortunately, most web applications tend to have
+more complex client-server interactions, reducing this overhead significantly.
 
@@
 by implementation convenience. It is certainly not the most performant choice,
-but can be easily replaced with something more suitable as the need arises.
+but can be trivially replaced with something more suitable as needed.
 
@@
 implementation, removing some performance enhancements and error handling
-clutter in the interest of clarity.
+clutter in the interest of clarity. The complete implementation is available
+for download, together with the Haste compiler, from our source repository at
+\lstinline!http://ekblad.cc/haste!.
 
@@
 with a compiler that generates the server binary, and once with one that
-generates Javascript for execution on the client. Conditional compilation is
-used for a select few functions, to enable slightly different behavior on the
-client and on the server as necessary. As we believe that a monadic
-programming style is a good fit for this application, we have chosen to use
-Haskell as the base of our solution, which by default leads us to choose GHC
-as our server-side compiler. For reasons outlined in section \ref{sec:haste},
-the Haste compiler was chosen to provide the client-side code.
+generates Javascript. Conditional compilation is used for a select few
+functions, to enable slightly different behavior on the client and on the
+server as necessary. As we believe that a monadic programming style is a good
+fit for this application, we have chosen to use Haskell as the base of our
+solution, which by default leads us to choose GHC as our server-side compiler.
+For reasons outlined in section \ref{sec:haste}, the Haste compiler was chosen
+to provide the client-side code.
 
@@
 The \lstinline!App! monad is where remote functions are declared, server state
-is initialized and program flow is finally handed over to the \lstinline!Client!
-monad. Its definition is as follows.
+is initialized and program flow is handed over to the \lstinline!Client! monad.
+Its definition is as follows.
 
@@
 type AppState = (CallID, [(CallID, Method)])
-type App = StateT AppState IO
+newtype App a = (StateT AppState IO a)
+  deriving (Functor, Applicative, Monad)
 \end{code}
@@
 next exported function, and its other state element contains a mapping from
-identifiers to exported functions. Of course, in reality, \lstinline!App! is a
-\lstinline!newtype!, to prevent arbitrary lifting of IO computations.
+identifiers to exported functions.
 
@@
 
-Aside from \lstinline!runClient!, this monad has two primary operations.
+Aside from \lstinline!runClient!, the \lstinline!App! monad has two primary
+operations, both of which have slightly different implementations on the client
+and server respectively.
 
@@
 
-The definition of \lstinline!liftServerIO! is very simple. Through the use of
-conditional compilation, on the client it simply ignores its argument and
-returns \lstinline!Useless!; recall that server-lifted IO computations are not
-observable on the client. On the server side, \lstinline!liftServerIO! executes
-its argument and wraps the computation's return value using the
+The client-side definition of \lstinline!liftServerIO! is as simple as can be.
+
+\begin{code}
+liftServerIO _ = App $ return Useless
+\end{code}
+
+Its argument is simply ignored, and an opaque dummy value is returned; recall
+that server-lifted IO computations are not observable on the client. The
+server-side implementation is only marginally more complex.
+
+\begin{code}
+liftServerIO m = App $ fmap Useful m
+\end{code}
+
+Its argument IO computation is executed, and the result is wrapped in the
 \lstinline!Useful! constructor. The counterpart function, \lstinline!mkUseful!
@@
 
+\begin{code}
+mkUseful :: Useless a -> Server a
+mkUseful (Useful x) = return x
+\end{code}
+
 \lstinline!export! is a more interesting function. Looking at its type, we see
@@
 class Exportable a where
-  mkExported :: a -> ([JSON] -> IO JSON)
-
-instance Serialize a =>
-         Exportable (Server a)
+  mkExported :: a -> ([JSON] -> Server JSON)
+
+instance Serialize a => Exportable (Server a) where
+  mkExported m = fmap toJSON m
+
 instance (Serialize a, Exportable b) =>
-         Exportable (a -> b)
+         Exportable (a -> b) where
+  mkExported f =
+    \(x:xs) -> mkExported (f $ fromJSON x) xs
 \end{code}
@@
 serializable value in the \lstinline!Server! monad can be exported. The
-\lstinline!mkExported! function makes use a well-known type class trick for
+\lstinline!mkExported! function makes use of a well-known type class trick for
 creating statically typed variadic functions, and works very much like the
 \lstinline!printf! function of Haskell's standard library.\ \cite{printf}
-From an \emph{n}-ary function, it creates a unary function which expects
-a list of \emph{n} JSON-encoded elements as its argument. When applied, its
-argument list is deserialized and passed to the exported function, and its
-return value is serialized. This allows us to easily bundle up all of an
-exported function's arguments and pass them across the network.
 
@@
 \begin{code}
-export _ = do
+export _ = App $ do
   (next_id, exports) <- get
@@
 \begin{code}
-export f = do
+export f = App $ do
   (next_id, exports) <- get
@@
 The second argument of the \lstinline!Export! constructor is, as the observant
-reader might have guessed, the list of arguments to be sent to the export.
-Essentially, an explicit closure is built up using the \lstinline!<.>! operator
-and sent off over the network for execution.
+reader might have guessed, the list of arguments to be sent to the export,
+to be updated as the exported function is applied using the \lstinline!<.>!
+operator.
 
@@
 (Export identifier args) <.> arg =
-  Export identifier (serialize arg : args)
+  Export identifier (toJSON arg : args)
 \end{code}
@@
 used to dispatch the proper functions for incoming calls from the client.
+
+\begin{code}
+onEvent :: [(CallID, Method)] -> JSON -> IO ()
+onEvent mapping incoming = do
+  let (nonce, identifier, args) = fromJSON incoming
+      Just f = lookup identifier mapping
+  result <- f args
+  webSocketSend $ toJSON (nonce, result)
+\end{code}
+
+The function corresponding to the RPC call's identifier is looked up in the
+identifier-function mapping and applied to the received list of arguments;
+the return value is bundled up with the a nonce, provided by the client to
+match each return value to the RPC call that produced it since there may be
+several RPC calls in progress at the same time, and sent back to the client.
+
+Note that during normal operation, it is not possible for the client to submit
+an RPC call with a non-existent call identifier, hence the irrefutable pattern
+match on \lstinline!Just f!. Should this pattern match fail, this is a sure
+sign of malicious tampering; the resulting exception is caught and the session
+is dropped as it is no longer meaningful to continue.
 
@@
   webSocketSend $
-    serialize (nonce, identifier, reverse args)  
+    toJSON (nonce, identifier, reverse args)  
   takeMVar mv
@@
 
-A nonce is used to pair up return values with the calls that produced them.
 After a call is dispatched, \lstinline!onServer! blocks, waiting for its
@@
 onMessage :: JSON -> Client ()
-onMessage msg = do
-  let (nonce, result) = deserialize msg
+onMessage response = do
+  let (nonce, result) = fromJSON response
   mv <- findResultMVar nonce
@@
 
-The corresponding server-side event loop, which handles incoming calls from the
-client, is similarly straightforward.
-
-\begin{code}
-onEvent :: [(CallID, Method)] -> JSON -> IO ()
-onEvent mapping msg = do
-  let (nonce, identifier, args) = deserialize msg
-      Just f = lookup identifier mapping
-  result <- f args
-  webSocketSend $ serialize (nonce, result)
-\end{code}
-
-The function corresponding to the RPC call's identifier is looked up in the
-identifier-function mapping generated by the \lstinline!App! computation and
-applied to the list of arguments; the return value is bundled up with the
-call's nonce and sent back to the client. Note that during normal operation,
-it is not possible for the client to submit an RPC call with a non-existent
-call identifier, hence the irrefutable pattern match on \lstinline!Just f!.
-Should this pattern match fail, this is a sure sign of malicious tampering;
-the resulting exception is caught and the session is dropped as it is no longer
-meaningful to continue.
+As we can see, the implementation of our programming model is very
+straightforward and requires no bothersome compiler modifications or language
+extensions and is thus easily portable to other Haskell compilers.
 
@@
         to be written as a single, monolithic program.
-  \item We also want our language to be practical, in the sense that its output
+  \item We also want our compiler to be practical, in the sense that its output
         should not be disproportionately large or slow. This is more important
@@
 out Fay which unfortunately lacks critical Haskell features, most notably type
-classes.
+classes, and thus has no one to one server-side counterpart compiler.
 
@@
 megabytes for very simple programs - which rules it out for our use case, in
-spite of providing slightly better compatibility with low-level GHC features
-than Haste. Similarly, the code produced by the UHC Javascript backend tends to
+spite of providing very good compatibility with even low-level GHC features.
+Similarly, the code produced by the UHC Javascript backend tends to
 be relatively slow due to it lacking the more mature optimization capabilities
@@
 client-server communication itself, are easily expressed as monads, well
-supported in Haskell thanks to do-notation. Add the fact that one of this
-work's authors is the original author of Haste, giving us a better position to
-troubleshoot any code based thereupon, and Haste becomes the better choice.
-
-\subsection{Design of the Haste compiler}
+supported in Haskell thanks to do-notation and a culture of monadic design
+patterns. Add the fact that one of this work's authors is the original author
+of Haste, giving us a better position to troubleshoot any code based thereupon,
+and Haste becomes the better choice.
+
+\subsection{Design goals}
 
@@
 
-GHCJS takes the opposite view, that faithful emulation of the GHC runtime is
-paramount. Of course, this brings an impressive degree of compatibility, at the
-cost of the blowup in code size mentioned previously.
-
 \subsection{Implementation overview}
@@
 the last intermediate representation used by GHC before the final code
-generation takes place and has several benefits for use as Haste's target
+generation takes place and has several benefits for use as Haste's source
 language:
@@
 than as a simple number, but recall that the \lstinline!Int! type is defined by
-GHC as \lstinline!data Int = I# Int#!.
+GHC as \lstinline!data Int = I# Int#! where \lstinline!Int#! is the primitive
+type for machine integers.
 
@@
 code. This function exploits Javascript's ability to parse and execute
-arbitrary strings at run-time using the \lstinline!eval! function, coupled with
+arbitrary strings at run time using the \lstinline!eval! function, coupled with
 the fact that functions in Haste and in Javascript share the same
@@
 The \lstinline!ffi! function is typed using the same method as the
-\lstinline!serializify! function described in section \ref{sec:impl}.
+\lstinline!mkExported! function described in section \ref{sec:impl}.
 When applied to one or more arguments instantiating the \lstinline!Marshal!
@@
 the inverse \lstinline!unpack! function is applied to its return value before
-it is passed back into Haskell land.
+it is passed back into the Haskell world.
 
@@
 the ground up to produce code for both client and server simultaneously. It
-utilizes the new attributes mechanism introduced in C++11 to designate
-functions and data to live on either client or server side. Any calls to a
-function on the other side of the network and attempts to access remote data
-are implicit, requiring no extra annotations or scaffolding at the call site.
+utilizes the new attributes mechanism introduced in C++11\ \cite{sepples11} to
+designate functions and data to live on either client or server side. Any calls
+to a function on the other side of the network and attempts to access remote
+data are implicit, requiring no extra annotations or scaffolding at the call
+site.
 
@@
 
-\paragraph{Sunroof} In contrast to the Conductance and Duetto, Sunroof is an
+\paragraph{Sunroof} In contrast to Conductance and Duetto, Sunroof is an
 embedded language. Implemented as a Haskell library, it allows the programmer
@@
 communication is appropriate, as well as helps prevents accidental transmission
-of sensitive information intended to stay on the client.
+of sensitive information intended to stay on either side of the network.
 
@@
 would be entirely straightforward. This implementation not only allows for
-greater flexibility, but also eliminates any potentially hairy interaction with
+greater flexibility, but also eliminates the need to tangle with complex
 compiler internals.
@@
 data representation: the client side of things must be able to obtain some
-representation of any function it may want to call. In our solution, this
-representation is obtained through the use of the \lstinline!export! function,
-which when executed on the server pairs a function with a unique identifier and
-on the client, and when executed on the client returns said identifier so that
-the client may now refer to the function. While this has the advantage of being
-simple to implement, one major drawback of this method is that all functions
-must be explicitly exported in the \lstinline!App! monad prior to being called
-over the network.
+representation of any function it may want to call on the server.
+
+In our solution, this representation is obtained through the use of the \lstinline!export! function, which when executed on the server pairs a function with a
+unique identifier and on the client, and when executed on the client returns
+said identifier so that the client may now refer to the function. While this
+has the advantage of being simple to implement, one major drawback of this
+method is that all functions must be explicitly exported in the \lstinline!App!
+monad prior to being called over the network.
 
@@
 introduces the notion of ``static values''; values which are known at compile
-time. Codifying this in the type system, to enable it to be used as a basis for
-remote procedure calls, unfortunately requires some major changes to the
-compiler. Cloud Haskell has a stopgap measure for unmodified compilers wherein
-a remote table, pairing values with unique identifiers, are kept. This explicit
-bookkeeping relies on the programmer to assign appropriate types to both values
-themselves and their identifiers, breaking type safety.
+time. Codifying this concept in the type system, to enable it to be used as a
+basis for remote procedure calls, unfortunately requires some major changes to
+the compiler. Cloud Haskell has a stopgap measure for unmodified compilers
+wherein a remote table, pairing values with unique identifiers, are kept.
+This explicit bookkeeping relies on the programmer to assign appropriate types
+to both values themselves and their identifiers, breaking type safety.
 
 The astute reader may notice that this is exactly what the \lstinline!export!
-does as well, the difference being that \lstinline!export! links the identifier
-to the value it represents on the type level, making it impossible to call
-non-existent remote functions and break the program's type safety in other ways.
+function does as well, the difference being that \lstinline!export! links the
+identifier to the value it represents on the type level, making it impossible
+to call non-existent remote functions and break the program's type safety in
+other ways.
 
@@
 an identifier around, the client would instead pass the actual defunctionalized
-function. This would have the added benefit of allowing functions to be
-arbitrarily composed before being remotely invoked.
+code to the server for execution. This would have the added benefit of allowing
+functions to be arbitrarily composed before being remotely invoked.
 
@@
 wisdom of allowing server-side execution of what is essentially arbitrary code
-sent from the client which, for web applications, is completely untrustworthy.
-While analyzing code for improper behavior is certainly possible, implementing
-and enforcing a sufficiently strict-yet-flexible security policy for remote
-execution would be an unwelcome burden on the programmer.
+sent from the client which, in a web application context, is completely
+untrustworthy. While analyzing code for improper behavior is certainly
+possible, designing and enforcing a security policy sufficiently strict to
+ensure correct behavior while flexible enough to be practically useful would be
+an unwelcome burden on the programmer.
 
@@
 
-\paragraph{Client-centricity} Unlike any of the related works, our approach
+\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
 takes a firm stand, regarding the client as the driver in the client-server
@@
 This is easily remedied however, by a slight compiler modification. Instead of
-using \lstinline!main :: IO ()! as the program entry point, introduce a function
-\lstinline!appMain :: App Done! for this purpose, and let the runtime system
-start the application by invoking \lstinline!runApp config appMain!, where
-\lstinline!config! contains the URL and port configuration for the server, given
-by the programmer using a compiler flag. As this modification only affects the
-program's entry point, which is already given ``special'' treatment in Haskell
-compilers, and may even be implemented using the plugin mechanism of GHC,
-requiring no modification of the compiler code at all, we believe this to be a
-practical, non-intrusive way to remove this limitation, and are planning on
-including this modification in our implementation.
+using \lstinline!main :: IO ()! as the program entry point, we introduce a
+function \lstinline!appMain :: App Done! for this purpose, and let the runtime
+system start the application by invoking \lstinline!runApp config appMain!,
+where \lstinline!config! contains the URL and port configuration for the
+server, given by the programmer using a compiler flag.
+
+As this modification only affects the program's entry point, which is already
+given ``special'' treatment in Haskell compilers, and may even be implemented
+using the plugin mechanism of GHC, requiring no modification of the compiler
+itself, we believe this to be a practical, non-intrusive way to remove this
+limitation, and are planning on releasing an implementation which includes
+this limitation.
 
@@
 any Haskell program using \lstinline!unsafePerformIO!, even subverting the type
-system, this is not a problem with our solution.
+system, ``solving'' \lstinline!unsafePerformIO! is not within the scope of this
+work.
 
@@
 
-While this would indeed provide an effective defence towards attackers and
+While this does indeed provide an effective defence towards attackers and
 programming mistakes alike, there is value in being able to tell the two apart,
@@
 
+\bibitem[Lubbers, Greco (2010)]{websockets}
+P. Lubbers and F. Greco. "Html5 web sockets: A quantum leap in scalability for the web." SOA World Magazine (2010).
+
 \bibitem[Marlow, Peyton Jones (2004)]{fastcurry}
@@
 
+\bibitem[Stroustrup (2013)]{sepples11}
+B. Stroustrup. (2014, January 21). "C++11 - the new ISO C++ standard." Retrieved March 1, 2014, from http://www.stroustrup.com/C++11FAQ.html.
+
 \bibitem[Taylor (2013)]{printf}

### 2014-03-01 paper.tex
Remove "language du jour" part; unintended negative connotations.

@@
 writing the one in Javascript, which is notorious even among its proponents for
-being wonky and error-prone, and the other in the current server language
-\emph{du jour}.
-Then, two are glued together using whichever home-grown network protocol seems
-to fit the application. However, most web applications are conceptually
-monolithic entities, making this forced split an undesirable hindrance which
-introduces new possibilities for defects, adds development overhead and
-prevents code reuse.
+being wonky and error-prone, and the other in any compiled or
+server-interpreted language. Then, two are glued together using whichever
+home-grown network protocol seems to fit the application. However, most web
+applications are conceptually monolithic entities, making this forced split an
+undesirable hindrance which introduces new possibilities for defects, adds
+development overhead and prevents code reuse.
 

### 2014-03-01 paper.tex
Explain that "no compiler support" means that we need a HS2JS compiler.

@@
 which are not explicitly declared as server-side or client-side are usable by
-either side. Our solution is implemented as a Haskell library requiring minimal
-to no compiler support, ensuring portability and modularity.
-
-As our implementation depends on the Haste Haskell to Javascript compiler, this
-paper also goes into some detail about its design and implementation.
+either side.
+
+Recent advances in compiler technology from functional languages to Javascript
+have led to a wealth of compilers targeting the web space, and enabled the
+practical development of functional libraries and applications for
+the browser. This enables us to implement our solution as a simple Haskell
+library for any compiler capable of producing Javascript output, requiring no
+further modification to existing compilers.
+
+As our implementation targets the Haste Haskell to Javascript compiler, this
+paper also goes into some detail about its design and implementation as well
+as the alternatives available for compiling functional languages to a browser
+environment.
 

### 2014-03-01 paper.tex
appMain -> clientMain; remove explanation for Haste.App.

@@
 
-Our proposed model, dubbed ``Haste.App'' in an ingenious attempt at
-capitalizing on the recent popularity of the ``app'' buzzword, has the
-following properties:
+Our proposed model, dubbed ``Haste.App'', has the following properties:
 
@@
 \begin{code}
-appMain :: App Done
-appMain = do
+clientMain :: App Done
+clientMain = do
   greetings <- export helloServer
@@
 main :: IO ()
-main =
-  runApp (mkConfig "ws://localhost:1111" 1111) appMain
+main = runApp (mkConfig "ws://localhost:1111" 1111)
+              clientMain
 \end{code}
@@
 non-deterministic IO and allowing \lstinline!runApp! or its arguments to
-capture the result may lead to unpredictable behavior. The \lstinline!appMain!
+capture the result may lead to unpredictable behavior. The \lstinline!clientMain!
 function should for all intents and purposes be regarded as the program's entry
@@
 
-appMain :: App Done
-appMain = do
+clientMain :: App Done
+clientMain = do
   greetings <- export helloServer
@@
 To allow the use of server state while treating the final argument to
-\lstinline!runApp! (from here on always referred to as \lstinline!appMain!)
+\lstinline!runApp! (from here on always referred to as \lstinline!clientMain!)
 as the program's entry point, we may want to allow arbitrary IO computations to
@@
 \begin{code}
-appMain :: App Done
-appMain = do
+clientMain :: App Done
+clientMain = do
   recipients <- liftServerIO $ CC.newMVar []
@@
 main =
-  runApp (mkConfig "ws://localhost:1111" 1111) appMain
+  runApp (mkConfig "ws://localhost:1111" 1111)
+         clientMain
 \end{code}
@@
 using \lstinline!main :: IO ()! as the program entry point, we introduce a
-function \lstinline!appMain :: App Done! for this purpose, and let the runtime
-system start the application by invoking \lstinline!runApp config appMain!,
+function \lstinline!clientMain :: App Done! for this purpose, and let the runtime
+system start the application by invoking \lstinline!runApp config clientMain!,
 where \lstinline!config! contains the URL and port configuration for the

### 2014-03-01 paper.tex
Better title
Stress the point that IT'S THE CLIENT, DAMMIT!

@@
 
-\title{Type-safe client-server communication in functional web applications}
+\title{A seamless, client-centric programming model for type safe web applications}
 
@@
 \end{code}
+
+This is the entry point of our application. Recall that Haste.App applications
+are written from the client's perspective, making this the starting point of
+the client-side code.
 

### 2014-03-01 paper.tex
Use Koen's abstract. It's about 20% cooler.

@@
 \begin{abstract}
-Developing web applications, especially applications which need to communicate
-with a server, is often a painful experience. What is conceptually a single
-monolithic application is forcefully split into a client and a server
-explicitly communicating over the network, with all the code duplication,
-complexity and lost type safety this entails.
-
-We propose instead a development model in which a web application is written
-and compiled as a whole, allowing for synchronous, type safe client-server
-communication without the need for explicit networking code, as well as the
-sharing of code between client and server.
+We propose a new programming model for web applications which is (1)
+seamless; one program and one language is used to produce code for
+both client and server, (2) client-centric; the programmer takes the
+viewpoint of the client that runs code on the server rather than the
+other way around, and (3) functional and type-safe; everything is
+implemented as a Haskell library that implicitly takes care of all
+networking code. Our aim is to improve the painful and error-prone
+experience of today's standard development methods, in which clients
+and servers are coded in different languages and communicate with each
+other using ad-hoc protocols. We present the design of our library
+called Haste.App, an example web application that uses it, and discuss
+the implementation and the compiler technology on which it depends.
 \end{abstract}

### 2014-03-01 paper.tex
Remove besserwisser words (of course, obviously, straightforward, etc.)

@@
 The \lstinline!export! function takes an arbitrary function, provided that all
-its arguments as well as its return value are serializable through the aptly
-named \lstinline!Serialize! type class, and produces a typed identifier which
+its arguments as well as its return value are serializable through the
+\lstinline!Serialize! type class, and produces a typed identifier which
 may be used to refer to the exported function. In this example, the type of
@@
 After the \lstinline!export! call, we enter the domain of client-exclusive code
-with the application of \lstinline!runClient!. This function, unsurprisingly,
-execute computations in the \lstinline!Client! monad which is essentially an
+with the application of \lstinline!runClient!. This function, execute
+computations in the \lstinline!Client! monad which is essentially an
 IO monad with cooperative multitasking added on top, to mitigate the fact that
@@
 
-This part is entirely straightforward; the URL of the server is provided for
-the benefit of the client and the port to listen on is provided for the
-benefit of the server. The final argument is the main function of our
-application. Figure \ref{lst:hello-server} gives the code of this example in
-its entirety.
+The URL of the server is provided for the benefit of the client and the port
+to listen on is provided for the benefit of the server. The final argument is
+the main function of our application. Figure \ref{lst:hello-server} gives the
+code of this example in its entirety.
 
@@
 
-The client implementation of \lstinline!export! is straightforward.
+The client implementation of \lstinline!export! is not very complex.
 
@@
 
-As we can see, the implementation of our programming model is very
-straightforward and requires no bothersome compiler modifications or language
-extensions and is thus easily portable to other Haskell compilers.
+As we can see, the implementation of our programming model is rather simple
+and requires no bothersome compiler modifications or language extensions,
+and is thus easily portable to other Haskell compilers.
 
@@
 In order to allow the same language to be used on both client and server, we
-obviously need some way to compile that language into Javascript. To this end,
-we make use of the Haste compiler\ \cite{haste}, started as an MSc thesis and
-continued as part of this work. Haste builds on the GHC compiler to provide
-the full Haskell language, including most GHC-specific extensions, in the
-browser.
+need some way to compile that language into Javascript. To this end, we make
+use of the Haste compiler\ \cite{haste}, started as an MSc thesis and continued
+as part of this work. Haste builds on the GHC compiler to provide the full
+Haskell language, including most GHC-specific extensions, in the browser.
 
@@
 Javascript \lstinline!Number! type, which is defined as the IEEE754 double
-precision floating point type. This obviously adds some overhead to operations
-on integers as overflow and non-integer divisions must be handled. However, this
+precision floating point type. This adds some overhead to operations on
+integers as overflow and non-integer divisions must be handled. However, this
 is common practice in hand-written Javascript as well, and is generally handled
@@
 is represented as \lstinline![0, 42]!, the leading \lstinline!0! representing
-the zeroth constructor of the \lstinline!Int! type and the \lstinline!42!,
-obviously, representing the ``machine'' integer. It may seem strange that a
-limited precision integer is represented using one level of indirection rather
+the zeroth constructor of the \lstinline!Int! type and the \lstinline!42!
+representing the ``machine'' integer. It may seem strange that a limited
+precision integer is represented using one level of indirection rather
 than as a simple number, but recall that the \lstinline!Int! type is defined by
@@
 forth between Haskell and Javascript, as long as she can come up with a way to
-translate this value between its Haskell and Javascript representations.
-Of course, not performing any translation at all is also a valid
-``translation'', which allows Haskell code to store any Javascript value for
-later retrieval without inspecting it and vice versa. The example given in
-figure \ref{lst:ffi} implements mutable variables using this custom Javascript
-interface.
+translate this value between its Haskell and Javascript representations. Not
+performing any translation at all is also a valid ``translation'', which allows
+Haskell code to store any Javascript value for later retrieval without
+inspecting it and vice versa. The example given in figure \ref{lst:ffi}
+implements mutable variables using this custom Javascript interface.
 
@@
 
-There are several benefits to this method, the most obvious being that new
+There are several benefits to this method, the most prominent being that new
 marshallable types can be added by simply instantiating a type class. Thanks
@@
 must be written in two separate languages - any code used to generate
-Javascript must of course be built solely from the primitives of the Sunroof
-language, precluding use of general Haskell code - makes code reuse hard.
+Javascript must be built solely from the primitives of the Sunroof
+language in order to be compilable into Javascript, precluding use of general
+Haskell code - makes code reuse hard.
 
@@
 modifying our implementation to use GHCJS or even the Javascript backend of UHC
-would be entirely straightforward. This implementation not only allows for
-greater flexibility, but also eliminates the need to tangle with complex
-compiler internals.
+would be trivial. This implementation not only allows for greater flexibility,
+but also eliminates the need to tangle with complex compiler internals.
 
 \paragraph{Inspiration and alternatives to \lstinline!export!} One crucial
-aspect of implementing cross-network function calls is, of course, the issue of
-data representation: the client side of things must be able to obtain some
+aspect of implementing cross-network function calls is the issue of data
+representation: the client side of things must be able to obtain some
 representation of any function it may want to call on the server.
@@
 insufficiently expressive, extending our existing model with an
-\lstinline!onClient! counterpart to \lstinline!onServer! would be relatively
-straightforward.
+\lstinline!onClient! counterpart to \lstinline!onServer! is a simple task.
 
@@
 communication with third parties and may thus, inadvertently or not, leak
-information about the application state. This is obviously highly undesirable
+information about the application state. This is of course highly undesirable
 for many applications, which is why there is ongoing work in controlling the

### 2014-03-01 paper.tex
Drastically cut down on "choosing a compiler"

@@
 After the \lstinline!export! call, we enter the domain of client-exclusive code
-with the application of \lstinline!runClient!. This function, execute
+with the application of \lstinline!runClient!. This function executes
 computations in the \lstinline!Client! monad which is essentially an
@@
 type AppState = (CallID, [(CallID, Method)])
-newtype App a = (StateT AppState IO a)
+newtype App a = App (StateT AppState IO a)
   deriving (Functor, Applicative, Monad)
@@
 
-In the face of this multitude of options, why did we settle on Haste? We make
-the following demands of a candidate source language:
+Essentially, our approach is portable to any language or compiler with the
+following properties:
 
@@
         to be written as a single, monolithic program.
-  \item We also want our compiler to be practical, in the sense that its output
-        should not be disproportionately large or slow. This is more important
-        for the client part, where we have to contend with slow connections and
-        the overhead of running within an interpreter, than for the server part.
+  \item We want the language to provide decent support for a monadic
+        programming style, as we our abstractions for cooperative multitasking
+        and synchronous client-server communication are neatly expressible in
+        this style.
 \end{itemize}
 
-Our first criterion rules out the dynamically typed Erlang. The second rules
-out Fay which unfortunately lacks critical Haskell features, most notably type
-classes, and thus has no one to one server-side counterpart compiler.
-
-This still leaves us with four options: Clean, GHCJS, Haste and UHC. However,
-the output from GHCJS is prohibitively large - often in the range of several
-megabytes for very simple programs - which rules it out for our use case, in
-spite of providing very good compatibility with even low-level GHC features.
-Similarly, the code produced by the UHC Javascript backend tends to
-be relatively slow due to it lacking the more mature optimization capabilities
-of GHC, as well as being in the same ballpark as GHCJS for size. Haste, in
-contrast, produces code which is generally one to two orders of magnitude
-smaller than that produced by GHCJS or UHC.\ \cite{haste}
-
-This leaves us choosing between Clean and Haste for our client compiler. Most
-of our abstractions, both the cooperative multitasking and the type-safe
-client-server communication itself, are easily expressed as monads, well
-supported in Haskell thanks to do-notation and a culture of monadic design
-patterns. Add the fact that one of this work's authors is the original author
-of Haste, giving us a better position to troubleshoot any code based thereupon,
-and Haste becomes the better choice.
+While several of the aforementioned compilers fullfil these criteria, the
+choice between them becomes almost arbitrary. We chose to base our solution on
+Haste as we, by virtue of its authorship, have an intimate knowledge of its
+internal workings, strengths and weaknesses. Without doubt, others may see many
+reasons to make a different choice.
 
@@
 runtime in Javascript. In the same vein, a conscious effort is made to generate
-as ``regular-looking'' code as possible. In some rare cases, this means having
-to sacrifice compatibility with low-level GHC features. Most notably, Haste
-does not support weak references or pre-emptive multitasking, as these would
-require modifications to the runtime system and generated code which confound
-the structure of the resulting Javascript and have a detrimental impact on code
-size and execution speed.
+as ``regular-looking'' code as possible. This results in large gains in code
+compactness and execution speed - in our experiences, compiler Haste programs
+are often an order of magnitude smaller and several times faster than an
+equivalent program compiled with some other Haskell to Javascript compiler.
+
+In some rare cases, however, this means having to sacrifice compatibility with
+low-level GHC features. Most notably, Haste does not support weak references
+or pre-emptive multitasking, as these would require modifications to the
+runtime system and generated code which confound the structure of the
+resulting Javascript and have a detrimental impact on code size and execution
+speed.
 

### 2014-03-01 paper.tex
Add blurb about why all of a sudden there's a bunch of Haste stuff.
Reference Cloud Haskell in implementation section.

@@
 
+This concept of client-side identifiers being sent to the server and used as
+indices into a table mapping identifiers to remotely accessible functions is an
+extension of the concept of ``static values'' introduced by Epstein et al with
+Cloud Haskell\ \cite{cloudhaskell}, which is discussed further in section
+\ref{sec:cloudhaskell}.
+
 \begin{code}
@@
 
-The second argument of the \lstinline!Export! constructor is, as the observant
-reader might have guessed, the list of arguments to be sent to the export,
-to be updated as the exported function is applied using the \lstinline!<.>!
-operator.
+The second argument of the \lstinline!Export! constructor is the list of
+arguments to be sent to the export, to be updated as the exported function is
+applied using the \lstinline!<.>! operator.
 
@@
 Haskell language, including most GHC-specific extensions, in the browser.
+
+As Haste has not been published elsewhere, we describe here some key elements
+of its design and implementation which are pertinent to this work.
 
@@
 
-\paragraph{Inspiration and alternatives to \lstinline!export!} One crucial
-aspect of implementing cross-network function calls is the issue of data
-representation: the client side of things must be able to obtain some
+\paragraph{Inspiration and alternatives to \lstinline!export!}
+\label{sec:cloudhaskell}
+One crucial aspect of implementing cross-network function calls is the issue of
+data representation: the client side of things must be able to obtain some
 representation of any function it may want to call on the server.

### 2014-03-01 paper.tex
Add client centricity to conclusions.

@@
 
-Our model is based on a simple blocking concurrency model rather than explicit
-continuations which is well suited for use with a GUI programming style based on
+Our model is client-centric, in that the client drives the application while
+the server takes on the role of passively serving client requests, and is based
+on a simple blocking concurrency model rather than explicit continuations.
+It is well suited for use with a GUI programming style based on
 self-contained processes with local state, and requires essentially no

### 2014-03-01 paper.tex
Note that Conductance and Duetto are quite young and quite unpublished.

@@
 possibility for both client and server to seamlessly modify variables located
-on the opposite end of the network.
+on the opposite end of the network. Conductance is quite new and has no
+relevant publications. It is, however, used for several large scale web
+applications.
 
@@
 data are implicit, requiring no extra annotations or scaffolding at the call
-site.
-
-Much like Conductance, Duetto suffers somewhat from its heritage: while the
+site. Duetto is still a highly experimental project, its first release being
+only a few months old, and has not been published in any academic venue.
+
+Like Conductance, Duetto suffers somewhat from its heritage: while the
 client-side code is not memory-unsafe, as it is not possible to generate

### 2014-03-01 paper.tex
Clarify that it's horrible when programs diverge.

@@
 main = do
-  captured <- randomRIO (False, True)
+  exportTwo <- randomRIO (False, True)
 
   runApp config $ do
-    diverged <- export $ \transmitted -> do
-      if captured == transmitted
-        then return ``All OK''
-        else return ``Diverged''
+    when exportTwo $ do
+      _ <- export (return () :: Server ())
+      return ()
+    fun <- export $ return "I'm a little teapot"
 
     runClient $ do
-      divergence <- onServer (diverged <.> captured)
-      alert divergence
-\end{code}
-
-The dialog box displayed by this program will read ``All OK'' roughly half
-of the times it is run, while in half of the runs it will read ``Diverged''.
-The reason for this is that the \lstinline!randomRIO! computation will be
-executed twice - once on the client, and once on the server. Whenever the
-computation returns differening results, the client and the server will have
-conflicting opinions on the actual value of \lstinline!captured!, causing any
-code which - quite reasonably - assumes that variables are consistent between
-client and server to fail.
+      onServer fun >>= alert
+\end{code}
+
+Whenever \lstinline!exportTwo! gets assigned a \lstinline!True! value, this
+application will call \lstinline!export! twice, assigning the identifier
+\lstinline!0! to the first exported computation, which is discarded, and
+\lstinline!1! to the second, \lstinline!fun!, which is kept and invoked.
+If, on the other hand, \lstinline!exportTwo! is \lstinline!False!,
+only the second computation will be exported, and will thus be assigned the
+identifier \lstinline!1!.
+
+This is perfectly fine as long as the client and server both assign the
+same identifier to \lstinline!fun!. However, since \lstinline!randomRIO! is
+not deterministic and since all computations called by \lstinline!main! will be
+executed on both the client and the server, there is a possibility that
+\lstinline!fun! will be assigned the identifier \lstinline!0! on the client and
+\lstinline!1! on the server. In this scenario, when the client attempts to call
+the \lstinline!return "I'm a little teapot"! computation over the network,
+the server instead executes the \lstinline!return ()! computation, which
+returns \lstinline!()! rather than \lstinline!String!, breaking the type safety
+of our program!
 

### 2014-03-01 paper.tex
Add real world applications to future work.

@@
 
-\subsection{The chatbox, revisited}
+\subsection{The chatbox, revisited}\label{sec:chatboxrevisited}
 
@@
 
+\paragraph{Real world applications} As Haste.App is quite new and experimental,
+it has yet to be used in the creation of large scale applications. While we
+have used it to implement some small applications, such as a spaced repetition
+vocabulary learning program and a more featureful variant on the chatbox
+example given in section \ref{sec:chatboxrevisited}, further investigation of
+its suitability for larger real world applications through the development of
+several larger scale examples is an important area of future work.
+
 \section{Conclusion}

### 2014-03-01 paper.tex
Add paragraph about why code sharing is important: GAEMS!!11

@@
 
+Interactive multiplayer games is one type of application where this code
+sharing may have a large impact. In order to ensure that players are not
+cheating, a game server must keep track of the entire game state and send
+updates to clients at regular intervals. However, due to network latency,
+waiting for server input before rendering each and every frame is completely
+impractical. Instead, the usual approach is to have each client continuously
+compute the state of the game to the best of its knowledge, rectifying any
+divergence from the game's ``official'' state whenever an update arrives from
+the server. In this scenario, it is easy to see how reusing much of the same
+game logic between the client and the server would be very important.
+
 Any and all communication between client and server is both strongly typed

### 2014-03-01 paper.tex
Replace monolithic by words without accidental negative connotations.

@@
 home-grown network protocol seems to fit the application. However, most web
-applications are conceptually monolithic entities, making this forced split an
+applications are conceptually single entities, making this forced split an
 undesirable hindrance which introduces new possibilities for defects, adds
@@
 described in the previous section. We propose an alternative programming model
-based on Haskell, in which web applications are written as monolithic entities
+based on Haskell, in which web applications are written as a single program
 rather than as two independent parts that just so happen to talk to each other.
@@
         suitable for server-side execution as we want our web applications
-        to be written as a single, monolithic program.
+        to be written and compiled as a single program.
   \item We want the language to provide decent support for a monadic

### 2014-03-01 paper.tex
Refer back to JS example to reinforce the advantage of typing.

@@
 
+\label{sec:jsexample}
 \begin{listingfloat}
@@
 Now that we have seen how to implement both network communication, we are ready
-to revisit the chatbox program, this time using our improved programming model.
-Since we are now writing the entire application, both client and server, as
-opposed to the client part from our motivating example, our program has three
-new responsibilities.
+to revisit the chatbox program from section \ref{sec:jsexample}, this time
+using our improved programming model. Since we are now writing the entire
+application, both client and server, as opposed to the client part from our
+motivating example, our program has three new responsibilities.
 
@@
 Having implemented our three server operations, all that's left is to tie them
-to the client. We start by creating an \lstinline!MVar! to hold the server
-state and exporting our three server operations.
+to the client. In this tying, we see our main advantage over the Javascript
+version in section \ref{sec:jsexample} in action: the \lstinline!export!
+function builds a strongly typed bridge between the client and the server,
+ensuring that any future enhancements to our chatbox program are made safely,
+in one place, instead of being spread about throughout two disjoint code bases.
 
@@
 
-%\acks
-%
-%Acknowledgments, if needed.
-
 % We recommend abbrvnat bibliography style.

### 2014-03-02 paper.tex
Add listing of core functions.

@@
 
-In order to make an RPC call using a identifier obtained from \lstinline!export!,
-we must supply it with an argument. This is done using the \lstinline!<.>!
-operator. It might be interesting to note that its type,
+In order to make an RPC call using an identifier obtained from
+\lstinline!export!, we must supply it with an argument. This is done using the
+\lstinline!<.>! operator. It might be interesting to note that its type,
 \linebreak
@@
 more complex client-server interactions, reducing this overhead significantly.
+
+A complete listing of the core functions in Haste.App is given in table
+\ref{tbl:api}, and their types are given in figure \ref{lst:api}.
+
+\begin{listingfloat}
+\begin{code}
+runApp       :: AppCfg -> App Done -> IO ()
+
+runClient    :: Client () -> App Done
+liftServerIO :: IO a -> App (Useless a)
+export       :: Exportable a
+             => a -> App (Export a)
+
+onServer     :: Export (Server a) -> Client a
+(<.>)        :: Serialize a
+             => Export (a -> b) -> a -> Export b
+
+mkUseful     :: Useless a -> Server a
+getSessionID :: Server SessionID
+\end{code}
+\caption{Types of the Haste.App core functions}
+\label{lst:api}
+\end{listingfloat}
+
+\begin{table}
+\renewcommand{\arraystretch}{1.5}
+\begin{center}
+\begin{tabular}{|r|l|}
+\hline
+Function & Purpose \\
+\hline
+\lstinline!runApp! & Begin execution of a Haste.App program. \\
+\lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!
+computation into the \lstinline!App! monad. Must be at the very end of
+an \lstinline!App! computation, which is enforced by the type system.} \\
+\lstinline!liftServerIO! & \parbox[t]{5cm}{Lift an IO computation into the
+\lstinline!App! monad. The computation and its result are exclusive to the
+server, as enforced by the type system, and are not observable on the client.} \\
+\lstinline!export! & \parbox[t]{5cm}{Make a server-side function available to
+be called remotely by the client.} \\
+\lstinline!onServer! & \parbox[t]{5cm}{Dispatch a remote call to the server and
+wait for its completion. The result of the remote computation is returned on
+the client after it completes.} \\
+\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!export!ed function to
+a serializable argument.} \\
+\lstinline!mkUseful! & \parbox[t]{5cm}{Extract the contents of a
+\lstinline!Useless! value produced by \lstinline!liftServerIO! on the server.} \\
+\lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
+the current session. This is a pure convenience function, to relieve
+programmers of the burden of session bookkeeping.} \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Core functions of Haste.App}
+\label{tbl:api}
+\end{table}
 
@@
 
-As this modification only affects the program's entry point, which is already
+As this modification only affects the program's entry, point which is already
 given ``special'' treatment in Haskell compilers, and may even be implemented

### 2014-05-02 paper.tex
Javascript -> JavaScript

@@
 The conventional method involves splitting your program into two logical parts,
-writing the one in Javascript, which is notorious even among its proponents for
+writing the one in JavaScript, which is notorious even among its proponents for
 being wonky and error-prone, and the other in any compiled or
@@
 
-Recent advances in compiler technology from functional languages to Javascript
+Recent advances in compiler technology from functional languages to JavaScript
 have led to a wealth of compilers targeting the web space, and enabled the
@@
 the browser. This enables us to implement our solution as a simple Haskell
-library for any compiler capable of producing Javascript output, requiring no
+library for any compiler capable of producing JavaScript output, requiring no
 further modification to existing compilers.
 
-As our implementation targets the Haste Haskell to Javascript compiler, this
+As our implementation targets the Haste Haskell to JavaScript compiler, this
 paper also goes into some detail about its design and implementation as well
@@
 
-Code written in Javascript, the only widely supported language for client-side
+Code written in JavaScript, the only widely supported language for client-side
 web applications, is often confusing and error-prone, much due to the
@@
 \end{code}
-\caption{Javascript chatbox implementation}
+\caption{JavaScript chatbox implementation}
 \label{lst:javascript-client}
@@
 IO monad with cooperative multitasking added on top, to mitigate the fact that
-Javascript has no native concurrency support. \lstinline!runClient! does not
+JavaScript has no native concurrency support. \lstinline!runClient! does not
 return, and is the only function with a return type of \lstinline!App Done!,
@@
 Having implemented our three server operations, all that's left is to tie them
-to the client. In this tying, we see our main advantage over the Javascript
+to the client. In this tying, we see our main advantage over the JavaScript
 version in section \ref{sec:jsexample} in action: the \lstinline!export!
@@
 and updates the chat log whenever one arrives. Note that unlike the
-\lstinline!onmessage! callback of the Javascript version of this example,
+\lstinline!onmessage! callback of the JavaScript version of this example,
 \lstinline!recvLoop! is acting as a completely self-contained process with
@@
 with a compiler that generates the server binary, and once with one that
-generates Javascript. Conditional compilation is used for a select few
+generates JavaScript. Conditional compilation is used for a select few
 functions, to enable slightly different behavior on the client and on the
@@
 As synchronous network communication is one of our stated goals, it is clear
-that we will need some kind of blocking primitive. Since Javascript does not
+that we will need some kind of blocking primitive. Since JavaScript does not
 support any kind of blocking, we will have to implement this ourselves.
@@
 computation into the monad, it is possible to implement cooperative
-multitasking on top of the non-concurrent Javascript runtime. This monad
+multitasking on top of the non-concurrent JavaScript runtime. This monad
 allows us to implement \lstinline!MVar!s as our blocking primitive, with the
@@
 In order to allow the same language to be used on both client and server, we
-need some way to compile that language into Javascript. To this end, we make
+need some way to compile that language into JavaScript. To this end, we make
 use of the Haste compiler\ \cite{haste}, started as an MSc thesis and continued
@@
 
-Haste is by no means the only Javascript-targeting compiler for a purely
+Haste is by no means the only JavaScript-targeting compiler for a purely
 functional language. In particular, the GHC-based GHCJS\ \cite{ghcjs} and
 UHC\ \cite{uhc} compilers are both capable of compiling standard Haskell into
-Javascript; the Fay\ \cite{fay} language was designed from the ground up to
+JavaScript; the Fay\ \cite{fay} language was designed from the ground up to
 target the web space using a subset of Haskell; and there exist solutions for
-compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to Javascript as
+compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to JavaScript as
 well.
@@
         the client-server communication channel.
-  \item The language must be compilable to both Javascript and a format
+  \item The language must be compilable to both JavaScript and a format
         suitable for server-side execution as we want our web applications
@@
 One such tradeoff is the decision to make use of the facilities provided by the
-web browser - most prominently the Javascript engine's native garbage
+web browser - most prominently the JavaScript engine's native garbage
 collector - as far as possible rather than implementing parts of the GHC
-runtime in Javascript. In the same vein, a conscious effort is made to generate
+runtime in JavaScript. In the same vein, a conscious effort is made to generate
 as ``regular-looking'' code as possible. This results in large gains in code
@@
 are often an order of magnitude smaller and several times faster than an
-equivalent program compiled with some other Haskell to Javascript compiler.
+equivalent program compiled with some other Haskell to JavaScript compiler.
 
@@
 runtime system and generated code which confound the structure of the
-resulting Javascript and have a detrimental impact on code size and execution
+resulting JavaScript and have a detrimental impact on code size and execution
 speed.
@@
         lambda calculus. When generating code for a high level target language
-        such as Javascript, where functions are first class objects, this
+        such as JavaScript, where functions are first class objects, this
         allows for a higher level translation than when doing traditional
@@
         application of most other functions. This allows for compiling most
-        function applications into simple Javascript function calls, limiting
+        function applications into simple JavaScript function calls, limiting
         the use of the slower dynamic techniques required to handle curried
@@
 The runtime data representation of Haste programs is kept as close to regular
-Javascript programs as possible. The numeric types are represented using the
-Javascript \lstinline!Number! type, which is defined as the IEEE754 double
+JavaScript programs as possible. The numeric types are represented using the
+JavaScript \lstinline!Number! type, which is defined as the IEEE754 double
 precision floating point type. This adds some overhead to operations on
 integers as overflow and non-integer divisions must be handled. However, this
-is common practice in hand-written Javascript as well, and is generally handled
-efficiently by Javascript engines.
+is common practice in hand-written JavaScript as well, and is generally handled
+efficiently by JavaScript engines.
 
@@
 
-Functions are represented as plain Javascript functions, one of the blessings
+Functions are represented as plain JavaScript functions, one of the blessings
 of targeting a high level language, and application can therefore be
-implemented as its Javascript counterpart in most cases. In the general case,
+implemented as its JavaScript counterpart in most cases. In the general case,
 however, functions may be curried. For such cases where the arity of an applied
@@
 
-\subsection{Interfacing with Javascript}
+\subsection{Interfacing with JavaScript}
 
@@
 low level. For this reason Haste sports its own method for interacting with
-Javascript as well, which allows the programmer to pass any value back and
-forth between Haskell and Javascript, as long as she can come up with a way to
-translate this value between its Haskell and Javascript representations. Not
+JavaScript as well, which allows the programmer to pass any value back and
+forth between Haskell and JavaScript, as long as she can come up with a way to
+translate this value between its Haskell and JavaScript representations. Not
 performing any translation at all is also a valid ``translation'', which allows
-Haskell code to store any Javascript value for later retrieval without
+Haskell code to store any JavaScript value for later retrieval without
 inspecting it and vice versa. The example given in figure \ref{lst:ffi}
-implements mutable variables using this custom Javascript interface.
+implements mutable variables using this custom JavaScript interface.
 
@@
 -- A MutableVar is completely opaque to Haskell code
--- and is only ever manipulated in Javascript. Thus,
+-- and is only ever manipulated in JavaScript. Thus,
 -- we use the Unpacked type to represent it,
@@
 The core of this interface consists of the \lstinline!ffi! function, which
-allows the programmer to create a Haskell function from arbitrary Javascript
-code. This function exploits Javascript's ability to parse and execute
+allows the programmer to create a Haskell function from arbitrary JavaScript
+code. This function exploits JavaScript's ability to parse and execute
 arbitrary strings at run time using the \lstinline!eval! function, coupled with
-the fact that functions in Haste and in Javascript share the same
+the fact that functions in Haste and in JavaScript share the same
 representation, to dynamically create a function object at runtime.
@@
 type class, the \lstinline!pack! function is applied to each argument,
-marshalling them into their respective Javascript representations, before they
+marshalling them into their respective JavaScript representations, before they
 are passed to the dynamically created function. When that function returns,
@@
 very non-intrusive, requiring only the use of the normal FFI to import
-Javascript's \lstinline!eval! function; no modification of the compiler is
+JavaScript's \lstinline!eval! function; no modification of the compiler is
 needed.
@@
 \paragraph{Conductance} Conductance is an application server built on
-StratifiedJS, a Javascript language extension which adds a few niceties such as
+StratifiedJS, a JavaScript language extension which adds a few niceties such as
 cooperative multitasking and more concise syntax for many common tasks.
@@
 While Conductance gets rid of the callback-based programming model endemic to
-regular Javascript, it still suffers from many of its usual drawbacks. In
-particular, the weak typing of Javascript poses a problem in that the
+regular JavaScript, it still suffers from many of its usual drawbacks. In
+particular, the weak typing of JavaScript poses a problem in that the
 programmer is in no way reprimanded by her tools for using server APIs
@@
 client-side code is not memory-unsafe, as it is not possible to generate
-memory-unsafe Javascript code, its server-side counterpart unfortunately is.
+memory-unsafe JavaScript code, its server-side counterpart unfortunately is.
 Our reservations expressed about how network communication in Duetto can be
@@
 embedded language. Implemented as a Haskell library, it allows the programmer
-to use Haskell to write code which is compiled to Javascript and executed on
-the client. The language can best be described as having Javascript semantics
+to use Haskell to write code which is compiled to JavaScript and executed on
+the client. The language can best be described as having JavaScript semantics
 with Haskell's type system. Communication between client and server is
@@
 must be written in two separate languages - any code used to generate
-Javascript must be built solely from the primitives of the Sunroof
-language in order to be compilable into Javascript, precluding use of general
+JavaScript must be built solely from the primitives of the Sunroof
+language in order to be compilable into JavaScript, precluding use of general
 Haskell code - makes code reuse hard.
@@
 Our programming model is implemented as a library, assuming only two Haskell
-compilers, one targeting Javascript and one targeting the programmer's server
-platform of choice. While we use Haste as our Javascript-targeting compiler,
-modifying our implementation to use GHCJS or even the Javascript backend of UHC
+compilers, one targeting JavaScript and one targeting the programmer's server
+platform of choice. While we use Haste as our JavaScript-targeting compiler,
+modifying our implementation to use GHCJS or even the JavaScript backend of UHC
 would be trivial. This implementation not only allows for greater flexibility,
@@
 compilers, while problematic due to the differing natures of native and
-Javascript compilation respectively, would be a significant improvement in this
+JavaScript compilation respectively, would be a significant improvement in this
 area.
@@
 \bibitem[Guthrie (2014)]{jserlang}
-G. Guthrie. (2014, January 1). "Your transpiler to Javascript toolbox". Retrieved March 1, 2014, from http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/.
+G. Guthrie. (2014, January 1). "Your transpiler to JavaScript toolbox". Retrieved March 1, 2014, from http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/.
 
@@
 \bibitem[Nazarov (2012)]{ghcjs}
-V. Nazarov. "GHCJS Haskell to Javascript Compiler". Retrieved March 1, 2014, from https://github.com/ghcjs/ghcjs.
+V. Nazarov. "GHCJS Haskell to JavaScript Compiler". Retrieved March 1, 2014, from https://github.com/ghcjs/ghcjs.
 

### 2014-05-02 paper.tex
Update related works intro, add AFAX section.

@@
 
-Other approaches to this problem exist, notably in the form of the
-Conductance\ \cite{conductance}, Duetto\ \cite{duetto} and
-Sunroof\ \cite{sunroof} projects, all of which implement some synchronous
-abstraction over the network between the client and the server, much like our
-solution.
-
-\paragraph{Conductance} Conductance is an application server built on
-StratifiedJS, a JavaScript language extension which adds a few niceties such as
-cooperative multitasking and more concise syntax for many common tasks.
+Several other approaches to the seamless client-server interaction exist. In
+general, these attacks tend to be of the ``all or nothing'' variety,
+introducing new languages or otherwise requiring custom full stack solutions.
+In contrast, our solution can be implemented entirely as a library and is
+portable to any pair of compilers supporting typed monadic programming.
+Moreover, Haste.App has a quite simple and controlled programming model with a
+clearly defined controller, which stands in contrast to most related work which
+embraces a more flexible but also more complex programming model.
+
+The more notable approaches to the problem are discussed further in this section.
+
+\paragraph{Conductance} Conductance\ \cite{conductance} is an application
+server built on StratifiedJS, a JavaScript language extension which adds a few
+niceties such as cooperative multitasking and more concise syntax for many
+common tasks.
 Conductance uses an RPC-based model for client-server communication, much like
@@
 
-\paragraph{Duetto} Duetto is a C++ compiler targeting the web, written from
-the ground up to produce code for both client and server simultaneously. It
-utilizes the new attributes mechanism introduced in C++11\ \cite{sepples11} to
-designate functions and data to live on either client or server side. Any calls
-to a function on the other side of the network and attempts to access remote
-data are implicit, requiring no extra annotations or scaffolding at the call
-site. Duetto is still a highly experimental project, its first release being
-only a few months old, and has not been published in any academic venue.
+\paragraph{Duetto} Duetto\ \cite{duetto} is a C++ compiler targeting the web,
+written from the ground up to produce code for both client and server
+simultaneously.
+It utilizes the new attributes mechanism introduced in C++11\ \cite{sepples11}
+to designate functions and data to live on either client or server side.
+Any calls to a function on the other side of the network and attempts to access
+remote data are implicit, requiring no extra annotations or scaffolding at the
+call site. Duetto is still a highly experimental project, its first release
+being only a few months old, and has not been published in any academic venue.
 
@@
 
-\paragraph{Sunroof} In contrast to Conductance and Duetto, Sunroof is an
-embedded language. Implemented as a Haskell library, it allows the programmer
-to use Haskell to write code which is compiled to JavaScript and executed on
-the client. The language can best be described as having JavaScript semantics
-with Haskell's type system. Communication between client and server is
-accomplished through the use of ``downlinks'' and ``uplinks'', allowing for
-data to be sent to and from the client respectively.
+\paragraph{Sunroof} In contrast to Conductance and Duetto,
+Sunroof\ \cite{sunroof} is an embedded language. Implemented as a Haskell
+library, it allows the programmer to use Haskell to write code which is
+compiled to JavaScript and executed on the client. The language can best be
+described as having JavaScript semantics with Haskell's type system.
+Communication between client and server is accomplished through the use
+of ``downlinks'' and ``uplinks'', allowing for data to be sent to and from the
+client respectively.
 
@@
 Haskell code - makes code reuse hard.
+
+\paragraph{AFAX} AFAX\ \cite{afax}, an F#-based solution, takes an approach
+quite similar to ours, using monads to allow client and server side to coexist
+in the same program. Unfortunately, using F# as the base of such a solution
+raises the issue of side effects. Since any expression in F# may be side
+effecting, it is quite possible with AFAX to perform a side effect on the client
+and then attempt to perform some action based on this side effect on the server.
+To cope with this, AFAX needs to introduce cumbersome extensions to the F# type
+system, making AFAX exclusive to Microsoft's F# compiler and operating system,
+whereas our solution is portable to any pair of Haskell compilers.
 
@@
 
+\bibitem[Petricek, Syme (2007)]{afax}
+T. Petricek, and Don Syme. "AFAX: Rich client/server web applications in F#." (2007).
+
 \bibitem[Peyton Jones (1992)]{stg}

### 2014-05-02 paper.tex
Address all language issues from R1.

@@
 being wonky and error-prone, and the other in any compiled or
-server-interpreted language. Then, two are glued together using whichever
+server-interpreted language. Then, the two are glued together using whichever
 home-grown network protocol seems to fit the application. However, most web
@@
 on the client, and the two communicate using type safe RPC calls. Functions
-which are not explicitly declared as server-side or client-side are usable by
+which are not explicitly declared as server side or client side are usable by
 either side.
@@
 
-Code written in JavaScript, the only widely supported language for client-side
+Code written in JavaScript, the only widely supported language for client side
 web applications, is often confusing and error-prone, much due to the
@@
 code contains no less than three asynchronous callbacks, two of which both read
-and and modify the application's global state. This makes the program flow
+and modify the application's global state. This makes the program flow
 non-obvious, and introduces unnecessary risk and complexity through the
@@
 
-There are many conveivable improvements to the mainstream web development model
+There are many conceivable improvements to the mainstream web development model
 described in the previous section. We propose an alternative programming model
@@
 are written from the client's perspective, making this the starting point of
-the client-side code.
+the client side code.
 
@@
 
-\subsection{Using server-side state}
+\subsection{Using server side state}
 
 While the Hello Server example illustrates how client-server communication is
-handled, most web applications need to keep some server-side state as well.
+handled, most web applications need to keep some server side state as well.
 At first glance, this would seem to be a trivial problem; can we not just
@@
 \end{code}
-\caption{Server-side state: a misguided first attempt}
+\caption{server side state: a misguided first attempt}
 \label{lst:bad-state}
@@
 
-While this may at first seem to solve the problem of server-side state, this
+While this may at first seem to solve the problem of server side state, this
 approach raises a number of questions. Does \lstinline!visitorref! live on the
@@
 client, the client has no obligation to even produce such a value.
-Figure \ref{lst:good-state} shows how to make proper use of server-side state.
+Figure \ref{lst:good-state} shows how to make proper use of server side state.
 
@@
 \end{code}
-\caption{Server-side state: doing it properly}
+\caption{server side state: doing it properly}
 \label{lst:good-state}
@@
 recipient list. Notice how the application's server state is passed in as the
-function's first \lstinline!Useless RcptList! argument.
+function's argument.
 
@@
 server, as enforced by the type system, and are not observable on the client.} \\
-\lstinline!export! & \parbox[t]{5cm}{Make a server-side function available to
+\lstinline!export! & \parbox[t]{5cm}{Make a server side function available to
 be called remotely by the client.} \\
@@
 fit for this application, we have chosen to use Haskell as the base of our
-solution, which by default leads us to choose GHC as our server-side compiler.
+solution, which by default leads us to choose GHC as our server side compiler.
 For reasons outlined in section \ref{sec:haste}, the Haste compiler was chosen
-to provide the client-side code.
+to provide the client side code.
 
@@
 As we can see, \lstinline!App! is a simple state monad, with underlying IO
-capabilities to allow server-side computations to be forked from within it.
+capabilities to allow server side computations to be forked from within it.
 Its \lstinline!CallID! state element contains the identifier to be given to the
@@
 all clients. This is necessary to ensure that any particular call identifier
-always refers to the same server-side function on all clients.
+always refers to the same server side function on all clients.
 
 After all common code has been executed, the program flow diverges between the
-client and the server; client-side, \lstinline!runClient! launches the
+client and the server; client side, \lstinline!runClient! launches the
 application's \lstinline!Client! computation whereas on the server, this
@@
 
-The client-side definition of \lstinline!liftServerIO! is as simple as can be.
+The client side definition of \lstinline!liftServerIO! is as simple as can be.
 
@@
 that server-lifted IO computations are not observable on the client. The
-server-side implementation is only marginally more complex.
+server side implementation is only marginally more complex.
 
@@
 
-This concept of client-side identifiers being sent to the server and used as
+This concept of client side identifiers being sent to the server and used as
 indices into a table mapping identifiers to remotely accessible functions is an
@@
 The function corresponding to the RPC call's identifier is looked up in the
-identifier-function mapping and applied to the received list of arguments;
-the return value is bundled up with the a nonce, provided by the client to
-match each return value to the RPC call that produced it since there may be
-several RPC calls in progress at the same time, and sent back to the client.
+identifier-function mapping and applied to the received list of arguments.
+The return value is paired with a nonce provided by the client to tie it to
+its corresponding RPC call, since there may be several such calls in progress
+at the same time. The pair is then sent back to the client.
 
@@
   \item The language must be compilable to both JavaScript and a format
-        suitable for server-side execution as we want our web applications
+        suitable for server side execution as we want our web applications
         to be written and compiled as a single program.
   \item We want the language to provide decent support for a monadic
-        programming style, as we our abstractions for cooperative multitasking
+        programming style, as our abstractions for cooperative multitasking
         and synchronous client-server communication are neatly expressible in
@@
 Like Conductance, Duetto suffers somewhat from its heritage: while the
-client-side code is not memory-unsafe, as it is not possible to generate
-memory-unsafe JavaScript code, its server-side counterpart unfortunately is.
+client side code is not memory-unsafe, as it is not possible to generate
+memory-unsafe JavaScript code, its server side counterpart unfortunately is.
 Our reservations expressed about how network communication in Duetto can be
@@
 
-Interactive multiplayer games is one type of application where this code
+Interactive multiplayer games are one type of application where this code
 sharing may have a large impact. In order to ensure that players are not
@@
 the compiler. Cloud Haskell has a stopgap measure for unmodified compilers
-wherein a remote table, pairing values with unique identifiers, are kept.
+wherein a remote table, pairing values with unique identifiers, is kept.
 This explicit bookkeeping relies on the programmer to assign appropriate types
@@
 unsuitable for our use case. Moreover, we are not entirely convinced about the
-wisdom of allowing server-side execution of what is essentially arbitrary code
+wisdom of allowing server side execution of what is essentially arbitrary code
 sent from the client which, in a web application context, is completely
@@
 or storage resource. The server may thus not call back into the client at
-arbitrary points but is instead limited to returning answers to client-side
+arbitrary points but is instead limited to returning answers to client side
 queries. This is clearly less flexible than the back-and-forth model of Sunroof
@@
 
-\paragraph{Preventing divergence} While our solution completely safe from the
-invocation point of the \lstinline!runApp! function onward, capturing
+\paragraph{Preventing divergence} While our solution is completely safe from
+the invocation point of the \lstinline!runApp! function onward, capturing
 IO-dependent variables introduced before this invocation may lead to the client

### 2014-05-02 paper.tex
F# -> F\#

@@
 
-\paragraph{AFAX} AFAX\ \cite{afax}, an F#-based solution, takes an approach
+\paragraph{AFAX} AFAX\ \cite{afax}, an F\#-based solution, takes an approach
 quite similar to ours, using monads to allow client and server side to coexist
-in the same program. Unfortunately, using F# as the base of such a solution
-raises the issue of side effects. Since any expression in F# may be side
+in the same program. Unfortunately, using F\# as the base of such a solution
+raises the issue of side effects. Since any expression in F\# may be side
 effecting, it is quite possible with AFAX to perform a side effect on the client
 and then attempt to perform some action based on this side effect on the server.
-To cope with this, AFAX needs to introduce cumbersome extensions to the F# type
-system, making AFAX exclusive to Microsoft's F# compiler and operating system,
-whereas our solution is portable to any pair of Haskell compilers.
+To cope with this, AFAX needs to introduce cumbersome extensions to the F\#
+type system, making AFAX exclusive to Microsoft's F\# compiler and operating
+system, whereas our solution is portable to any pair of Haskell compilers.
 
@@
 \bibitem[Petricek, Syme (2007)]{afax}
-T. Petricek, and Don Syme. "AFAX: Rich client/server web applications in F#." (2007).
+T. Petricek, and Don Syme. "AFAX: Rich client/server web applications in F\#." (2007).
 

### 2014-05-02 paper.tex
Remove split infinitive. Apparently, grammar can be controversial.
Update info on where to obtain Haste.

@@
 (character code 13) while the input box has focus. Finally, all that is left
-is to actually launch our application.
+is to launch our application.
 
@@
 clutter in the interest of clarity. The complete implementation is available
-for download, together with the Haste compiler, from our source repository at
-\lstinline!http://ekblad.cc/haste!.
+for download, together with the Haste compiler, from Hackage as well as from
+our website at \lstinline!http://haste-lang.org!.
 

### 2014-05-02 paper.tex
Add GWT to related work.

@@
 
+\paragraph{Google Web Toolkit} Google Web Toolkit\ \cite{gwt}, a Java
+compiler targeting the browser, provides its own solution to client-server
+interoperability as well. This solution is based on callbacks, forcing
+developers to code in a confusing continuation passing style. It also suffers
+from excessive boilerplate code and an error-prone configuration process.
+
 \paragraph{Duetto} Duetto\ \cite{duetto} is a C++ compiler targeting the web,
@@
 
+\bibitem[Wargolet (2011)]{gwt}
+S. Wargolet. "Google Web Toolkit. Technical report 12." University of Wisconsin-Platterville Department of Computer Science and Software Engineering, 2011.
+
 \end{thebibliography}

### 2014-05-02 paper.tex
Remove defensive Haste-justification.

@@
 compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to JavaScript as
-well.
+well. While the aforementioned compilers are the ones most interesting for
+purely functional programming, there exist a wealth of other
+JavaScript-targeting compilers, for virtually any language.
 
@@
 
-While several of the aforementioned compilers fullfil these criteria, the
-choice between them becomes almost arbitrary. We chose to base our solution on
-Haste as we, by virtue of its authorship, have an intimate knowledge of its
-internal workings, strengths and weaknesses. Without doubt, others may see many
-reasons to make a different choice.
-
-\subsection{Design goals}
-
-Haste is first and foremost intended to be a practical compiler; the ability
-to produce code which is lean and fast enough to be used in the real world is
-critical. To this end, we make several design choices and tradeoffs, some of
-which would be viewed as detrimental in a more purist context.
-
-One such tradeoff is the decision to make use of the facilities provided by the
-web browser - most prominently the JavaScript engine's native garbage
-collector - as far as possible rather than implementing parts of the GHC
-runtime in JavaScript. In the same vein, a conscious effort is made to generate
-as ``regular-looking'' code as possible. This results in large gains in code
-compactness and execution speed - in our experiences, compiler Haste programs
-are often an order of magnitude smaller and several times faster than an
-equivalent program compiled with some other Haskell to JavaScript compiler.
-
-In some rare cases, however, this means having to sacrifice compatibility with
-low-level GHC features. Most notably, Haste does not support weak references
-or pre-emptive multitasking, as these would require modifications to the
-runtime system and generated code which confound the structure of the
-resulting JavaScript and have a detrimental impact on code size and execution
-speed.
+As several of the aforementioned compilers fullfil these criteria, the
+choice between them becomes almost arbitrary. Indeed, as Haste.App is compiler
+agnostic, this decision boils down to one's personal preference. We chose to
+base our solution on Haste as we, by virtue of its authorship, have an intimate
+knowledge of its internal workings, strengths and weaknesses.
+Without doubt, others may see many reasons to make a different choice.
 
@@
 
-As previously mentioned, Haste offloads much of the heavy lifting of
-compilation - parsing, type checking, intermediate code generation and many
-optimizations - onto GHC, and takes over code generation after the STG
-generation step, at the very end of the compilation process. STG\ \cite{stg} is
-the last intermediate representation used by GHC before the final code
-generation takes place and has several benefits for use as Haste's source
-language:
+Haste offloads much of the heavy lifting of compilation - parsing,
+type checking, intermediate code generation and many optimizations - onto GHC,
+and takes over code generation after the STG generation step, at the very end
+of the compilation process. STG\ \cite{stg} is the last intermediate
+representation used by GHC before the final code generation takes place and
+has several benefits for use as Haste's source language:
 
@@
 immutability of Haskell, this model gives programmers a not-so-subtle hint
-as to how they may want to structure their programs. Should this model prove
-insufficiently expressive, extending our existing model with an
-\lstinline!onClient! counterpart to \lstinline!onServer! is a simple task.
+as to how they may want to structure their programs. Extending our existing
+model with an \lstinline!onClient! counterpart to \lstinline!onServer! would be
+a simple task, but we are not quite convinced that there is value in doing so.
 
@@
 itself, we believe this to be a practical, non-intrusive way to remove this
-limitation, and are planning on releasing an implementation which includes
-this limitation.
+limitation.
 

### 2014-05-02 paper.tex
Add Ocsigen to related work.

@@
 
+\paragraph{Ocsigen} Ocsigen\ \cite{ocsigen} enables the development of
+client-server web applications using O'Caml. It accomplishes typed, seamless
+communication by exposing mutable variables across the network. While this is a
+quite flexible programming model, we believe that this uncontrolled, implicit
+information flow makes programs harder to follow, debug, secure and optimize.
+
 \paragraph{AFAX} AFAX\ \cite{afax}, an F\#-based solution, takes an approach
@@
 
+\bibitem[Balat (2006)]{ocsigen}
+V. Balat. "Ocsigen: typing web interaction with objective Caml." Proceedings of the 2006 workshop on ML. ACM, 2006.
+
 \bibitem[Bracker, Gill (2014)]{sunroof}

### 2014-05-02 paper.tex
Add Opa to related work.

@@
 to exercise extra caution to avoid inadvertently sending large amounts of data
-over the network.
+over the network or leak sensitive information.
+
+\paragraph{Opa} The Opa framework\ \cite{opa}, much like Conductance, enables
+seamless communication within the JavaScript language itself.
+While Opa improves upon Conductance by introducing non-mandatory type checking
+to the JavaScript world, its communication model is based on implicit
+information flows, allowing the server to read and update mutable state on the
+client and vice versa. While this is a quite flexible programming model, we
+believe that this uncontrolled, implicit information flow makes programs harder
+to follow, debug, secure and optimize.
 
@@
 \paragraph{Ocsigen} Ocsigen\ \cite{ocsigen} enables the development of
-client-server web applications using O'Caml. It accomplishes typed, seamless
-communication by exposing mutable variables across the network. While this is a
-quite flexible programming model, we believe that this uncontrolled, implicit
-information flow makes programs harder to follow, debug, secure and optimize.
+client-server web applications using O'Caml. Much like Opa, it accomplishes
+typed, seamless communication by exposing mutable variables across the network,
+giving it many of the same drawbacks and benefits. While Ocsigen is a full
+stack solution, denying the developer some flexibility in choosing their tools,
+it should be noted that said stack is rather comprehensive and well tested.
 
@@
 
+\bibitem[The Opa framework for JavaScript (2014)]{opa}
+The Opa framework for JavaScript. Retrieved May 2, 2014, from http://opalang.org.
+
 \bibitem[Petricek, Syme (2007)]{afax}

### 2014-05-02 paper.tex
Update abstract, add Links, Ur/Web and HOP.

@@
 viewpoint of the client that runs code on the server rather than the
-other way around, and (3) functional and type-safe; everything is
-implemented as a Haskell library that implicitly takes care of all
-networking code. Our aim is to improve the painful and error-prone
-experience of today's standard development methods, in which clients
-and servers are coded in different languages and communicate with each
-other using ad-hoc protocols. We present the design of our library
-called Haste.App, an example web application that uses it, and discuss
+other way around, (3) functional and type-safe; and (4) portable;
+everything is implemented as a Haskell library that implicitly takes
+care of all networking code. Our aim is to improve the painful and
+error-prone experience of today's standard development methods,
+in which clients and servers are coded in different languages and
+communicate with each other using ad-hoc protocols.
+We present the design of our library called Haste.App,
+an example web application that uses it, and discuss
 the implementation and the compiler technology on which it depends.
@@
 
+\paragraph{HOP, Links, Ur/Web and others} In addition to solutions which work
+within existing languages, there are several languages specifically created to
+target the web domain. These languages tend to target not only the client and
+server tiers, but the database tier as well. These languages bring many new,
+interesting ideas to the table but suffers from not being compatible with any
+more established languages.
+
 \paragraph{Advantages to our approach} We believe that our approach has a

### 2014-05-05 paper.tex
Dedicated languages solve another problem.

@@
 \paragraph{HOP, Links, Ur/Web and others} In addition to solutions which work
-within existing languages, there are several languages specifically created to
-target the web domain. These languages tend to target not only the client and
-server tiers, but the database tier as well. These languages bring many new,
-interesting ideas to the table but suffers from not being compatible with any
-more established languages.
+within existing languages, there are several languages specifically crafted
+targeting the web domain. These languages target not only the client and server
+tiers but the database tier as well, and incorporate several interesting new
+ideas such as more expressive type systems and inclusion of typed inline XML
+code.\ \cite{hop}\cite{links}\cite{urweb} As our solution aims to bring typed,
+seamless communication into the existing Haskell ecosystem without language
+modifications, these languages solve a different set of problems.
 
@@
 
+\bibitem[Chlipala (2010)]{urweb}
+A. Chlipala. "Ur: statically-typed metaprogramming with type-level record computation." ACM Sigplan Notices. Vol. 45. No. 6. ACM, 2010.
+
 \bibitem[Claessen (1999)]{concurrencymonad}
 K. Claessen. "Functional Pearls: A poor man's concurrency monad." Journal of Functional Programming 9 (1999): 313-324.
+
+\bibitem[Cooper et al (1999)]{links}
+E. Cooper, S. Lindley, P. Wadler, and J. Yallop. Links: Web programming without tiers. In Formal Methods for Components and Objects (pp. 266-296). Springer Berlin Heidelberg, 2007.
 
@@
 
+\bibitem[Serrano et al (2006)]{hop}
+M. Serrano, E. Gallesio, and F. Loitsch. "Hop: a language for programming the web 2. 0." OOPSLA Companion. 2006.
+
 \bibitem[Stroustrup (2013)]{sepples11}

### 2014-05-07 paper.tex
Flip names, binaries -> executables, prev. work in intro.

@@
 
-\authorinfo{Koen Claessen and Anton Ekblad}
+\authorinfo{Anton Ekblad and Koen Claessen}
            {Chalmers University of Technology}
-           {\{koen,antonek\}@chalmers.se}
+           {\{antonek,koen\}@chalmers.se}
 
@@
 
-Several attempts have been made at tackling this problem, but none is, in our
-view, satisfactory. In this paper, we propose a functional programming model in
-which a web application is written as a single program from which client and
-server binaries are generated during compilation. Type annotations in the source
-program control which parts are executed on the server and which are executed
-on the client, and the two communicate using type safe RPC calls. Functions
-which are not explicitly declared as server side or client side are usable by
-either side.
+Several solutions to this problem have been proposed, as discussed in section
+\ref{sec:related}, but the perfect one has yet to be found. In this paper,
+we propose a functional programming model in which a web application is
+written as a single program from which client and server executables are
+generated during compilation. Type annotations in the source program control
+which parts are executed on the server and which are executed on the client,
+and the two communicate using type safe RPC calls. Functions which are not
+explicitly declared as server side or client side are usable by either side.
 
 Recent advances in compiler technology from functional languages to JavaScript
-have led to a wealth of compilers targeting the web space, and enabled the
+have led to a wealth of compilers targeting the web space, and have enabled the
 practical development of functional libraries and applications for
@@
 
-As our implementation targets the Haste Haskell to JavaScript compiler, this
-paper also goes into some detail about its design and implementation as well
-as the alternatives available for compiling functional languages to a browser
-environment.
+As our implementation targets the Haste Haskell to JavaScript
+compiler\ \cite{haste}, this paper also goes into some detail about its design
+and implementation as well as the alternatives available for compiling
+functional languages to a browser environment.
 
@@
   var msgbox = document.getElementById('message');
-  var sock = new WebSocket('ws://example.com);
+  var sock = new WebSocket('ws://example.com');
 
@@
 \subsection{Related work}
+\label{sec:related}
 

### 2014-05-07 paper.tex
Useless -> Remote, main is entry point

@@
   \item Our model takes the view that the client side is the main driver when
-        developing web applications and accordingly delegates the server to
+        developing web applications and accordingly assigns the server
         the role of a computational and/or storage resource, tasked with
@@
 \begin{code}
-clientMain :: App Done
-clientMain = do
+main :: App Done
+main = do
   greetings <- export helloServer
@@
 
-This is the entry point of our application. Recall that Haste.App applications
-are written from the client's perspective, making this the starting point of
-the client side code.
+The \lstinline!main! function is, as usual, the entry point of our application.
+In contrast to traditional applications which live either on the client or on
+the server and begin in the \lstinline!IO! monad, Haste.App applications live
+on both and begin execution in the \lstinline!App! monad which provides some
+crucial tools to facilitate typed communication between the two.
 
@@
 
-Finally, we need to launch our application and inform the client part of the
-whereabouts of its server counterpart. In this example the server will listen
-for connecting clients on port 1111, and since we will only run this
-application locally the client should attempt to contact the server over the
-loopback interface.
-
-\begin{code}
-main :: IO ()
-main = runApp (mkConfig "ws://localhost:1111" 1111)
-              clientMain
-\end{code}
-
-The URL of the server is provided for the benefit of the client and the port
-to listen on is provided for the benefit of the server. The final argument is
-the main function of our application. Figure \ref{lst:hello-server} gives the
-code of this example in its entirety.
-
-Like \lstinline!runClient!, \lstinline!runApp! does not return, and is intended
-to be the only computation executed by \lstinline!main!. In fact, performing
-non-deterministic IO and allowing \lstinline!runApp! or its arguments to
-capture the result may lead to unpredictable behavior. The \lstinline!clientMain!
-function should for all intents and purposes be regarded as the program's entry
-point. A non-intrusive method of enforcing this using the type system is
-described in section \ref{sec:limitations}, but is not used in the examples
-given here as it relies on the use of the GHC plugin system.
+Figure \ref{lst:hello-server} gives the code of this example in its entirety.
 
@@
 
-clientMain :: App Done
-clientMain = do
+main :: App Done
+main = do
   greetings <- export helloServer
@@
     onServer (greetings <.> name)
-
-main :: IO ()
-main = runApp (mkConfig "ws://localhost:1111" 1111)
-              clientMain
 \end{code}
@@
 handled, most web applications need to keep some server side state as well.
-At first glance, this would seem to be a trivial problem; can we not just
-create our state holding elements in \lstinline!main! as usual?
-Figure \ref{lst:bad-state} describes an attempt at a solution using this idea.
-
-\begin{listingfloat}
-\begin{code}
-main = do
-  visitorref <- newIORef 0
-  runApp config $ do
-    count <- export . liftIO $ do
-      atomicModifyIORef visitorref (\v -> (v+1, v+1))
-
-    runClient $ do
-      visitors <- onServer count
-      alert ("Your are visitor #" ++ show visitors)
-\end{code}
-\caption{server side state: a misguided first attempt}
-\label{lst:bad-state}
-\end{listingfloat}
-
-While this may at first seem to solve the problem of server side state, this
-approach raises a number of questions. Does \lstinline!visitorref! live on the
-client or the server, or perhaps on both? What happens if our state holding
-element is initialized in a non-deterministic way, and what if we assign it
-a value on the server and then attempt to read that value on the client?
-Recall that, as explained in section \ref{sec:helloserver}, the function
-passed to \lstinline!runApp! should for all intents and purposes be regarded as
-the program's entry point.
-
-To allow the use of server state while treating the final argument to
-\lstinline!runApp! (from here on always referred to as \lstinline!clientMain!)
-as the program's entry point, we may want to allow arbitrary IO computations to
-be lifted into the \lstinline!App! monad:
-
-\begin{code}
-main = runApp config $ do
-  visitorref <- liftIO $ newIORef 0
-  ...
-\end{code}
-
-However, this does not answer the question of where \lstinline!visitorref!
-lives or where any side effects caused by its creation will take place.
-To solve this problem once and for all, we need to introduce a way to lift
-arbitrary IO computations, but ensure that said computations are executed on
-the server and nowhere else. This is accomplished using a more restricted
-version of \lstinline!liftIO!:
-
-\begin{code}
-liftServerIO :: IO a -> App (Useless a)
+How can we create state holding elements for the server which are not
+accessible to the client?
+
+To accomplish this, we need to introduce a way to lift arbitrary IO
+computations, but ensure that said computations are executed on the server and
+nowhere else. This is accomplished using a more restricted version of
+\lstinline!liftIO!:
+
+\begin{code}
+liftServerIO :: IO a -> App (Remote a)
 \end{code}
@@
 In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
-is wrapped in the \lstinline!Useless! type. As the name implies, this type is
-completely opaque and thus useless unless its contained value can somehow be
-extracted, something which is only possible using a single function:
-
-\begin{code}
-mkUseful :: Useless a -> Server a
+is wrapped in the abstract \lstinline!Remote! type. As the name implies,
+values of this type live exclusively on the server. In order to inspect such a
+value is needs to be extracted somehow, something which is only possible using
+a single function:
+
+\begin{code}
+fromRemote :: Remote a -> Server a
 \end{code}
@@
 Therein lies the secret to restricting \lstinline!liftServerIO! to the
-server: the only function capable of turning a \lstinline!Useless! value into
+server: the only function capable of turning a \lstinline!Remote! value into
 a useful one lives in the server monad. Any client side code is thus free to
@@
 \begin{code}
-main =
-  runApp config $ do
-    uselessref <- liftServerIO $ newIORef 0
+main = do
+  remoteref <- liftServerIO $ newIORef 0
 
   count <- export $ do
-      r <- mkUseful uselessref
+    r <- fromRemote remoteref
     liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))
@@
 \begin{code}
-srvHello :: Useless RcptList -> Server ()
-srvHello uRcpts = do
-  recipients <- mkUseful uRcpts
+srvHello :: Remote RcptList -> Server ()
+srvHello remoteRcpts = do
+  recipients <- fromRemote remoteRcpts
   sid <- getSessionID
@@
 \begin{code}
-srvSend :: Useless RcptList -> String -> Server ()
-srvSend uRcpts message = do
-    rcpts <- mkUseful uRcpts
+srvSend :: Remote RcptList -> String -> Server ()
+srvSend remoteRcpts message = do
+    rcpts <- fromRemote remoteRcpts
     liftIO $ do
@@
 \begin{code}
-srvAwait :: Useless RcptList -> Server String
-srvAwait uRcpts = do
-  rcpts <- mkUseful uRcpts
+srvAwait :: Remote RcptList -> Server String
+srvAwait remoteRcpts = do
+  rcpts <- fromRemote remoteRcpts
   sid <- getSessionID
@@
 \begin{code}
-clientMain :: App Done
-clientMain = do
+main :: App Done
+main = do
   recipients <- liftServerIO $ CC.newMVar []
@@
 input box and send its contents off to the server whenever the user hits return
-(character code 13) while the input box has focus. Finally, all that is left
-is to launch our application.
-
-\begin{code}
-main :: IO ()
-main =
-  runApp (mkConfig "ws://localhost:1111" 1111)
-         clientMain
-\end{code}
+(character code 13) while the input box has focus.
 
 The discerning reader may be slightly annoyed at the need to extract the
-contents from \lstinline!Useless! values at each point of use. Indeed, in a
+contents from \lstinline!Remote! values at each point of use. Indeed, in a
 simple example such as this, the source clutter caused by this becomes a
@@
 runClient    :: Client () -> App Done
-liftServerIO :: IO a -> App (Useless a)
+liftServerIO :: IO a -> App (Remote a)
 export       :: Exportable a
@@
 
-mkUseful     :: Useless a -> Server a
+fromRemote   :: Remote a -> Server a
 getSessionID :: Server SessionID
@@
 a serializable argument.} \\
-\lstinline!mkUseful! & \parbox[t]{5cm}{Extract the contents of a
-\lstinline!Useless! value produced by \lstinline!liftServerIO! on the server.} \\
+\lstinline!fromRemote! & \parbox[t]{5cm}{Extract the contents of a
+\lstinline!Remote! value produced by \lstinline!liftServerIO! on the server.} \\
 \lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
@@
 \begin{code}
-data Useless a = Useful a | Useless
+data Remote a = Here a | There
 data Export a = Export CallID [JSON]
 
-liftServerIO :: IO a -> App (Useless a)
+liftServerIO :: IO a -> App (Remote a)
 export :: Exportable a => a -> App (Export a)
@@
 \begin{code}
-liftServerIO _ = App $ return Useless
+liftServerIO _ = App $ return There
 \end{code}
@@
 \begin{code}
-liftServerIO m = App $ fmap Useful m
+liftServerIO m = App $ fmap Here m
 \end{code}
@@
 Its argument IO computation is executed, and the result is wrapped in the
-\lstinline!Useful! constructor. The counterpart function, \lstinline!mkUseful!
-simply unwraps the \lstinline!Useful! constructor and returns the contained
+\lstinline!Remote! constructor. The counterpart function, \lstinline!fromRemote!
+simply unwraps the \lstinline!Here! constructor and returns the contained
 value in the \lstinline!Server! monad.
@@
 \begin{code}
-mkUseful :: Useless a -> Server a
-mkUseful (Useful x) = return x
+fromRemote :: Remote a -> Server a
+fromRemote (Here x) = return x
 \end{code}
@@
 
-\paragraph{Preventing divergence} While our solution is completely safe from
-the invocation point of the \lstinline!runApp! function onward, capturing
-IO-dependent variables introduced before this invocation may lead to the client
-and server execution paths diverging. Recall that our solution does not modify
-either of the compilers used. As such, both will produce code to execute all
-computations called by \lstinline!main!. Consider the following example:
-
-\begin{code}
-main = do
-  exportTwo <- randomRIO (False, True)
-
-  runApp config $ do
-    when exportTwo $ do
-      _ <- export (return () :: Server ())
-      return ()
-    fun <- export $ return "I'm a little teapot"
-
-    runClient $ do
-      onServer fun >>= alert
-\end{code}
-
-Whenever \lstinline!exportTwo! gets assigned a \lstinline!True! value, this
-application will call \lstinline!export! twice, assigning the identifier
-\lstinline!0! to the first exported computation, which is discarded, and
-\lstinline!1! to the second, \lstinline!fun!, which is kept and invoked.
-If, on the other hand, \lstinline!exportTwo! is \lstinline!False!,
-only the second computation will be exported, and will thus be assigned the
-identifier \lstinline!1!.
-
-This is perfectly fine as long as the client and server both assign the
-same identifier to \lstinline!fun!. However, since \lstinline!randomRIO! is
-not deterministic and since all computations called by \lstinline!main! will be
-executed on both the client and the server, there is a possibility that
-\lstinline!fun! will be assigned the identifier \lstinline!0! on the client and
-\lstinline!1! on the server. In this scenario, when the client attempts to call
-the \lstinline!return "I'm a little teapot"! computation over the network,
-the server instead executes the \lstinline!return ()! computation, which
-returns \lstinline!()! rather than \lstinline!String!, breaking the type safety
-of our program!
-
-This is easily remedied however, by a slight compiler modification. Instead of
-using \lstinline!main :: IO ()! as the program entry point, we introduce a
-function \lstinline!clientMain :: App Done! for this purpose, and let the runtime
-system start the application by invoking \lstinline!runApp config clientMain!,
-where \lstinline!config! contains the URL and port configuration for the
-server, given by the programmer using a compiler flag.
-
-As this modification only affects the program's entry, point which is already
-given ``special'' treatment in Haskell compilers, and may even be implemented
-using the plugin mechanism of GHC, requiring no modification of the compiler
-itself, we believe this to be a practical, non-intrusive way to remove this
-limitation.
-
-The astute reader may notice that the same applies to code using
-\lstinline!unsafePerformIO! to declare IO-dependent top-level values, and that
-this problem is unaffected by changing the type of the program's entry point.
-While this is indeed true, we feel that since it is quite possible to break
-any Haskell program using \lstinline!unsafePerformIO!, even subverting the type
-system, ``solving'' \lstinline!unsafePerformIO! is not within the scope of this
-work.
-
 \paragraph{Environment consistency} As our programming model uses two different

### 2014-05-07 paper.tex
Further emphasize that it's a library.

@@
 It is well suited for use with a GUI programming style based on
-self-contained processes with local state, and requires essentially no
-modification of existing tools or compilers.
+self-contained processes with local state, and requires no modification of
+existing tools or compilers, being implemented completely as a library.
 

### 2014-05-07 paper.tex
GWT, Opa, Conductance, Sunroof fleshed out a little.

@@
 
-\paragraph{Conductance} Conductance\ \cite{conductance} is an application
+\paragraph{Conductance and Opa} Conductance\ \cite{conductance} is an application
 server built on StratifiedJS, a JavaScript language extension which adds a few
@@
 
-\paragraph{Opa} The Opa framework\ \cite{opa}, much like Conductance, enables
-seamless communication within the JavaScript language itself.
-While Opa improves upon Conductance by introducing non-mandatory type checking
-to the JavaScript world, its communication model is based on implicit
-information flows, allowing the server to read and update mutable state on the
-client and vice versa. While this is a quite flexible programming model, we
-believe that this uncontrolled, implicit information flow makes programs harder
-to follow, debug, secure and optimize.
+The Opa framework\ \cite{opa}, another JavaScript framework, is an improvement
+over Conductance by introducing non-mandatory type checking to the JavaScript
+world. Its communication model is based on implicit information flows, allowing
+the server to read and update mutable state on the client and vice versa.
+While this is a quite flexible programming model, we believe that this
+uncontrolled, implicit information flow makes programs harder to follow, debug,
+secure and optimize.
 
@@
 interoperability as well. This solution is based on callbacks, forcing
-developers to code in a confusing continuation passing style. It also suffers
-from excessive boilerplate code and an error-prone configuration process.
+developers to write code in a continuation passing style. It also suffers
+from excessive boilerplate code and an error prone configuration process.
+The programming model shares Haste.App's client centricity, relegating the
+server to serving client requests.
 
@@
 language in order to be compilable into JavaScript, precluding use of general
-Haskell code - makes code reuse hard.
+Haskell code - makes code reuse hard. As the JavaScript DSL is executed from
+a native Haskell host, Sunroof's programming model can be said to be somewhat
+server centric, but with quite some flexibility due to its back and forth
+communication model.
 
@@
 
-\paragraph{Advantages to our approach} We believe that our approach has a
+\paragraph{Advantages of our approach} We believe that our approach has a
 number of distinct advantages to the aforementioned attacks on the problem.
@@
 
-\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
+\paragraph{Client-centricity} Unlike most related work, our approach
 takes a firm stand, regarding the client as the driver in the client-server

### 2014-05-07 paper.tex
Fix "we're using Haskell!!" six pages into a Haskell Symposium paper.

@@
 functions, to enable slightly different behavior on the client and on the
-server as necessary. As we believe that a monadic programming style is a good
-fit for this application, we have chosen to use Haskell as the base of our
-solution, which by default leads us to choose GHC as our server side compiler.
-For reasons outlined in section \ref{sec:haste}, the Haste compiler was chosen
-to provide the client side code.
+server as necessary. Using Haskell as the base language of our solution leads
+us to choose GHC as our server side compiler by default. We chose the Haste
+compiler to provide the client side code, mainly owing to our great familiarity
+with it and its handy ability to make use of vanilla Haskell packages from
+Hackage.
 

### 2014-05-07 paper.tex
Add bootstrapping info to first example.

@@
 
-\titlebanner{-- ICFP '14 draft submission --}        % These are ignored unless
+\titlebanner{-- Haskell Symposium '14 draft submission --}        % These are ignored unless
 \preprintfooter{-- -- -- -- draft -- -- -- --}   % 'preprint' option specified.
@@
 main = do
-  greetings <- export helloServer
+  greetings <- remote helloServer
 
@@
 
-The \lstinline!export! function takes an arbitrary function, provided that all
+The \lstinline!remote! function takes an arbitrary function, provided that all
 its arguments as well as its return value are serializable through the
 \lstinline!Serialize! type class, and produces a typed identifier which
-may be used to refer to the exported function. In this example, the type of
-\lstinline!greetings! is \lstinline!Export (String -> Server ())!,
+may be used to refer to the remote function. In this example, the type of
+\lstinline!greetings! is \lstinline!Remote (String -> Server ())!,
 indicating that the identifier refers
-to an exported function with a single \lstinline!String! argument and no return
-value. Exported functions all live in the \lstinline!Server! monad.
+to a remote function with a single \lstinline!String! argument and no return
+value. Remote functions all live in the \lstinline!Server! monad.
 This part of the program is executed on both the server and the client, albeit
@@
 
-After the \lstinline!export! call, we enter the domain of client-exclusive code
+After the \lstinline!remote! call, we enter the domain of client-exclusive code
 with the application of \lstinline!runClient!. This function executes
@@
 In order to make an RPC call using an identifier obtained from
-\lstinline!export!, we must supply it with an argument. This is done using the
+\lstinline!remote!, we must supply it with an argument. This is done using the
 \lstinline!<.>! operator. It might be interesting to note that its type,
 \linebreak
-\lstinline!Serialize a => Export (a -> b) -> a -> Export b!,
+\lstinline!Serialize a => Remote (a -> b) -> a -> Remote b!,
 is very similar to the type of the \lstinline!<*>! operator over applicative
 functors. This is not a coincidence; \lstinline!<.>! performs the same role
-for the \lstinline!Export! type as \lstinline!<*>! performs for applicative
+for the \lstinline!Remote! type as \lstinline!<*>! performs for applicative
 functors. The reason for using a separate operator for this instead of making
-\lstinline!Export! an instance of \lstinline!Applicative! is that since
-functions embedded in the \lstinline!Export! type exist only to be called over
+\lstinline!Remote! an instance of \lstinline!Applicative! is that since
+functions embedded in the \lstinline!Remote! type exist only to be called over
 a network, such functions must only be applied to arguments which can be
-serialized and sent over a network connection. When an \lstinline!Export!
+serialized and sent over a network connection. When a \lstinline!Remote!
 function is applied to an argument using \lstinline!<.>!, the argument is
-serialized and stored inside the resulting \lstinline!Export! object, awaiting
-dispatch. \lstinline!Export! values can thus be seen as explicit
+serialized and stored inside the resulting \lstinline!Remote! object, awaiting
+dispatch. \lstinline!Remote! computations can thus be seen as explicit
 representations of closures.
 
-After applying the value obtained from the user to the exported function,
+After applying the value obtained from the user to the remote function,
 we apply the \lstinline!onServer! function to the result, which dispatches
@@
 \label{lst:hello-server}
+\end{listingfloat}
+
+To run this example, an address and a port must be provided so that the client
+knows which server to contact. There are several ways of doing this:
+using the GHC plugin system, through Template Haskell or by slightly altering
+how program entry points are treated in a compiler or wrapper script, to name
+a few. A non-intrusive method when using the GHC/Haste compiler pair would
+be to pass add \lstinline!-main-is setup! to both compilers' command line
+and add the \lstinline!setup! function in figure \ref{lst:setup} to the source
+code. This will instruct the server binary to listen on the port 1111 when
+started, and the client to attempt contact with that port on the local machine.
+
+The exact mechanism chosen to provide the host and port are implementation
+specific, and will in the interest of clarity not be discussed further.
+
+\begin{listingfloat}
+\begin{code}
+setup :: IO ()
+setup =
+  runApp (mkConfig "ws://localhost:1111" 1111) main
+\end{code}
+\caption{Bootstrapping a Haste.App program}
+\label{lst:setup}
 \end{listingfloat}
@@
 \begin{code}
-fromRemote :: Remote a -> Server a
+use :: Remote a -> Server a
 \end{code}
@@
 
-  count <- export $ do
-    r <- fromRemote remoteref
+  count <- remote $ do
+    r <- use remoteref
     liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))
@@
 srvHello remoteRcpts = do
-  recipients <- fromRemote remoteRcpts
+  recipients <- use remoteRcpts
   sid <- getSessionID
@@
 srvSend remoteRcpts message = do
-    rcpts <- fromRemote remoteRcpts
+    rcpts <- use remoteRcpts
     liftIO $ do
@@
 srvAwait remoteRcpts = do
-  rcpts <- fromRemote remoteRcpts
+  rcpts <- use remoteRcpts
   sid <- getSessionID
@@
 to the client. In this tying, we see our main advantage over the JavaScript
-version in section \ref{sec:jsexample} in action: the \lstinline!export!
+version in section \ref{sec:jsexample} in action: the \lstinline!remote!
 function builds a strongly typed bridge between the client and the server,
@@
 
-  hello <- export $ srvHello recipients
-  awaitMsg <- export $ srvAwait recipients
-  sendMsg <- export $ srvSend recipients
+  hello <- remote $ srvHello recipients
+  awaitMsg <- remote $ srvAwait recipients
+  sendMsg <- remote $ srvSend recipients
 
@@
 Notice that the \lstinline!recipients! list is passed to our three server
-operations \emph{before} they are exported; since \lstinline!recipients! is
+operations \emph{before} they are imported; since \lstinline!recipients! is
 a mutable reference created on the server and inaccessible to client code,
@@
 proceeds to register itself with the server's recipient list using the
-\lstinline!hello! export.
+\lstinline!hello! remote computation.
 
@@
 runClient    :: Client () -> App Done
-liftServerIO :: IO a -> App (Remote a)
-export       :: Exportable a
-             => a -> App (Export a)
-
-onServer     :: Export (Server a) -> Client a
+liftServerIO :: IO a -> App (RemoteValue a)
+remote       :: Remotable a
+             => a -> App (Remote a)
+
+onServer     :: Remote (Server a) -> Client a
 (<.>)        :: Serialize a
-             => Export (a -> b) -> a -> Export b
-
-fromRemote   :: Remote a -> Server a
+             => Remote (a -> b) -> a -> Remote b
+
+use          :: RemoteValue a -> Server a
 getSessionID :: Server SessionID
@@
 server, as enforced by the type system, and are not observable on the client.} \\
-\lstinline!export! & \parbox[t]{5cm}{Make a server side function available to
+\lstinline!remote! & \parbox[t]{5cm}{Make a server side function available to
 be called remotely by the client.} \\
@@
 the client after it completes.} \\
-\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!export!ed function to
+\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!remote! function to
 a serializable argument.} \\
-\lstinline!fromRemote! & \parbox[t]{5cm}{Extract the contents of a
-\lstinline!Remote! value produced by \lstinline!liftServerIO! on the server.} \\
+\lstinline!use! & \parbox[t]{5cm}{Extract the contents of a
+\lstinline!RemoteValue! produced by \lstinline!liftServerIO! on the server.} \\
 \lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
@@
 Its \lstinline!CallID! state element contains the identifier to be given to the
-next exported function, and its other state element contains a mapping from
-identifiers to exported functions.
+next remote function, and its other state element contains a mapping from
+identifiers to remote functions.
 
@@
 \begin{code}
-data Remote a = Here a | There
-data Export a = Export CallID [JSON]
-
-liftServerIO :: IO a -> App (Remote a)
-export :: Exportable a => a -> App (Export a)
+data RemoteValue a = Here a | There
+data Remote a = Remote CallID [JSON]
+
+liftServerIO :: IO a -> App (RemoteValue a)
+remote :: Remotable a => a -> App (Remote a)
 \end{code}
@@
 Its argument IO computation is executed, and the result is wrapped in the
-\lstinline!Remote! constructor. The counterpart function, \lstinline!fromRemote!
+\lstinline!Remote! constructor. The counterpart function, \lstinline!use!
 simply unwraps the \lstinline!Here! constructor and returns the contained
@@
 \begin{code}
-fromRemote :: Remote a -> Server a
-fromRemote (Here x) = return x
-\end{code}
-
-\lstinline!export! is a more interesting function. Looking at its type, we see
-that it accepts any \lstinline!Exportable! value for its argument.
-\lstinline!Exportable! is defined as follows.
-
-\begin{code}
-class Exportable a where
-  mkExported :: a -> ([JSON] -> Server JSON)
-
-instance Serialize a => Exportable (Server a) where
-  mkExported m = fmap toJSON m
-
-instance (Serialize a, Exportable b) =>
-         Exportable (a -> b) where
-  mkExported f =
-    \(x:xs) -> mkExported (f $ fromJSON x) xs
+use :: Remote a -> Server a
+use (Here x) = return x
+\end{code}
+
+\lstinline!remote! is a more interesting function. Looking at its type, we see
+that it accepts any \lstinline!Remotable! value for its argument.
+\lstinline!Remotable! is defined as follows.
+
+\begin{code}
+class Remotable a where
+  mkRemote :: a -> ([JSON] -> Server JSON)
+
+instance Serialize a => Remotable (Server a) where
+  mkRemote m = fmap toJSON m
+
+instance (Serialize a, Remotable b) =>
+         Remotable (a -> b) where
+  mkRemote f =
+    \(x:xs) -> mkRemote (f $ fromJSON x) xs
 \end{code}
@@
 In essence, any function, over any number of arguments, which returns a
-serializable value in the \lstinline!Server! monad can be exported. The
-\lstinline!mkExported! function makes use of a well-known type class trick for
+serializable value in the \lstinline!Server! monad can be imported. The
+\lstinline!mkRemote! function makes use of a well-known type class trick for
 creating statically typed variadic functions, and works very much like the
@@
 
-The client implementation of \lstinline!export! is not very complex.
-
-\begin{code}
-export _ = App $ do
-  (next_id, exports) <- get
-  put (next_id+1, exports)
-  return (Export next_id [])
-\end{code}
-
-As the exported function lives on the server, the client only needs an
-identifier to be able to call on it. The exported function is thus ignored,
+The client implementation of \lstinline!remote! is not very complex.
+
+\begin{code}
+remote _ = App $ do
+  (next_id, remotes) <- get
+  put (next_id+1, remotes)
+  return (Remote next_id [])
+\end{code}
+
+As the remote function lives on the server, the client only needs an
+identifier to be able to call on it. The remote function is thus ignored,
 so that it can be optimized out of existence in the client version of the code.
@@
 \begin{code}
-export f = App $ do
-  (next_id, exports) <- get
-  put (next_id+1, (next_id, mkExported f) : exports)
-  return (Export next_id [])
-\end{code}
-
-The second argument of the \lstinline!Export! constructor is the list of
-arguments to be sent to the export, to be updated as the exported function is
+remote f = App $ do
+  (next_id, remotes) <- get
+  put (next_id+1, (next_id, mkRemote f) : remotes)
+  return (Remote next_id [])
+\end{code}
+
+The second argument of the \lstinline!Remote! constructor is the list of
+arguments to be sent to the function, to be updated as the remote function is
 applied using the \lstinline!<.>! operator.
@@
 (<.>) :: Serialize a
-      => Export (a -> b) -> a -> Export b
-(Export identifier args) <.> arg =
-  Export identifier (toJSON arg : args)
+      => Remote (a -> b) -> a -> Remote b
+(Remote identifier args) <.> arg =
+  Remote identifier (toJSON arg : args)
 \end{code}
@@
 onServer :: Serialize a
-         => Export (Server a) -> Client a
-onServer (Export identifier args) = do
+         => Remote (Server a) -> Client a
+onServer (Remote identifier args) = do
   nonce <- get
@@
 The \lstinline!ffi! function is typed using the same method as the
-\lstinline!mkExported! function described in section \ref{sec:impl}.
+\lstinline!mkRemote! function described in section \ref{sec:impl}.
 When applied to one or more arguments instantiating the \lstinline!Marshal!
@@
 
-Several other approaches to the seamless client-server interaction exist. In
-general, these attacks tend to be of the ``all or nothing'' variety,
+Several other approaches to seamless client-server interaction exist. In
+general, these proposed solutions tend to be of the ``all or nothing'' variety,
 introducing new languages or otherwise requiring custom full stack solutions.
@@
 
-\paragraph{Inspiration and alternatives to \lstinline!export!}
+\paragraph{Inspiration and alternatives to \lstinline!remote!}
 \label{sec:cloudhaskell}
@@
 
-In our solution, this representation is obtained through the use of the \lstinline!export! function, which when executed on the server pairs a function with a
+In our solution, this representation is obtained through the use of the \lstinline!remote! function, which when executed on the server pairs a function with a
 unique identifier and on the client, and when executed on the client returns
@@
 has the advantage of being simple to implement, one major drawback of this
-method is that all functions must be explicitly exported in the \lstinline!App!
+method is that all functions must be explicitly imported in the \lstinline!App!
 monad prior to being called over the network.
@@
 
-The astute reader may notice that this is exactly what the \lstinline!export!
-function does as well, the difference being that \lstinline!export! links the
+The astute reader may notice that this is exactly what the \lstinline!remote!
+function does as well, the difference being that \lstinline!remote! links the
 identifier to the value it represents on the type level, making it impossible

### 2014-05-07 paper.tex
Rephrase RemoteValue introduction.

@@
 \begin{code}
-liftServerIO :: IO a -> App (Remote a)
+liftServerIO :: IO a -> App (RemoteValue a)
 \end{code}
@@
 In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
-is wrapped in the abstract \lstinline!Remote! type. As the name implies,
+is wrapped in the abstract \lstinline!RemoteValue! type. As the name implies,
 values of this type live exclusively on the server. In order to inspect such a
@@
 \begin{code}
-use :: Remote a -> Server a
+use :: RemoteValue a -> Server a
 \end{code}
@@
 Therein lies the secret to restricting \lstinline!liftServerIO! to the
-server: the only function capable of turning a \lstinline!Remote! value into
-a useful one lives in the server monad. Any client side code is thus free to
-completely ignore executing computations lifted using \lstinline!liftServerIO!;
-since the result of a server lifted computation is never observable on the
-client, the client has no obligation to even produce such a value.
-Figure \ref{lst:good-state} shows how to make proper use of server side state.
+server: the only function capable of turning a \lstinline!RemoteValue! into
+one that can be inspected lives in the server monad. Any client side code is
+thus free to completely ignore executing computations lifted using
+\lstinline!liftServerIO!; since the result of a server lifted computation is
+never observable on the client, the client has no obligation to even produce
+such a value. Figure \ref{lst:good-state} shows how to make proper use of
+server side state.
 

### 2014-05-08 paper.tex
Restructure implementation stuff

@@
 be to pass add \lstinline!-main-is setup! to both compilers' command line
-and add the \lstinline!setup! function in figure \ref{lst:setup} to the source
-code. This will instruct the server binary to listen on the port 1111 when
+and add the \lstinline!setup! function to the source code.
+This will instruct the server binary to listen on the port 1111 when
 started, and the client to attempt contact with that port on the local machine.
 
-The exact mechanism chosen to provide the host and port are implementation
-specific, and will in the interest of clarity not be discussed further.
-
-\begin{listingfloat}
 \begin{code}
@@
 \end{code}
-\caption{Bootstrapping a Haste.App program}
-\label{lst:setup}
-\end{listingfloat}
+
+The exact mechanism chosen to provide the host and port are implementation
+specific, and will in the interest of clarity not be discussed further.
 
@@
 
-Aside from \lstinline!runClient!, the \lstinline!App! monad has two primary
-operations, both of which have slightly different implementations on the client
-and server respectively.
-
-\begin{code}
-data RemoteValue a = Here a | There
+The workings of the \lstinline!App! monad basically hinges on two abstract data
+types: \lstinline!RemoteValue! and \lstinline!Remote!. \lstinline!RemoteValue!
+is used to represent values which are available to the server but not to the
+client. \lstinline!Remote! denotes functions which live on the server but can
+be invoked remotely by the client. The implementation of these types and the
+functions that operate on them differ between the client and the server.
+
+\paragraph{Client side implementations}
+We begin by looking at the client side implementation.
+
+\begin{code}
+data RemoteValue a = There
 data Remote a = Remote CallID [JSON]
-
-liftServerIO :: IO a -> App (RemoteValue a)
+\end{code}
+
+As values of the \lstinline!RemoteValue! type are not accessible to the client,
+they are represented by a simple dummy value. The \lstinline!Remote! type is
+slightly more useful, containing the ID of a remote function and a list of the
+serialized arguments to be passed when invoking it. In essence, the
+\lstinline!Remote! type is an explicit representation of a remote closure.
+Such closures can be applied to values using the \lstinline!<.>! operator.
+
+\begin{code}
+(<.>) :: Serialize a
+      => Remote (a -> b) -> a -> Remote b
+(Remote identifier args) <.> arg =
+  Remote identifier (toJSON arg : args)
+\end{code}
+
+The \lstinline!remote! function is used to bring server side functions into
+scope on the client as \lstinline!Remote! functions. It is implemented using a
+simple counter which keeps track of how many functions have been imported so
+far and thus which ID to assign to the next remote function.
+
+\begin{code}
 remote :: Remotable a => a -> App (Remote a)
-\end{code}
-
-The client side definition of \lstinline!liftServerIO! is as simple as can be.
-
-\begin{code}
-liftServerIO _ = App $ return There
-\end{code}
-
-Its argument is simply ignored, and an opaque dummy value is returned; recall
-that server-lifted IO computations are not observable on the client. The
-server side implementation is only marginally more complex.
-
-\begin{code}
-liftServerIO m = App $ fmap Here m
-\end{code}
-
-Its argument IO computation is executed, and the result is wrapped in the
-\lstinline!Remote! constructor. The counterpart function, \lstinline!use!
-simply unwraps the \lstinline!Here! constructor and returns the contained
-value in the \lstinline!Server! monad.
-
-\begin{code}
-use :: Remote a -> Server a
-use (Here x) = return x
-\end{code}
-
-\lstinline!remote! is a more interesting function. Looking at its type, we see
-that it accepts any \lstinline!Remotable! value for its argument.
+remote _ = App $ do
+  (next_id, remotes) <- get
+  put (next_id+1, remotes)
+  return (Remote next_id [])
+\end{code}
+
+As the remote function lives on the server, the client only needs an
+identifier to be able to call on it. The remote function is thus ignored,
+so that it can be optimized out of existence in the client executable. Looking
+at its type, we can see that \lstinline!remote! accepts any argument
+instantiating the \lstinline!Remotable! class.
 \lstinline!Remotable! is defined as follows.
@@
 
-The client implementation of \lstinline!remote! is not very complex.
-
-\begin{code}
-remote _ = App $ do
+The final function operating on these types is \lstinline!liftServerIO!, used
+to initialize state holding elements and perform other setup functionality on
+the server.
+
+\begin{code}
+liftServerIO :: IO a -> App (RemoteValue a)
+liftServerIO _ = App $ return There
+\end{code}
+
+As we can see, the implementation is as simple as can be. Since
+\lstinline!RemoteValue!s are represented by a dummy value on the client, we
+just return said value.
+
+\paragraph{Server side implementations}
+The server side representation of the \lstinline!RemoteValue! and
+\lstinline!Remote! types are in a sense the opposites of their client side
+counterparts.
+
+\begin{code}
+data RemoteValue a = Here a
+data Remote a = Remote
+\end{code}
+
+Where the client is able to do something useful with the \lstinline!Remote!
+type but not with \lstinline!RemoteValue! the server has no way to inspect
+\lstinline!Remote! functions, and thus only has a no-op implementation of the
+\lstinline!<.>! operator. On the other hand, the server has full access to the
+contents of \lstinline!RemoteValue!s via the \lstinline!use! function.
+
+\begin{code}
+use :: RemoteValue a -> Server a
+use (Here x) = return x
+\end{code}
+
+Those values are produced by the \lstinline!remoteIO! function, which is quite
+simple. It merely executes its argument and wraps its argument in the
+\lstinline!RemoteValue! type.
+
+\begin{code}
+remoteIO :: IO a -> Server (RemoteValue a)
+remoteIO m = App $ fmap Here m
+\end{code}
+
+The server version of \lstinline!remote! is a little more complex than its
+client side counterpart. In addition to keeping track of the identifier of the
+next remote function, the server side \lstinline!remote! pairs up remote
+functions with these identifiers in an identifier-function mapping.
+
+\begin{code}
+remote f = App $ do
   (next_id, remotes) <- get
-  put (next_id+1, remotes)
-  return (Remote next_id [])
-\end{code}
-
-As the remote function lives on the server, the client only needs an
-identifier to be able to call on it. The remote function is thus ignored,
-so that it can be optimized out of existence in the client version of the code.
-The server version is only marginally more complex in that it updates the
-identifier-function mapping table as well.
+  put (next_id+1, (next_id, mkRemote f) : remotes)
+  return Remote
+\end{code}
 
@@
 
-\begin{code}
-remote f = App $ do
-  (next_id, remotes) <- get
-  put (next_id+1, (next_id, mkRemote f) : remotes)
-  return (Remote next_id [])
-\end{code}
-
-The second argument of the \lstinline!Remote! constructor is the list of
-arguments to be sent to the function, to be updated as the remote function is
-applied using the \lstinline!<.>! operator.
-
-\begin{code}
-(<.>) :: Serialize a
-      => Remote (a -> b) -> a -> Remote b
-(Remote identifier args) <.> arg =
-  Remote identifier (toJSON arg : args)
-\end{code}
-
+\paragraph{The server side dispatcher}
 After the \lstinline!App! computation finishes, the identifier-function mapping

### 2014-05-08 paper.tex
typo
Kill use/RemoteValue

@@
 
-While explaining of the properties of our solution is all well and good,
+While explaining the properties of our solution is all well and good,
 nothing compares to a good old Hello World example to convey the idea.
@@
 \begin{code}
-liftServerIO :: IO a -> App (RemoteValue a)
+liftServerIO :: IO a -> App (Server a)
 \end{code}
@@
 In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
-is wrapped in the abstract \lstinline!RemoteValue! type. As the name implies,
-values of this type live exclusively on the server. In order to inspect such a
-value is needs to be extracted somehow, something which is only possible using
-a single function:
-
-\begin{code}
-use :: RemoteValue a -> Server a
-\end{code}
-
-Therein lies the secret to restricting \lstinline!liftServerIO! to the
-server: the only function capable of turning a \lstinline!RemoteValue! into
-one that can be inspected lives in the server monad. Any client side code is
-thus free to completely ignore executing computations lifted using
-\lstinline!liftServerIO!; since the result of a server lifted computation is
-never observable on the client, the client has no obligation to even produce
-such a value. Figure \ref{lst:good-state} shows how to make proper use of
-server side state.
+is contained within the \lstinline!Server! monad, completely out of reach of the
+client. Any client side code is thus free to completely ignore executing
+computations lifted using \lstinline!liftServerIO!; since the result of a
+server lifted computation is never observable on the client, the client has no
+obligation to even produce such a value. Figure \ref{lst:good-state} shows how
+to make proper use of server side state.
 
@@
   count <- remote $ do
-    r <- use remoteref
+    r <- remoteref
     liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))
@@
 \begin{code}
-srvHello :: Remote RcptList -> Server ()
+srvHello :: Server RcptList -> Server ()
 srvHello remoteRcpts = do
-  recipients <- use remoteRcpts
+  recipients <- remoteRcpts
   sid <- getSessionID
@@
 \begin{code}
-srvSend :: Remote RcptList -> String -> Server ()
+srvSend :: Server RcptList -> String -> Server ()
 srvSend remoteRcpts message = do
-    rcpts <- use remoteRcpts
+    rcpts <- remoteRcpts
     liftIO $ do
@@
 \begin{code}
-srvAwait :: Remote RcptList -> Server String
+srvAwait :: Server RcptList -> Server String
 srvAwait remoteRcpts = do
-  rcpts <- use remoteRcpts
+  rcpts <- remoteRcpts
   sid <- getSessionID
@@
 
-use          :: RemoteValue a -> Server a
 getSessionID :: Server SessionID
@@
 \begin{code}
-data RemoteValue a = Here a
 data Remote a = Remote
@@
 Where the client is able to do something useful with the \lstinline!Remote!
-type but not with \lstinline!RemoteValue! the server has no way to inspect
-\lstinline!Remote! functions, and thus only has a no-op implementation of the
-\lstinline!<.>! operator. On the other hand, the server has full access to the
-contents of \lstinline!RemoteValue!s via the \lstinline!use! function.
-
-\begin{code}
-use :: RemoteValue a -> Server a
-use (Here x) = return x
-\end{code}
-
-Those values are produced by the \lstinline!remoteIO! function, which is quite
-simple. It merely executes its argument and wraps its argument in the
-\lstinline!RemoteValue! type.
-
-\begin{code}
-remoteIO :: IO a -> Server (RemoteValue a)
-remoteIO m = App $ fmap Here m
+type, the server has no way to inspect \lstinline!Remote! functions, and thus
+only has a no-op implementation of the \lstinline!<.>! operator. On the other
+hand, it does have full access to the values and side effects produced by
+the \lstinline!liftServerIO! function, which is quite simple. The side effect
+is performed immediately and its result saved for later, as state holding
+elements should only be initialized once.
+
+\begin{code}
+liftServerIO :: IO a -> App (Server a)
+liftServerIO m = App $ do
+  x <- liftIO m
+  return (return x)
 \end{code}

### 2014-05-08 paper.tex
Stuff.

@@
 viewpoint of the client that runs code on the server rather than the
-other way around, (3) functional and type-safe; and (4) portable;
+other way around, (3) functional and type-safe, and (4) portable;
 everything is implemented as a Haskell library that implicitly takes
@@
 technologies such as WebSockets\ \cite{websockets}, we may come up with
-something like the program in figure \ref{lst:javascript-client}.
+something like the program in figure \ref{lst:javascript-client} for our
+client program. In addition, a corresponding server program would need to be
+written to handle distribution of messages among clients. We will not give such
+an implementation here, as we do not believe it necessary to state the problem
+at hand.
 
@@
 and add the \lstinline!setup! function to the source code.
-This will instruct the server binary to listen on the port 1111 when
-started, and the client to attempt contact with that port on the local machine.
 
@@
 
+This will instruct the server binary to listen on the port 1111 when
+started, and the client to attempt contact with that port on the local machine.
 The exact mechanism chosen to provide the host and port are implementation
@@
     CC.modifyMVar recipients $ \cs ->
-      return ((sid, rcptMVar):cs ,())
+      return ((sid, rcptMVar):cs, ())
 \end{code}
@@
 runClient    :: Client () -> App Done
-liftServerIO :: IO a -> App (RemoteValue a)
+liftServerIO :: IO a -> App (Server a)
 remote       :: Remotable a
@@
 a serializable argument.} \\
-\lstinline!use! & \parbox[t]{5cm}{Extract the contents of a
-\lstinline!RemoteValue! produced by \lstinline!liftServerIO! on the server.} \\
 \lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
@@
 
-The workings of the \lstinline!App! monad basically hinges on two abstract data
-types: \lstinline!RemoteValue! and \lstinline!Remote!. \lstinline!RemoteValue!
-is used to represent values which are available to the server but not to the
-client. \lstinline!Remote! denotes functions which live on the server but can
+The workings of the \lstinline!App! monad basically hinges on the
+\lstinline!Server! and \lstinline!Remote! abstract data types.
+\lstinline!Server! is the monad wherein any server side code is contained, and
+\lstinline!Remote! denotes functions which live on the server but can
 be invoked remotely by the client. The implementation of these types and the
@@
 \paragraph{Client side implementations}
-We begin by looking at the client side implementation.
-
-\begin{code}
-data RemoteValue a = There
+We begin by looking at the client side implementation for those two types.
+
+\begin{code}
+data Server a = Server
 data Remote a = Remote CallID [JSON]
@@
 
-As values of the \lstinline!RemoteValue! type are not accessible to the client,
-they are represented by a simple dummy value. The \lstinline!Remote! type is
-slightly more useful, containing the ID of a remote function and a list of the
-serialized arguments to be passed when invoking it. In essence, the
-\lstinline!Remote! type is an explicit representation of a remote closure.
-Such closures can be applied to values using the \lstinline!<.>! operator.
+The \lstinline!Server! monad is quite uninteresting to the client; since
+operations performed within it can not be observed by the client in any way,
+such computations are simply represented by a dummy value.
+The \lstinline!Remote! type contains the identifier of a remote function and a
+list of the serialized arguments to be passed when invoking it. In essence,
+it is an explicit representation of a remote closure. Such closures can be
+applied to values using the \lstinline!<.>! operator.
 
@@
 simple counter which keeps track of how many functions have been imported so
-far and thus which ID to assign to the next remote function.
+far and thus which identifier to assign to the next remote function.
 
@@
 \begin{code}
-liftServerIO :: IO a -> App (RemoteValue a)
-liftServerIO _ = App $ return There
+liftServerIO :: IO a -> App (Server a)
+liftServerIO _ = App Server
 \end{code}
@@
 As we can see, the implementation is as simple as can be. Since
-\lstinline!RemoteValue!s are represented by a dummy value on the client, we
+\lstinline!Server! is represented by a dummy value on the client, we
 just return said value.
@@
 \paragraph{Server side implementations}
-The server side representation of the \lstinline!RemoteValue! and
+The server side representation of the \lstinline!Server! and
 \lstinline!Remote! types are in a sense the opposites of their client side
@@
 \begin{code}
+newtype Server a = Server (ReaderT SessionInfo IO a)
+  deriving (Functor, Applicative, Monad, MonadIO)
 data Remote a = Remote
@@
 Where the client is able to do something useful with the \lstinline!Remote!
-type, the server has no way to inspect \lstinline!Remote! functions, and thus
-only has a no-op implementation of the \lstinline!<.>! operator. On the other
-hand, it does have full access to the values and side effects produced by
-the \lstinline!liftServerIO! function, which is quite simple. The side effect
-is performed immediately and its result saved for later, as state holding
-elements should only be initialized once.
+type but can't touch \lstinline!Server! values, the server has no way to
+inspect \lstinline!Remote! functions, and thus only has a no-op implementation
+of the \lstinline!<.>! operator. On the other hand, it does have full access to
+the values and side effects of the \lstinline!Server! monad, which is an IO
+monad with some additional session data for the convenience of server
+side code.
+
+\lstinline!Server! values are produced by the \lstinline!liftServerIO! and
+\lstinline!remote! functions. \lstinline!liftServerIO! is quite simple:
+the function executes its argument immediately and the result is returned,
+tucked away within the \lstinline!Server! monad.
 

### 2014-05-08 paper.tex
Improve description of liftServerIO.
Even better liftServerIO!!11~1

@@
 
-In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
-is contained within the \lstinline!Server! monad, completely out of reach of the
-client. Any client side code is thus free to completely ignore executing
+\lstinline!liftServerIO! performs its argument computation once on the server,
+in the \lstinline!App! monad, and then returns the result of said computation
+inside the \lstinline!Server! monad so that it is only reachable by server side
+code.
+Any client side code is thus free to completely ignore executing
 computations lifted using \lstinline!liftServerIO!; since the result of a

### 2014-05-08 paper.tex
ServerDummy + RemoteDummy

@@
 \begin{code}
-data Server a = Server
+data Server a = ServerDummy
 data Remote a = Remote CallID [JSON]
@@
 liftServerIO :: IO a -> App (Server a)
-liftServerIO _ = App Server
+liftServerIO _ = App ServerDummy
 \end{code}
@@
   deriving (Functor, Applicative, Monad, MonadIO)
-data Remote a = Remote
+data Remote a = RemoteDummy
 \end{code}
@@
   put (next_id+1, (next_id, mkRemote f) : remotes)
-  return Remote
+  return RemoteDummy
 \end{code}

### 2014-05-12 paper.tex
ServerDummy -> return ServerDummy
pass add -main-is -> add -main-is

@@
 a few. A non-intrusive method when using the GHC/Haste compiler pair would
-be to pass add \lstinline!-main-is setup! to both compilers' command line
+be to add \lstinline!-main-is setup! to both compilers' command line
 and add the \lstinline!setup! function to the source code.
@@
 liftServerIO :: IO a -> App (Server a)
-liftServerIO _ = App ServerDummy
+liftServerIO _ = App $ return ServerDummy
 \end{code}

### 2014-05-12 paper.tex
clarity -> brevity wrt. why we don't talk more about launching Haste.App apps.
Remove runApp from Haste API.

@@
 The exact mechanism chosen to provide the host and port are implementation
-specific, and will in the interest of clarity not be discussed further.
+specific, and will in the interest of brevity not be discussed further.
 
@@
 \begin{code}
-runApp       :: AppCfg -> App Done -> IO ()
-
 runClient    :: Client () -> App Done
@@
 \hline
-\lstinline!runApp! & Begin execution of a Haste.App program. \\
 \lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!

### 2014-06-19 paper.tex
Fix Remotable implementation.
Fix JS syntax highlighting.

@@
 \begin{listingfloat}
-\begin{code}
+\begin{lstlisting}[language=JavaScript]
 function handshake(sock) {sock.send('helo');}
@@
 };
-\end{code}
+\end{lstlisting}
 \caption{JavaScript chatbox implementation}
@@
 instance Serialize a => Remotable (Server a) where
-  mkRemote m = fmap toJSON m
+  mkRemote m = \_ -> fmap toJSON m
 

### 2014-06-19 paper.tex
Fix omitted "top" on p2.

@@
 respectively. When a message arrives, it is prepended to the chat log, making
-the most recent message appear at the of the log window, and when the user hits
-the return key in the input text box the message contained therein is sent and
-the input text box is cleared.
+the most recent message appear at the top of the log window, and when the user
+hits the return key in the input text box the message contained therein is sent
+and the input text box is cleared.
 

### 2014-06-19 paper.tex
Fix garbled sentence on P10

@@
 In our solution, this representation is obtained through the use of the \lstinline!remote! function, which when executed on the server pairs a function with a
-unique identifier and on the client, and when executed on the client returns
+unique identifier, and when executed on the client returns
 said identifier so that the client may now refer to the function. While this

### 2014-06-19 paper.tex
Remove code listing of example program, unnecessary to repeat the code.

@@
 value. Remote functions all live in the \lstinline!Server! monad.
-This part of the program is executed on both the server and the client, albeit
-with slightly different side effects, as described in section \ref{sec:impl}.
+The part of the program contained within the \lstinline!App! monad is executed
+on both the server and the client, albeit with slightly different side effects,
+as described in section \ref{sec:impl}.
 
@@
 RPC call returns.
-
-Figure \ref{lst:hello-server} gives the code of this example in its entirety.
-
-\begin{listingfloat}
-\begin{code}
-import Haste.App
-
-helloServer :: String -> Server ()
-helloServer name =
-  liftIO $ putStrLn (name ++ " says hello!")
-
-main :: App Done
-main = do
-  greetings <- remote helloServer
-
-  runClient $ do
-    name <- prompt "Hi there, what is your name?"
-    onServer (greetings <.> name)
-\end{code}
-\caption{A seamless programming model: Hello Server}
-\label{lst:hello-server}
-\end{listingfloat}
 

### 2014-06-19 paper.tex
Remove "the handshake is a simple operation" from server example section.
Make it more explicit that Conc is an IO monad.

@@
 
-The handshake is a simple operation; an \lstinline!MVar! is associated with the
+An \lstinline!MVar! is associated with the
 connecting client's session identifier, and the pair is prepended to the
@@
 same semantics as their regular Haskell counterpart.\ \cite{ffi}
-This concurrency-enhanced monad is used as the basis of the \lstinline!Client!
-monad.
+This concurrency-enhanced IO monad is used as the basis of the
+\lstinline!Client! monad.
 

### 2014-06-19 paper.tex
Be clearer about why server-side state is wrapped in the Server monad.
Be more explicit about result MVars.

@@
 recipient list. Notice how the application's server state is passed in as the
-function's argument.
+function's argument, wrapped in the \lstinline!Server! monad in order to
+prevent client-side inspection.
 
@@
 type Nonce = Int
-type Client = StateT Nonce Conc
+type ClientState = (Nonce, Map Nonce (MVar JSON))
+type Client = StateT ClientState Conc
 \end{code}
@@
 \begin{code}
+newResult :: Client (Nonce, MVar JSON)
+newResult = do
+  (nonce, m) <- get
+  mv <- liftIO newEmptyMVar
+  put (nonce+1, insert nonce var m)
+  return (nonce, mv)
+
 onServer :: Serialize a
@@
 onServer (Remote identifier args) = do
-  nonce <- get
-  put (nonce + 1)
-  mv <- createResultMVar nonce
+  (nonce, mv) <- newResult
   webSocketSend $
     toJSON (nonce, identifier, reverse args)  
-  takeMVar mv
-\end{code}
+  fromJSON <$> takeMVar mv
+\end{code}
+
+The \lstinline!createResultMVar! function creates a new \lstinline!MVar!,
+paired with its corresponding nonce in the
 
@@
   let (nonce, result) = fromJSON response
-  mv <- findResultMVar nonce
-  putMVar mv result
+  (n, m) <- get
+  put (n, delete nonce m)
+  putMVar (m ! nonce) result
 \end{code}

### 2014-06-19 paper.tex
Use Chan rather than MVar for state example; be clearer about errors there.

@@
 
-type Recipient = (SessionID, CC.MVar String)
+type Recipient = (SessionID, CC.Chan String)
 type RcptList = CC.MVar [Recipient]
@@
       recipients <- CC.readMVar rcpts
-      mapM_ (CC.forkIO . deliver message) recipients
-  where
-    deliver msg (_,rcptMVar) = CC.putMVar rcptMVar msg
+      mapM_ (\rcpt -> writeChan rcpt message) recipients
 \end{code}
@@
 The send function is slightly more complex. The incoming message is written to
-the \lstinline!MVar! corresponding to each active session. Note the use of
-\lstinline!forkIO!. Some recipients may be slow to empty their \lstinline!MVar!
-to make room for a new message. In this case, we don't want to force all
-recipients to wait for the slowest one, so we spawn a new lightweight thread
-to handle delivery for each recipient.
+the \lstinline!Chan! corresponding to each active session.
 
@@
     case lookup sid recipients of
-      Just mv -> CC.takeMVar mv
+      Just mv -> CC.readChan mv
       _       -> fail "Unregistered session!"
@@
 The final server operation, notifying users of pending messages, finds the
-appropriate \lstinline!MVar! to wait on by searching the recipient list for the
+appropriate \lstinline!Chan! to wait on by searching the recipient list for the
 session identifier of the calling user, and then blocks until a message arrives
@@
 server. If this is the case, we simply drop the session by throwing an error;
-the session will be automatically cleaned up on the server side and an exception
-is thrown to the client.
+an exception will be thrown to the client. No server side state needs to be
+cleaned up as the very lack of such state was our reason for dropping the
+session.
 

### 2014-06-19 paper.tex
Fix Chan code.

@@
   liftIO $ do
-    rcptMVar <- CC.newEmptyMVar
+    rcptChan <- CC.newChan
     CC.modifyMVar recipients $ \cs ->
-      return ((sid, rcptMVar):cs, ())
+      return ((sid, rcptChan):cs, ())
 \end{code}
@@
       recipients <- CC.readMVar rcpts
-      mapM_ (\rcpt -> writeChan rcpt message) recipients
+      mapM_ (flip CC.writeChan message) recipients
 \end{code}

### 2014-06-25 haskell10f-ekblad.tex
Some formatting fixes + copyright info for publishing.
Add acknowledgement for RAWFP funding.
Fix bug in JS example.

--- 2014-06-19 paper.tex
@@
 
-\documentclass[preprint]{sigplanconf}
+\documentclass{sigplanconf}
 
@@
 
-\conferenceinfo{ICFP~'14}{September 1--3, 2014, Gothenburg, Sweden} 
+\exclusivelicense
+\conferenceinfo{ICFP~'14}{September 6, 2014, Gothenburg, Sweden}
 \copyrightyear{2014}
-\copyrightdata{978-1-nnnn-nnnn-n/14/09}
-\doi{nnnnnnn.nnnnnnn}
+\copyrightdata{978-1-4503-3041-1/14/09}
+\doi{2633357.2633367}
 
@@
 % traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
 
@@
 
-\title{A seamless, client-centric programming model for type safe web applications}
+\title{A Seamless, Client-Centric Programming Model for Type Safe Web Applications}
 
@@
         msgbox.value = '';
-        chat(msg);
+        chat(sock, msg);
       }
@@
 
+\acks
+This work has been partially funded by the Swedish Foundation for Strategic Research, under grant RAWFP.
+
 % We recommend abbrvnat bibliography style.

### 2014-06-25 haskell10f-ekblad.tex
Make code horribly ugly to fix overflow in PS.

@@
 function handshake(sock) {sock.send('helo');}
-function chat(sock, msg) {sock.send('text' + msg);}
+function chat(sock,msg) {sock.send('text'+msg);}
 
 window.onload = function() {
-  var logbox = document.getElementById('log');
-  var msgbox = document.getElementById('message');
+  var log = document.getElementById('log');
+  var msg = document.getElementById('msg');
   var sock = new WebSocket('ws://example.com');
@@
   sock.onmessage = function(e) {
-    logbox.value = e.data + NEWLINE + logbox.value;
+    log.value = e.data + NEWLINE + log.value;
   };
@@
     handshake(sock);
-    msgbox.addEventListener('keydown', function(e) {
+    msg.addEventListener('keydown', function(e) {
       if(e.keyCode == 13) {
-        var msg = msgbox.value;
-        msgbox.value = '';
+        var msg = msg.value;
+        msg.value = '';
         chat(sock, msg);
@@
 send and receive text messages in real time - we opt for a very simple
-design. Two UI elements, \lstinline!logbox! and \lstinline!msgbox!, represent
+design. Two UI elements, \lstinline!log! and \lstinline!msg!, represent
 the chat log and the text area where the user inputs their messages
@@
   runClient $ do
-    name <- prompt "Hi there, what is your name?"
+    name <- prompt "Hello, what is your name?"
     onServer (greetings <.> name)
@@
 setup :: IO ()
-setup =
-  runApp (mkConfig "ws://localhost:1111" 1111) main
+setup = runApp cfg main
+  where cfg = mkConfig "ws://localhost:1111" 1111
 \end{code}
@@
     r <- remoteref
-    liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))
+    liftIO $ atomicModifyIORef r (\v -> (v+1,v+1))
 
@@
   runClient $ do
-    withElems ["log","message"] $ \[log,msgbox] -> do
+    withElems ["log","msg"] $ \[log,msg] -> do
       onServer hello
@@
 In this case the variable \lstinline!log! will be bound to the node
-with the identifier ``log'', and \lstinline!msgbox! will be bound to the node
-identified by ``message''. These are the same DOM nodes that were referenced in
+with the identifier ``log'', and \lstinline!msg! will be bound to the node
+identified by ``msg''. These are the same DOM nodes that were referenced in
 our original example, and refer to the chat log window and the text input
@@
 \begin{code}
-      let recvLoop chatlines = do
-            setProp log "value" $ unlines chatlines
+      let recvLoop hist = do
+            setProp log "value" $ unlines hist
             message <- onServer awaitMsg
-            recvLoop (message : chatlines)
+            recvLoop (message : hist)
       fork $ recvLoop []
@@
 \begin{code}
-      msgbox `onEvent` OnKeyPress $ \13 -> do
-        msg <- getProp msgbox "value"
-        setProp msgbox "value" ""
-        onServer (sendMsg <.> msg)
+      msg `onEvent` OnKeyPress $ \13 -> do
+        msgtext <- getProp msg "value"
+        setProp msg "value" ""
+        onServer (sendMsg <.> msgtext)
 \end{code}
@@
 
-instance Serialize a => Remotable (Server a) where
+instance Serialize a =>
+         Remotable (Server a) where
   mkRemote m = \_ -> fmap toJSON m
@@
 \begin{code}
-newtype Server a = Server (ReaderT SessionInfo IO a)
-  deriving (Functor, Applicative, Monad, MonadIO)
+newtype Server a =
+  Server (ReaderT SessionInfo IO a)
+  deriving (Functor,Applicative,Monad,MonadIO)
 data Remote a = RemoteDummy
@@
   (next_id, remotes) <- get
-  put (next_id+1, (next_id, mkRemote f) : remotes)
+  put (next_id+1, (next_id,mkRemote f):remotes)
   return RemoteDummy
@@
 onEvent :: [(CallID, Method)] -> JSON -> IO ()
-onEvent mapping incoming = do
-  let (nonce, identifier, args) = fromJSON incoming
-      Just f = lookup identifier mapping
+onEvent mapping json = do
+  let (nonce, ident, args) = fromJSON json
+      Just f = lookup ident mapping
   result <- f args
@@
 type Nonce = Int
-type ClientState = (Nonce, Map Nonce (MVar JSON))
+type ClientState = (Nonce,Map Nonce (MVar JSON))
 type Client = StateT ClientState Conc
@@
 
--- A MutableVar is completely opaque to Haskell code
--- and is only ever manipulated in JavaScript. Thus,
--- we use the Unpacked type to represent it,
--- indicating a completely opaque value.
+-- A MutableVar is completely opaque to Haskell
+-- code and is only ever manipulated in
+-- JavaScript. Thus, we use the Unpacked type
+-- to represent it, indicating a completely
+-- opaque value.
 newtype MutableVar a = MV Unpacked
@@
 
-newMutable :: Marshal a => a -> IO (MutableVar a)
-newMutable = ffi "(function(x) {return {val: x};})"
-
-setMutable :: Marshal a => MutableVar a -> a -> IO ()
-setMutable = ffi "(function(m, x) {m.val = x;})"
-
-getMutable :: Marshal a => MutableVar a -> IO a
-getMutable = ffi "(function(m) {return m.val;})"
+newMutable :: Marshal a =>
+              a -> IO (MutableVar a)
+newMutable = ffi "(function(x){return {val:x};})"
+
+setMutable :: Marshal a =>
+              MutableVar a -> a -> IO ()
+setMutable = ffi "(function(m, x){m.val = x;})"
+
+getMutable :: Marshal a =>
+              MutableVar a -> IO a
+getMutable = ffi "(function(m){return m.val;})"
 \end{code}
